Vglut2-based glutamatergic signaling in central noradrenergic neurons is dispensable for normal breathing and chemosensory reflexes

Version 1

Abstract

Abstract

Central noradrenergic (NA) neurons are key constituents of the respiratory homeostatic network. NA dysfunction is implicated in several developmental respiratory disorders including Central Congenital Hyperventilation Syndrome (CCHS), Sudden Infant Death Syndrome (SIDS) and Rett Syndrome. The current unchallenged paradigm in the field, supported by multiple studies, is that glutamate co-transmission in subsets of central NA neurons plays a role in breathing control. If true, NA-glutamate co-transmission may also be mechanistically important in respiratory disorders. However, the requirement of NA derived glutamate in breathing has not been directly tested and the extent of glutamate co-transmission in the central NA system remains uncharacterized. Therefore, we fully characterized the cumulative fate maps and acute adult expression patterns of all three Vesicular Glutamate Transporters (Slc17a7 (Vglut1), Slc17a6 (Vglut2), and Slc17a8 (Vglut3)) in NA neurons, identifying a novel dynamic expression pattern for Vglut2 and an undescribed co-expression domain for Vglut3 in the NA system. Our functional studies showed that loss of Vglut2 throughout the NA system failed to alter breathing or metabolism under room air, hypercapnia, or hypoxia in unrestrained and conscious mice, which demonstrates that Vglut2-based glutamatergic signaling within the central NA system is not required for normal baseline breathing and hypercapnic, hypoxic chemosensory reflexes. These outcomes challenge the current understanding of central NA neurons in the control of breathing and suggests that glutamate may be not a critical target to understand NA neuron dysfunction in respiratory diseases.

Introduction

Breathing is a vital and life-sustaining function supporting homeostatic processes, most critically maintaining blood pH/CO2 and O2 levels within a narrow physiological range. Respiratory homeostasis is mediated by neuron-modulated lung ventilation adjustments in response to physiological deviations resulting in high pCO2 or low pO2 blood and tissue levels, known as the hypercapnic and hypoxic reflexes respectively (Del Negro et al., 2018; Dick et al., 2018). These reflexes are part of a complex brainstem neural network that integrates a multitude of information streams across the central and peripheral nervous systems to regulate respiratory output. Within this brainstem network, central noradrenergic (NA) neurons are known to be an important component that plays a variety of roles in modulating breathing. Furthermore, various perturbations across the central NA system have been implicated in several developmental disorders with respiratory and chemosensory features such as Central Congenital Hyperventilation Syndrome (CCHS), Sudden Infant Death Syndrome (SIDS) and Rett Syndrome (Beltrán-Castillo et al., 2017; Feldman et al., 2013; Gauda et al., 2007; Viemari, 2008). Thus, understanding how central NA neurons modulate respiratory chemoreflexes or chemosensory breathing is critically important for the development of new diagnostic and therapeutic interventions to address respiratory pathophysiology.

Central noradrenergic neurons are commonly thought to exert their effect on breathing through their primary neurotransmitter noradrenaline and adrenaline. However, in addition to noradrenaline, several studies suggest that glutamate is co-transmitted in subsets of central NA neurons. Vesicular glutamate transporter 2 (Vglut2), a gene marker of glutamatergic signaling, has been shown to be co-expressed in subsets of central NA neurons including, C1/A1, C2/A2, A5, and LC in adult rats and mice (DePuy et al., 2013; Souza et al., 2022a; Stornetta et al., 2002a, 2002b; Yang et al., 2021), though NA specific expression of related Vglut1 and Vglut3 transporters remains unknown. Additionally, it has been well documented that central NA neurons co-expressing Vglut2 innervate key respiratory centers, such as preBötzinger complex and parafacial region (pFRG), as well as many other autonomic brainstem and forebrain centers that could indirectly drive a change in breathing if perturbed, i.e., cardio-vascular function, the intermediolateral nucleus (IML)-sympathetic system and metabolism (Supplemental table 1). Thus, it has currently become a dominant paradigm in the field of respiratory physiology that NA-based glutamate release is a critical form of neurotransmission in the control of breathing. This paradigm is supported by multiple papers. Abbott et al. (2014) showed that Vglut2 is required for an increase in respiratory rate when anterior C1 neurons are unilaterally opto-genetically stimulated. Malheiros-Lima et al. (2020) showed that Vglut2-expressing C1 neurons project to the pFRG region, and hypoxic breathing was blunted after blockade of ionotropic glutamatergic receptors at the pFRG site in anesthetized rats, together supporting a role for anterior C1 neurons releasing glutamate at the pFRG site in turn to regulate breathing under hypoxia. Similarly, Malheiros-Lima et al. (2022) and Malheiros-Lima et al. (2018) showed that Vglut2-expressing C1 neurons project to the NA A5 region and the preBötzinger complex respectively, and again the blockade of ionotropic glutamatergic receptors at the A5 region or preBötzinger complex reduced the increase in phrenic nerve activities or in respiratory frequency elicited by optogenetic stimulation of C1 cells in an anesthetized preparation. In addition, Guyenet et al. (2013) and others speculate that the apparent lack of plasmalemmal monoamine transporter in C1 fibers indicates reduced or absent noradrenergic or adrenergic signaling due a lack of re-uptake and neurotransmitter pool depletion (Comer et al., 1998; Lorang et al., 1994). Cumulatively, the studies argue for glutamate as the predominant functional neurotransmitter for C1 NA neurons in the breathing neural network. To our knowledge, this dominant perspective has not been otherwise previously challenged. Although these studies are informative, the evidence supporting the role of NA-based Vglut2 signaling in respiratory control are either indirect and circumstantial or cannot be seen as physiological given experimental limitations, such as the focal nature of opto-genetic stimulation. Thus, it is not yet clear what the requirement is for NA-based glutamatergic signaling in homeostatic breathing in the awake and unrestrained animal and how that might inform upon disease.

To better understand the role of NA-based glutamatergic signaling in breathing, we sought to both fully characterize the molecular profiles of the central noradrenergic system with respect to glutamate co-expression and to determine the requirement of Vglut2-based glutamatergic release in respiratory control under physiological chemosensory challenges in awake and unrestrained mice. We first fully characterized the recombinase-based cumulative fate maps for Vglut1, Vglut2 and Vglut3 expression and compared those maps to their real time expression profiles in central NA neurons by RNA in situ hybridization in adult mice. We found a novel dynamic expression pattern for Vglut2 and an entirely undescribed co-expression domain for Vglut3 in the central NA system. Secondly, to determine if Vglut2-based glutamatergic signaling in NA neurons is required for respiratory homeostasis, we conditionally ablated Vglut2 in all NA neurons and tested respiratory, chemosensory, and metabolic function in unrestrained and conscious mice. Using the same genetic model in prior studies (Abbott et al., 2014), conditional deletion of Vglut2 in NA neurons did not significantly impact breathing under room air, hypercapnic, or hypoxic conditions. These results demonstrate, for the first time, that NA Vglut2-based glutamatergic signaling is dispensable for respiratory control, which challenges the prevalent perspective on the role of C1 and other Vglut2 expressing NA neurons in respiratory homeostasis and suggests that glutamate may not be a critical target to understand NA neuron dysfunction in respiratory diseases.

Results

Cumulative fate maps of central noradrenergic neurons co-expressing Vglut1, Vglut2 and Vglut3

To fully characterize the expression profiles of all three glutamate markers Vglut1, Vglut2 and Vglut3 in central noradrenergic neurons, we first used an intersectional genetic strategy (Figure 1A). We bred three Cre drivers Slc17a7Cre (Vglut1-Cre), Slc17a6Cre (Vglut2-Cre) and Slc17a8Cre (Vglut3-Cre) to Dbhp2a-Flpo (DBH-p2a-Flpo) (targeting noradrenergic neurons) mice respectively. The three compound lines of Vglut1-Cre/+; DBH-p2a-Flpo/+, Vglut2-Cre/+; DBH-p2a-Flpo/+, and Vglut3-Cre/+; DBH-p2a-Flpo/+ were then crossed with the Rosa26RC::FLTG (RC::FLTG) intersectional reporter mice which express tdTomato in cells expressing only flippase (Flpo) while expressing eGFP in cells co-expressing both Flpo and Cre recombinases. Thus, in each of the three intersectional reporter crosses [e.g. 1) Vglut1-Cre/+; DBH-p2a-Flpo/+; RC::FLTG/+, 2) Vglut2-Cre/+; DBH-p2a-Flpo/+; RC::FLTG/+ and 3) Vglut3-Cre/+; DBH-p2a-Flpo/+; RC::FLTG/+], NA neurons without any Vglut1/2/3 expression are labelled by red fluorescent protein (tdTomato) while NA neurons co-expressing either Vglut1, Vglut2, or Vglut3 are labeled by green fluorescent protein (eGFP). By using this method, we characterized and quantified the expression profiles of all three vesicular glutamate transporters Vglut1, Vglut2 and Vglut3 across every NA nucleus in adult mice including A7, Locus Coeruleus (LC), sub CD/CV, A5, and throughout the anterior - posterior dimension of C1/A1, C2/A2 (Figure 1B-D). Vglut1-Cre was not co-expressed in any central NA nuclei in adult mice. However, Vglut2-Cre and Vglut3-Cre both showed co-expression in central NA neurons. Vglut3-Cre-expressing NA neurons are restricted to posterior C2/A2 and posterior C1/A1, and the majority is in posterior C2/A2 where 26.9 ± 3.16% NA neurons are Vglut3-Cre positive. In posterior C1/A1, only 1.26 ± 0.559% NA neurons are Vglut3-Cre positive. Surprisingly, 84.6 ± 3.75% NA neurons in total have Vglut2-Cre co-expression and each NA nucleus was predominantly labelled by Vglut2-Cre expression. The percentage of Vglut2-Cre positive NA neurons in A7, LC, A5, sub CD/CV, anterior and posterior part of C1/A1, anterior and posterior part of C2/A2 is 69.8 ± 8.29%, 80.5 ± 3.78%, 84.8 ± 11.3%, 51.5 ± 2.61%, 99.0 ± 0.681%, 95.7 ± 2.40%, 96.1 ± 1.98%, 97.8 ± 0.771%, respectively (Mean ± SEM). The presence of Vglut2-Cre co-expression in anterior NA groups was unexpected as previous in situ data in adult rats found Vglut2 positive NA neurons only in the posterior C2/A2 and C1/A1 (DePuy et al., 2013; Stornetta et al., 2002a, 2002b). Notably, however, the Vglut2 co-expression in the LC region agrees with Yang et al. (2021) which also showed that a high percentage of LC NA neurons are Vglut2 positive by using another intersectional strategy in mice, ThFlpo; Slc17a6Cre; Rosa26Ai65 (TH-Flpo; Vglut2-Cre; Ai65). The differences in our and other fate maps compared to in situ hybridization may either reflect early gene expression that is down regulated in the adult or may reflect low levels of expression not detectable by in situ hybridization but that are nonetheless sufficient to effect recombination in the intersectional genetic strategy.

Cumulative fate maps of glutamate co-expressing NA neurons characterized by intersectional genetics.

(A) Breeding schematic. Vglut1-Cre/+, Vglut2-Cre/+ and Vglut3-Cre/+ were each bred with DBH-p2a-Flpo/+ mice respectively. These Vglut1, Vglut2, or Vglut3-Cre/+; DBH-p2a-Flpo/+ compound lines were then crossed with the RC::FLTG intersectional reporter mice. In each of the three intersectional reporter crosses, NA neurons co-expressing either Vglut1, Vglut2, or Vglut3 are labeled by green fluorescent protein (eGFP) while NA neurons without any Vglut1,2, or 3 expression are labelled by red fluorescent protein (tdTomato). Shown as an example is the Vglut2-Cre/+; DBH-p2a-Flpo/+; RC::FLTG/+ intersectional reporter line. (B) Quantification of the percentage of Vglut2-coexpressing NA neurons among NA neurons in each brainstem NA nuclei including A7, LC, A5, sub CD/CV, anterior C1/A1 and C2/A2, posterior C1/A1 and C2/A2. (C) Quantification of the percentage of Vglut3-coexpressing NA neurons among NA neurons in each NA nuclei. (D) Fluorescent expression of tdTomato (red) and eGFP (green) in coronal sections of Vglut1, Vglut2, or Vglut3 intersectional reporter lines in brainstem NA nuclei. Nucleus is labeled by DAPI (blue). Scale bar 50μm.

Real time mRNA expression patterns of Vglut1, Vglut2 and Vglut3 in central noradrenergic neurons in adult mice

DePuy et al. (2013), Stornetta et al. (2002a) and Stornetta et al. (2002b) showed that Vglut2 co-expression is only located in the C2/A2 and C1/A1 region in the adult rat but not in anterior NA nuclei in the brainstem by in situ hybridization. To verify if the real time NA-based Vglut2 expression pattern in adult mice is the same as in adult rats and to further characterize the expression patterns of the other two vesicular glutamate transporters (Vglut1 and Vglut3) in adult mice, we performed fluorescent RNA in situ hybridization experiments. We co-stained Vglut1, Vglut2, Vglut3 with DBH respectively in brain tissue of adult mice and characterized the colocalization of Vglut1/2/3 with DBH in brainstem NA nuclei A7, LC, A5, sub CD/CV, anterior C1/A1 and C2/A2, and posterior C1/A1 and C2/A2 (Figure 2A-C). No Vglut1/DBH double positive neurons were detected in any part of the central NA system, consistent with our fate map. Vglut3 mRNA colocalization with DBH is only found in the posterior part of C2/A2 where 27.1 ± 1.86% DBH positive neurons have Vglut3 co-localization, again consistent with our fate map. For Vglut2, we only found detectable levels of Vglut2 mRNA in C1/A1 and C2/A2 NA region, but not in A7, LC, A5, sub CD/CV. Quantitatively, the percentage of Vglut2/DBH double positive neurons in anterior C1/A1, anterior C2/A2, posterior C1/A1 and posterior C2/A2 is 84.7 ± 5.82%, 66.3 ± 0.335%, 35.7 ± 3.01% and 90.1 ± 2.45%, respectively. This result is consistent with the previous in situ data in adult rats (DePuy et al., 2013; Stornetta et al., 2002a, 2002b) suggesting that there is no obvious expression difference of Vglut2 in the central NA system between mouse and rat. However, the Vglut2 in situ data did not show expression in anterior NA populations. This difference between our in situ data and fate map data supports our previous hypothesis that many NA neurons expressed Vglut2 at some point from early development toward adulthood but the expression of Vglut2 is turned back down during adulthood. Interestingly, Yang et al. (2021) did show adult Vglut2 co-expression in LC by a viral injection of both Cre and Flpo-dependent eYFP into the LC of a bi-transgenic mouse with both TH-Flpo and Vglut2-Cre. To verify this result, we injected an AAV virus containing Cre-dependent tdTomato (pAAV-EF1a-DIO-tdTomato-WPRE) into the LC region of adult Vglut2-Cre mice. Consistent with Yang et al. (2021), we observed sparse Vglut2-Cre and TH (immunopositive) double positive neurons in the LC (Figure 2 supplement 1). Additionally, Souza et al. (2022a) showed about 28.5% A5 neurons are Vglut2 mRNA positive in adult rats by RNA scope, a method which is more sensitive than the traditional in situ hybridization we used here. These data suggest that some adult NA neurons in LC and A5 have Vglut2 co-expression, but the Vglut2 mRNA level cannot be detected by traditional in situ hybridization.

Characterization of real time co-expression of Vglut1, Vglut2 and Vglut3 in brainstem NA neurons in adult mice by fluorescent in situ hybridization.

(A) Representative images of Vglut1, Vglut2, or Vglut3 and DBH double ISH in coronal sections of WT mice in all of brainstem NA nuclei in adult mice. DBH (green), Vglut1/2/3 (red), DAPI (cyan). Scale bar 50μm. (B) Quantification of the percentage of Vglut2 co-expression in NA neurons in each NA nuclei. (C) Quantification of the percentage of Vglut3 co-expression in NA neurons in each NA nuclei.

Vglut2 expression is effectively knocked down in the whole central noradrenergic system in DBH-Cre; Vglut2 cKO mice

To investigate the requirement of Vglut2-based glutamatergic signaling in the central NA system in breathing under physiological challenges, we used Dbh-Cre; Slc17a6flox/flox (DBH-Cre; Vglut2 cKO) as the mouse model in which Vglut2 is inactivated in all noradrenergic neurons. This is the same model used by Abbott et al. (2014) to show that Vglut2 expression in the NA system was required for a modest increase in respiratory frequency following unilateral optogenetic stimulation of anterior C1. To verify if Vglut2 expression is effectively removed from central NA neurons in DBH-Cre; Vglut2 cKO mice, we carried out fluorescent RNA in situ hybridization for Vglut2 and DBH in the mouse brainstems and compared the Vglut2 signal intensity in DBH positive neurons in DBH-Cre; Vglut2 cKO and its littermate controls. We found that Vglut2 mRNA expression is 96.4 ± 0.226% decreased compared to controls, indicating that the Vglut2 is effectively recombined to abrogate expression in NA neurons (Figure 3A-B).

DBH with Vglut2 co-staining by fluorescent in situ hybridization confirms that Vglut2 expression is disrupted in the whole central NA system in DBH-Cre; Vglut2 cKO mice.

(A) Representative images of Vglut2 and DBH double ISH in anterior C1/A1, posterior C1/A1, anterior C2/A2 and posterior C2/A2 in control and DBH-Cre; Vglut2 cKO mice. DBH (green), Vglut2 (red), DAPI (cyan). Scale bar 20μm.(B) Quantification of intensity of Vglut2 signal in control and DBH-Cre; Vglut2 cKO mice.

Vglut2-based glutamatergic signaling in central NA neurons is not required for baseline breathing nor for the hypercapnic ventilatory reflex under distinct CO2 challenges

Multiple prior studies provide indirect evidence that NA glutamate transmission may play roles in respiratory function, particularly that anterior C1 neurons mediate the hypoxic ventilatory response through the pFRG/RTN and/or the NA A5 group (Malheiros-Lima et al., 2022, 2020). Additionally, the A5 group has been implicated in the hypercapnic reflex (Haxhiu et al., 1996). Thus, we sought to determine if Vglut2-based glutamatergic signaling in central NA neurons is necessary to regulate baseline breathing and the hypercapnic chemoreflex in awake and unrestrained animals. We used whole-body barometric plethysmography to measure the breathing and metabolism of conscious and free-moving mice. After a five-day habituation protocol, the ventilation response of DBH-Cre; Vglut2 cKO mice and their littermate controls was measured under room air and hypercapnia (5% CO2) (Figure 4A). Surprisingly, DBH-Cre; Vglut2 cKO mice didn’t show significant differences in overall respiratory output (VE/VO2), respiratory rate (Vf), tidal volume (VT), minute ventilation (VE) and metabolism demand (VO2) compared to their littermate controls under both room air and 5% CO2 challenge (Figure 4B and Table 1-2). These results suggested that Vglut2-based glutamatergic signaling is not required for regulating baseline breathing and hypercapnic ventilatory reflex under 5% CO2.

Vglut2 conditional knockout in central noradrenergic neurons fails to alter baseline breathing and the hypercapnic ventilatory reflex (5% CO2).

(A) Mouse model schematic and experimental protocol including habituation and hypercapnia protocol. (B) Under both room air and hypercapnia (5% CO2), DBH-Cre; Vglut2 cKO mice didn’t show significant changes in respiratory output (VE/VO2), minute ventilation (VE), respiratory rate (Vf), tidal volume (VT), and metabolism demand (VO2). Linear mixed-effects regression model, *p<0.05.

To further investigate if Vglut2-based glutamatergic signaling in central NA neurons is required for ventilation response of higher CO2 challenges, we also measured the breathing of DBH-Cre; Vglut2 cKO mice under 7% CO2 and 10% CO2. We found that under the 7% CO2 condition, DBH-Cre; Vglut2 cKO mice showed only a significantly decreased tidal volume (VT), but the overall respiratory output (VE/VO2), respiratory rate (Vf), minute ventilation (VE) and metabolism demand (VO2) didn’t show a significant difference between the mutant mice and their sibling controls (Figure 5A-B and Table 3). This result suggests that Vglut2-based glutamatergic signaling in central NA neurons is not required or at least is not as important as previously expected for regulating the hypercapnic ventilatory reflex under 7% CO2.

Vglut2 conditional knockout in central noradrenergic neurons fails to alter the majority of hypercapnic ventilation parameters and metabolism under 7% CO2.

(A) Mouse model schematic and experimental protocol including habituation and hypercapnia protocol. (B) Knocking out Vglut2 in the whole NA system failed to alter ventilation (VE/VO2, Vf, VE) and metabolism (VO2) but only showed a significantly reduce in VT under 7% CO2 condition. Linear mixed-effects regression model, *p<0.05.

Under 10% CO2 challenge, none of the overall respiratory outputs (VE/VO2), respiratory rate (Vf), tidal volume (VT), minute ventilation (VE) and metabolism demand (VO2) were significantly different between DBH-Cre; Vglut2 cKO mice and their littermate controls, suggesting that Vglut2-based glutamatergic signaling in central NA neurons is not required for the hypercapnic ventilatory reflex under 10% CO2 (Figure 6A-B and Table 4).

Vglut2 conditional knockout in central noradrenergic neurons fails to alter the hypercapnic ventilatory reflex (10% CO2).

(A) Mouse model schematic and experimental protocol including habituation and hypercapnia protocol. (B) Knocking out Vglut2 in the whole NA system didn’t significantly alter ventilation (VE/VO2, Vf, VT, VE) and metabolism (VO2) under 10% CO2 condition. Linear mixed-effects regression model, *p<0.05.

Vglut2-based glutamatergic signaling in central NA neurons is not required for the hypoxic ventilatory reflex under 10% O2

To determine if Vglut2-based glutamatergic signaling in central NA neurons is necessary to regulate the hypoxic chemoreflex, we measured the ventilation response of DBH-Cre; Vglut2 cKO mice under 10% O2. We analyzed the five respiratory parameters in three 5-min time periods under 10% O2 challenge. No significant difference was found in the overall respiratory output (VE/VO2), respiratory rate (Vf), tidal volume (VT), minute ventilation (VE) and metabolism demand (VO2) between DBH-Cre; Vglut2 cKO mice and controls in either of the three 5-min 10% O2 periods (Figure 7A-B and Table 5). These data suggest that Vglut2-based glutamatergic signaling in central NA neurons is not required for the hypoxic ventilatory reflex under 10% O2.

Vglut2 conditional knockout in central noradrenergic neurons fails to alter the hypoxic ventilatory reflex (10% O2).

(A) Mouse model schematic and experimental protocol including habituation and hypoxia protocol. (B) Knocking out Vglut2 in the whole NA system didn’t show significant breathing changes including VE/VO2, VE, Vf, VT, and VO2 in either of the three five-minute time periods under hypoxia (10% O2). Linear mixed-effects regression model, *p<0.05.

Discussion

Neuron co-transmission of two or more neurotransmitters across excitatory, inhibitory, and neuro-modulatory facets is becoming an increasingly appreciated phenomenon in the central nervous system. Functional interrogation of co-transmission is essential not only to understand the distinct and cooperative roles that multiple signaling molecules may play at the synapse but also to determine the most critical targets for potential therapeutics. To our knowledge, the unopposed paradigm in the field until now has been that vesicular glutamate transporter 2 (Vglut2)-based glutamate transmission from central noradrenergic neurons is important in respiratory homeostasis. Anatomically, it has long been appreciated that central NA neurons co-express Vglut2, the major glutamate marker among three glutamate transporters (DePuy et al., 2013; Souza et al., 2022a; Stornetta et al., 2002a, 2002b; Yang et al., 2021). Also, it has been well documented that NA Vglut2 positive fibers innervate known central respiratory centers or other autonomic centers, that, when perturbed, could disrupt respiratory homeostasis (Supplemental Table 1). Functionally, multiple reports have argued that Vglut2-based glutamate transmission plays a role in respiratory homeostasis. It has also been suggested that, at least for C1 neurons, the apparent lack of a plasmalemmal monoamine transporter may attenuate or eliminate noradrenergic or adrenergic release from these fibers. However as discussed below, much of the functional evidence supporting a role for Vglut2-based glutamate transmission from central noradrenergic neurons is circumstantial or of a non-physiological nature, and absent or attenuated release of adrenaline and noradrenaline has not been demonstrated. Nonetheless, the dominant perspective that NA-Vglut2 glutamate transmission plays a role in breathing has remained unchallenged. Our studies here uncover a novel dynamic expression pattern for Vglut2 and an entirely undescribed co-expression domain for Vglut3 in the central NA neurons, and in contrast to prior work, our results show that loss of Vglut2 in NA neurons does not appreciably change baseline or chemosensory breathing in conscious and free-moving mice.

Vglut2 has been shown to be co-expressed in subsets of central NA neurons, however, to our knowledge, the potential for the central NA neurons to express the other glutamate transporters Vglut1 or Vglut3 has not been reported at the anatomical level. To determine if central NA neurons express either Vglut1 or Vglut3 and to confirm the Vglut2 expression, we carried out cumulative intersectional fate mapping using Vglut1-Cre/+; DBH-p2a-Flpo/+; RC::FLTG/+, Vglut2-Cre/+; DBH-p2a-Flpo/+; RC::FLTG/+ and Vglut3-Cre/+; DBH-p2a-Flpo/+; RC::FLTG/+ compound mouse lines. As this approach maps the cumulative history of expression, we also examined acute adult expression of Vglut1, Vglut2 and Vglut3 using fluorescent RNA in situ hybridization. Our in situ results for Vglut2 agreed with earlier published outcomes (DePuy et al., 2013; Stornetta et al., 2002a, 2002b) showing as much as 80% expression in posterior NA neurons. However, there was no appreciable signal in anterior groups (A7, LC, A5, sub CD/CV). This lack of expression was in stark contrast to the cumulative fate map for Vglut2, showing as much as 75% of the anterior central NA neurons were recombined by the Vglut2-Cre over the lifetime of the animal. To further interrogate the acute activity of the Vglut2-Cre in the Locus Coeruleus in the adult mouse, we injected a Cre responsive virus in LC of Vglut2-Cre mice. The outcomes from the cumulative fate map and acute injections are in agreement with published results from Yang et al. (2021). Notably our use of the DBH-p2a-Flpo avoided expression in the PBN that was seen in the TH-Flpo mouse. In addition, 28.5% A5 neurons were shown to be Vglut2 positive in adult rats by RNA scope (Souza et al., 2022b). It is notable that both our lab and work published by (DePuy et al., 2013; Stornetta et al., 2002a, 2002b) failed to detect Vglut2 expression in anterior groups, but that apparently very low levels can be detected by more sensitive recombinase and RNA scope methods. Additional electrophysiological evidence suggests that this expression is functional (Yang et al., 2021). Our data suggests that Vglut2 co-expression in anterior NA groups is time dependent, though the underlying temporal dynamics of Vglut2 remain unknown. It is notable that we see nearly 80% of the LC and 85% of A5 recombined in the intersectional strategy representing lifetime expression, but our and other studies find only 20% of the LC and 28.5% of A5 may be expressing Vglut2 at any given moment in adults. This suggests either a temporal restriction in expression as the animal matures or may reflect dynamic change driven by behavioral or physiological experiences.

For Vglut1, neither fate map nor in situ hybridization revealed any co-expression within the central NA system. However, Vglut3 expression was found in the posterior part of NA neurons including C2/A2 and C1/A1, which is first characterized by this study. The majority of this Vglut3 co-expression is restricted to the posterior part of the C2/A2 NA population. This NA region has not been heavily implicated in respiratory control nonetheless it readily lends itself to interrogation by intersectional genetic methods and in situ hybridization. Given its distal expression from anterior C1, it is unlikely that Vglut3 would compensate for loss of Vglut2 in anterior C1, though this has not been ruled out. Further, while Vglut3 is a marker of glutamatergic neurons, its role in neurotransmission is not wholly defined as Vglut3 is typically found in soma and dendrites (Fremeau et al., 2004).

Based on the Vglut2 co-expression in NA neurons and their projections to multiple brain regions important in respiration, several studies suggest that glutamate in NA populations is important in breathing control. Those studies provide a mix of circumstantial, indirect, or limited evidence for the potential role of Vglut2-based glutamatergic signaling in central NA neurons in breathing regulation. However, it has not yet been shown whether or not NA-based Vglut2 signaling is required for respiratory control due to experimental caveats or limitations.

Abbott et al. (2014) showed that conditionally knocking out Vglut2 in the whole central NA system attenuated a modest increase in respiratory rate resulting from unilateral C1 optogenetic stimulation. Absolute values and other respiratory parameters, including tidal volume (VT) and minute ventilation (VE) were not reported, making direct comparison difficult. While the optogenetic stimulation did drive a Vglut2 dependent increase in respiratory rate, it is not clear if the resulting change in breathing reflects a native feature of the network that might be engaged in chemosensory responses, exercise, or volitional control. Thus, it remains possible that strong optogenetic stimulation of a small, isolated population drives a Vglut2 dependent but ectopic function that is not typically engaged in the normal operation of the breathing network. Such ectopic outcomes could come from several potential, non-exclusive mechanisms. First, unilateral stimulation could drive a network hysteresis or dysregulation akin to a focal injury that is not a typical biological feature (Ducros et al., 2003). This point is well made in Abbott et al. (2013) in their discussion of C1 optogenetic stimulation where they state, “resulting cardiorespiratory response pattern should not be considered strictly “physiological,” because a physiological response is never initiated by selective activation of a single cluster of CNS neurons and the various subsets of C1 neurons are presumably never recruited en bloc under any physiological condition”. Second, strong stimulation could result in overwhelming synaptic mechanisms (i.e., glial uptake) resulting in glutamate spillover and unintended cross talk to affect extra-synaptic neurons proximal to targeted NA fibers. Activity dependent spillover transmission has been documented elsewhere in the nervous system (Henneberger et al., 2020; Hülsmann et al., 2000). Third, the outcomes could be secondary to a different autonomic or behavioral function requiring NA-derived glutamatergic signaling involving metabolism or circulation (DePuy et al., 2013; Yang et al., 2021).

More circumstantially, Malheiros-Lima et al. (2020) suggested C1 neurons potentially release glutamate at the pFRG site to regulate breathing under hypoxia by providing three indirect lines of evidence: 1) Vglut2-expressing C1 neurons project to the pFRG region; 2) Increased breathing was blunted after blockade of ionotropic glutamatergic receptors at the pFRG site under anesthesia using chemical hypoxia; 3) Depletion of C1 neurons eliminated the increased breathing elicited by hypoxia. However, 1) the glutamatergic signaling targeting the pFRG region is not necessarily from C1 NA neurons since glutamatergic neurons from other regions also project to the pFRG site (Yang and Feldman, 2018).Thus, it is possible that the glutamatergic signaling that is required for hypoxic response is derived from other glutamatergic neurons such as the preBötzinger complex and the RTN itself; 2) C1 neuron depletion knocks out all the signaling modalities in the C1 population including both noradrenaline and glutamate. It is not clear which signaling drives the hypoxic response.

Similarly, Malheiros-Lima et al. (2022) and Malheiros-Lima et al. (2018) showed that Vglut2-expressing C1 neurons project to the NA A5 region and the preBötzinger complex respectively. The blockade of ionotropic glutamatergic receptors at the A5 region or preBötzinger complex reduced the increase in phrenic nerve activities or the breathing frequency caused by optogenetic stimulation of C1 cells in anesthetized preparation. Together these results suggest that Vglut2-expressing C1 neurons communicate with A5 or preBötzinger complex neurons by releasing glutamate in turn to regulate phrenic nerve activities or breathing frequency. Again, these two studies showed that C1 neurons are important to regulate phrenic nerve activities or breathing frequency, however, it is not necessarily through a direct C1-A5 or C1-preBötzinger complex glutamatergic pathway, whereas an indirect multi-synaptic pathway originating with C1 noradrenaline and ending with glutamate transmission from an intermediate to A5 or preBötzinger complex cannot be ruled out. Importantly, in all four studies noted here, respiratory measurements were limited or proxies of breathing under anesthesia were used. The role of NA-Vglut2 signaling was not reported in the context of an awake, unrestrained animal under normoxic, hypoxic, or hypercapnic breathing.

Thus, the limitations and indirect lines of the evidence in these studies raised the question as to whether or not glutamatergic signaling in central NA neurons is required or necessary to modulate respiratory homeostasis under more physiologically relevant circumstances and if NA-derived glutamate is an important mechanism to understand the pathophysiology of NA neurons in respiratory diseases. The goal of our study was to test the requirement of NA-based glutamatergic signaling in breathing homeostasis in the awake and unrestrained animal under conditions that would engage the whole respiratory control network rather than a singular component that may drive hysteresis or artifactual outcomes. Our in vivo breathing data under room air, hypoxic, and multiple hypercapnic conditions failed to show the requirement of NA-derived Vglut2 in normal breathing and respiratory chemoreflexes directly. To ensure a rigorous and robust conclusion, first we use exactly the same mouse model, DBH-Cre; Vglut2 cKO, as Abbott et al. (2014) used, and confirmed, by a second method (fluorescent RNA in situ hybridization) that Vglut2 expression was indeed abrogated in the whole central NA system. Our results are in strong agreement with DePuy et al. (2013) that recombination is efficient. Notably, we assayed for expression at the cell body, rather than the fibers and see no singular cell body that is scored positive. Second, we measured the breathing of conscious, unrestrained, and habituated mice under physiological challenges including normoxia, hypercapnia and hypoxia. Third, multiple respiratory and metabolic parameters were measured and reported in absolute values including respiratory rate, tidal volume, minute ventilation, oxygen consumption, overall respiratory output (minute ventilation normalized to oxygen consumption) in order to visualize breathing function comprehensively. Fourth, our experimental design was overpowered (i.e., n=16-21 vs 5-13 needed for power based on power analysis and n=7-8 reported in other manuscripts). Each respiratory/metabolic parameter for each physiological condition including room air, hypercapnia, and hypoxia was derived by quantifying the entire respiratory trace and analyzed between mutant and sibling controls by using a powerful linear mixed-effects regression model with animal type (experimental vs. control) as fixed effects, animal ID as a random effect and including sex as a variable (there was no sex specific effect) (Lusk et al., 2022).

Despite our experimental design and the extensive nature of our measurements, there remain several considerations in the interpretation of our data. First, we uncovered a novel NA co-expressed glutamate transporter, Vglut3. It remains a formal possibility that Vglut3 somehow compensates for the loss of Vglut2, however we argue that this is unlikely for the following reasons. First, our model is the same used in the other study where an effect was seen with optogenetic stimulation. Second, focus has been on anterior C1 whereas Vglut3 is expressed in the posterior CA domains. Another consideration is that pre- or post-natal developmental compensation corrects for the loss of NA derived glutamate. Again, we argue that is not likely, as we are using the same adult model that provided the most direct evidence and first functionally demonstrated a potential role for C1-derived glutamate in breathing. Second, Abbott et al. (2014) highlight that Vglut2 negative NA neurons showed no obvious abnormalities in number, morphology, and projection patterns. Third, we have examined the requirement of NA-expressed Vglut2 in P7 neonate mice in the auto-resuscitation reflex and saw no differences (data not shown). However, this could be better tested by using the Dbh-CreERT2 (DBH-CreERT2) (Stubbusch et al., 2011) to more acutely remove Vglut2 expression (though this leaves a window of 2-3 weeks for compensation to occur). Yet another consideration in interpreting our results is that we only address the loss of NA-Vglut2 based signaling across the entire NA system. It remains a formal possibility that we have removed Vglut2 dependent signaling from two counter balancing regions of the NA system and are therefore left with null results whereas removal from anterior C1 alone would show a requirement. To test this possibility, it would require an intersection conditional approach, something that has not yet been shown to be efficient and effective and would be beyond the scope of these studies. Additionally, in our studies, we account for potential metabolic changes that may underlie or be coordinate with changes in breathing. However, in this study we did not test for potential changes in blood pressure or other autonomic functions as well as affective state that may, in some obscure way, compensate for changes in breathing. Lastly, our results are loss of function in nature. We do not test sufficiency or gain of function and cannot fully rule out a role for NA-Vglut2 based glutamatergic signaling in the control of breathing.

From a translational perspective, our data questions whether or not glutamatergic signaling in NA neurons is likely to be a key mechanism and therefore a therapeutic target for breathing disorders, such as Rett syndrome and SIDS. Mecp2-deficient mice (a mouse model of Rett Syndrome that phenocopies many human symptoms) showed both a deficiency of NA populations (reduced number of NA neurons in C2/A2 and C1/A1 group) and highly variable respiratory rhythm at around 4-5 weeks of age (Roux et al., 2007; Viemari et al., 2005). 80% Rett Syndrome patients experience breathing issues, such as unstable breathing, episodes of hyperventilation and breathing holds, during the whole lifespan including both childhood and adulthood (Ramirez et al., 2020). SIDS decedents show NA abnormalities and are hypothesized to ultimately succumb from a failure in cardiorespiratory autoresuscitation (Chigr et al., 1989; Garcia et al., 2013; Kopp et al., 1993; Mansouri et al., 2001; Ozawa et al., 2003; Takashima and Becker, 1991). However, DBH-Cre; Vglut2 cKO mice, have normal baseline and chemosensory respiratory parameters as adults and neonate mice show normal autoresuscitation indicating that perturbed NA based glutamatergic signaling may not be a key driver in these or other related respiratory pathophysiologies.

In conclusion, our studies demonstrated the extent of glutamate as a functional vesicular glutamate transporter co-expression in central noradrenergic system. We were able to characterize a novel Vglut3-expressing NA population in the posterior C2/A2 nuclei and found a different Vglut2 co-expression pattern in anterior NA groups from a cumulative intersectional fate map compared to a viral fate map and in situ hybridization in the adult, suggesting that anterior NA populations may have a time-dependent or experience-dependent manner for Vglut2 expression. Furthermore, despite prior studies providing indirect and circumstantial evidence that noradrenergic-based glutamatergic signaling may play a role in control of breathing, our studies that remove Vglut2 expression from noradrenergic neurons failed to perturb room air breathing, the hypercapnic ventilatory reflex, and the hypoxic ventilatory reflex, which challenges the current dominant perspective and provides further insight into the potential role of NA-glutamate transmission in the control of breathing. Based on our data, we suggest that noradrenergic-based glutamatergic signaling is not required for regulating baseline breathing or the hypercapnic and hypoxic respiratory chemoreflexes, and glutamate may not be a critical mechanistic component of NA neuron dysfunction in respiratory disorders.

Materials and Methods

Ethical Approval

Studies were approved by the Baylor College of Medicine Institutional Animal Care and Use Committee (IACUC) under protocol AN-6171, and all experiments reported here were performed in accordance with relevant guidelines and regulations.

Breeding, Genetic Background, and Maintenance of Mice

We maintained all our heterozygous mouse strains by backcrossing to wildtype C57BL/6J mice and homozygous mouse strains by sibling crosses. For immunofluorescence experiments, heterozygous B6;129S-Slc17a7tm1.1(cre)Hze/J (Slc17a7Cre, Vglut1-Cre) (Jax Stock No: 023527), Slc17a6tm2(cre)Lowl/J (Slc17a6Cre, Vglut2-Cre) (Jax Stock No: 016963) and B6;129S-Slc17a8tm1.1(cre)Hze/J (Slc17a8Cre, Vglut3-Cre) (Jax Stock No: 018147) were mated with heterozygous Dbhem2.1(flpo)Rray (Dbhp2a-Flpo, DBH-p2a-Flpo) (MMRRC ID: 41575) (Sun and Ray, 2016) mice respectively to derive compound lines with both Cre and Flpo alleles (Vglut1-Cre/+; DBH-p2a-Flpo/+, Vglut2-Cre/+; DBH-p2a-Flpo/+, and Vglut3-Cre/+; DBH-p2a-Flpo/+). Then these three compound lines were each mated with homozygous B6.Cg-Gt(ROSA)26Sortm1.3(CAG-tdTomato,-EGFP)Pjen/J (Rosa26RC::FLTG, RC::FLTG) (Jax Stock No: 026932) to derive three different intersectional mouse lines Vglut1-Cre/+; DBH-p2a-Flpo/+; RC::FLTG/+, Vglut2-Cre/+; DBH-p2a-Flpo/+; RC::FLTG/+ and Vglut3-Cre/+; DBH-p2a-Flpo/+; RC::FLTG/+. For in situ hybridization experiments, wildtype C57BL/6J mice were ordered from the Center of Comparative Medicine (CCM), Baylor College of Medicine. For NA-Vglut2 conditional loss of function in situ hybridization and plethysmography experiments, hemizygous transgene Tg(Dbh-cre)KH212Gsat (Dbh-Cre, DBH-Cre) (MMRRC ID: 036778-UCD GENSAT) mice were mated with homozygous Slc17a6tm1Lowl/J (Slc17a6flox/flox) (Jax Stock No: 012898) to derive Dbh-Cre; Slc17a6flox/+. Dbh-Cre; Slc17a6flox/+ mice were mated with Slc17a6flox/flox to derive Dbh-Cre; Slc17a6flox/flox (DBH-Cre; Vglut2 cKO). Sibling mice that lacked the Cre allele or carried the Cre allele but lacked the floxed Vglut2 alleles were used as controls. Rosa26 specific primers for the RC::FLTG mice were 5ʹ-GCACTTGCTCTCCCAAAGTC, 5ʹ-GGGCGTACTTGGCATATGAT, and 5ʹ-CTTTAAGCCTGCCCAGAAGA (Ray et al., 2011) and yield a 495 bp band (targeted) and 330 bp band (wt). Cre-specific primers for all Cre drivers were 5ʹ-ATCGCCATCTTCCAGCAGGCGCACCATTGCCC and 5ʹ-GCATTTCTGGGGATTGCTTA and yielded a 550 bp band if positive. Flpo-specific primers for DBH-p2a-Flpo are 5’-CACGCCCAGGTACTTGTTCT and 5’-CCACAGCAAGAAGATGCTGA (Sun and Ray, 2016) and yielded a 226 bp band if positive. Slc17a6flox specific primers for Vglut2-floxed mice are available at The Jackson Laboratory website (https://www.jax.org/strain/012898).

Immunofluorescence staining

Vglut1, Vglut2, or Vglut3-Cre/+; DBH-p2a-Flpo/+; RC::FLTG/+ adult mice were sacrificed and transcardially perfused with 0.1 M phosphate-buffered saline (PBS) then with 4% paraformaldehyde (PFA) in PBS. Mouse brains were dissected out and fixed for 2h in 4% PFA before a PBS rinse and dehydration in 30% sucrose in PBS. Brains were embedded in OCT blocks, sectioned at the thickness of 30μm, mounted on slides, and stored at -80°C until they were ready for staining. The slides were hydrated in 0.1% Triton-X in PBS (PBST) for 15 mins, blocked with 5% donkey serum in 0.1% PBST for 1h at room temperature and then incubated with primary antibodies for 72 hours at 4°C in 0.1% PBST with 5% donkey serum. Tissues were washed in 0.1% PBST 3 times for 10 min each and then incubated with secondary antibodies for 2hs at room temperature in 0.1% PBST with 5% donkey serum. Slides were washed with 0.1% PBST for 10 min and washed in PBS twice for 10 min each, stained for DAPI, washed 3 times for 10 min each with PBS and mounted in ProLong Glass (Invitrogen). The following primary and secondary antibodies were used: chicken anti-GFP (1:1,000, Abcam ab13970), rabbit anti-dsRed (1:1,000, Clontech 632496), donkey anti-chicken Cy2 (1:500, Jackson 703-225-155), donkey anti-rabbit Cy3 (1:500, Jackson 711-165-152).

In situ hybridization

Mouse brains were dissected out from adult mice with 6-8 week of age, sectioned into 25µm brain sections and mounted on slides. We generated a digoxigenin (DIG)-labeled mRNA antisense probe against Vglut1, Vglut2, and Vglut3 and fluorescein (FITC)-labeled mRNA against DBH using reverse-transcribed mouse cDNA as a template and an RNA DIG or FITC-labeling kits from Roche (Sigma). Primer and probe sequences for the Vglut1, Vglut2, Vglut3 and DBH probes are available in the Allen Brain Atlas (http://www.brain-map.org). For the Vglut2 and DBH double ISH in DBH-Cre; Vglut2 cKO and their littermate controls, we generated a new Vglut2 probe targeting exon 2 of Vglut2 specifically (Tong et al., 2007) and the probe sequence is 892-1144bp as Slc17a6 transcript variant 1 and the size is 253bp. ISH was performed by the RNA In Situ Hybridization Core at Baylor College of Medicine using an automated robotic platform as previously described (Yaylaoglu et al., 2005) with modifications of the protocol for double ISH. Modifications in brief [see Yaylaoglu et al. (2005) for buffer descriptions]: both probes were hybridized to the tissue simultaneously. After the described washes and blocking steps the DIG-labeled probes were visualized using tyramide-Cy3 Plus (1/75 dilution, 15-minute incubation, Akoya Biosciences). After washes in TNT, the remaining HRP-activity was quenched by a 10-minute incubation in 0.2M HCl. The sections were then washed in TNT, blocked in TNB for 15 minutes, and incubated at room temperature for 30 minutes with HRP-labeled sheep anti-FITC antibody (1/500 in TNB, Roche/Sigma). Following washes in TNT, the FITC-labeled probe was visualized using tyramide-FITC Plus (1/50 dilution, 15-minute incubation, Akoya Biosciences). Following washes in TNT, the slides were stained with DAPI (invitrogen), washed again, removed from the machine, and mounted in ProLong Diamond (Invitrogen).

Viral Injection

To verify Vglut2 co-expression in LC in adult mice (Yang et al., 2021), 6-week-old Vglut2-Cre mice were injected with pAAV-EF1a-DIO-tdTomato-WPRE virus (RRID:Addgene_133786, obtained from Joshua Ortiz at the Optogenetics and Viral Vectors Core at the Jan and Dan Duncan Neurological Research Institute, 1ul at 7.90E+10 Gc/mL) into the LC (coordinates from bregma anteroposterior −5.4 mm, lateral + 0.8 mm and dorsoventral – 4.0 mm) and allowed to incubate for 4 weeks.

Plethysmography

Plethysmography on conscious and free-moving mice was carried out as described in (Ray et al., 2011) on 6-8-week-old adult animals with at least 16 animals in both the experimental and control group for each experiment. Animals were subjected to a five-day habituation protocol with each day including several minutes of handling, temperature taken by rectal probe and at least 30 min exposure in the plethysmography chamber (Martinez et al., 2019). Plethysmography was then performed for each mouse within 1 week of the last day of habituation. On the day of testing, mice were taken out from their home cage, weighed, and rectal temperature was taken. Animals were then placed into a flow-through, temperature-controlled (about 32°C with real-time temperature recording) plethysmography chamber and allowed to acclimate for at least 45 min in room air (21% O2/79% N2) conditions. After acclimation and measurement under room air, the chamber gas was switched to a hypercapnic or hypoxic mixture of 5% CO2/21% O2/74% N2, 7% CO2/21% O2/72% N2, 10% CO2/21% O2/69% N2, or 10% O2/90% N2, depending on the protocol, for 20 min. Chamber gas was then switched back to room air for another 20 min. The animals were removed from the chamber and rectal temperature was measured immediately after the mice were taken out from the chamber.

Plethysmography data analysis and statistics

Details were previously described in (Martinez et al., 2019; Ray et al., 2011). Plethysmography pressure changes were measured using a Validyne DP45 differential pressure transducer, CD15 carrier demodulator and a reference chamber, and were recorded with LabChart Pro in real time. Respiratory waveforms were analyzed by the BASSPRO module of the Breathe Easy software to determine respiratory frequency (Vf), tidal volume (VT), minute ventilation (VE), oxygen consumption (VO2) and ventilatory equivalents for oxygen (VE/VO2) (Lusk et al., 2022). A power analysis was performed using the reported effect size in Abbott et al. (2014) which used the same mouse model as we used here and 5-13 mice were necessary to observe a statistically significant result (Abbott et al. (2014) was able to see a significant difference between two groups with n=7 mice). In our experiments, the sample size for each group (DBH-Cre; Vglut2 cKO and control) exceeded 13. Results (Vf, VE, VT, VO2, VE/VO2) for room air and hypercapnic or hypoxic data were compared between DBH-Cre; Vglut2 cKO cohorts and sibling controls using a linear mixed-effects regression model with animal type (experimental vs. control) as fixed effects and animal ID as a random effect (Lusk et al., 2022). A p < 0.05 was used to indicate statistical significance, and individual data points, mean, and standard error of the mean are shown on all charts. The graphs were plotted by Prism 8.

Image quantification

Images were taken by using a Zeiss upright epifluorescent microscope and a Zeiss LSM 880 with Airyscan FAST confocal microscope. Images were captured using Zen software with z-stack function from top to bottom with 0.34μm intervals, exported, and then analyzed in Imaris using the spots and surface functions. For quantification of immunofluorescence staining, each GFP positive area coincident with DAPI (to denote nuclear localization) was defined as Vglut2 or Vglut3-expressing NA neurons while each tdTomato positive area coincident with DAPI was defined as NA neurons without any Vglut2 or Vglut3 co-expression. For quantification of Vglut2 or Vglut3 with DBH double in situ hybridization in adult WT mice, DBH positive areas coinciding with DAPI were identified as NA neurons and the DBH positive areas overlapped with Vglut2/3 positive pixels and DAPI were defined as NA neurons colocalized with Vglut2/3. The number of Vglut2/3 positive NA neurons and Vglut2/3 negative NA neurons was counted in each image for both immunofluorescence and in situ experiments and the percentage of Vglut2/3 positive NA neurons among all NA neurons in each NA nucleus in each mouse brainstem was calculated every other brain section unilaterally. For quantification of Vglut2 and DBH double ISH in DBH-Cre; Vglut2 cKO and their littermate controls, Vglut2 pixel intensities in DBH positive areas coincident with DAPI (NA neurons) were measured in mutant and control images separately. The Vglut2 pixel intensity in NA neurons of control brains was normalized as 1 and the relative Vglut2 pixel intensity of NA neurons in mutant brains compared to that in control brains was calculated. At least three mouse brains were examined for each set of experiments in each group. All quantitative results were graphed using Prism 8.

Data availability

All the data supporting this study is available from the corresponding author upon request.

eLife Assessment

Chang et al. provide glutamate co-expression profiles in the central noradrenergic system and test the requirement of Vglut2-based glutamatergic release in respiratory and metabolic activity under physiologically relevant gas challenges. Their experiments provide compelling evidence that conditional deletion of Vglut2 in noradrenergic neurons does not impact steady-state breathing or metabolic activity in room air, hypercapnia, or hypoxia. This study provides an important contribution to our understanding of how noradrenergic neurons regulate respiratory homeostasis in conscious adult mice.

Reviewer #1 (Public Review):

Summary: Chang et al. provide glutamate co-expression profiles in the central noradrenergic system and test the requirement of Vglut2-based glutamatergic release in respiratory and metabolic activity under physiologically relevant gas challenges. Their experiments show that conditional deletion of Vglut2 in NA neurons does not impact steady-state breathing or metabolic activity in room air, hypercapnia, or hypoxia. Their observations challenge the importance of glutamatergic signaling from Vglut2 expressing NA neurons in normal respiratory homeostasis in conscious adult mice.

Strengths: The comprehensive Vglut1, Vglut2, and Vglut3 co-expression profiles in the central noradrenergic system and the combined measurements of breathing and oxygen consumption are two major strengths of this study. Observations from these experiments provide previously undescribed insights into (1) expression patterns for subtypes of the vesicular glutamate transporter protein in the noradrenergic system and (2) the dispensable nature of Vglut2-dependent glutamate signaling from noradrenergic neurons to breathing responses to physiologically relevant gas challenges in adult conscious mice.

Weaknesses: Although the cellular expression profiles for the vesicular glutamate transporters are provided, the study fails to document that glutamatergic-based signaling originating from noradrenergic neurons is evident at the cellular level under normal, hypoxic, and/or hypercapnic conditions. This limits the reader's understanding of why conditional Vglut2 knockdown is dispensable for breathing under the conditions tested.

Reviewer #2 (Public Review):

The authors characterized the recombinase-based cumulative fate maps for vesicular glutamate transporters (Vglut1, Vglut2 and Vglut3) expression and compared those maps to their real-time expression profiles in central NA neurons by RNA in situ hybridization in adult mice. Authors have revealed a new and intriguing expression pattern for Vglut2, along with an entirely uncharted co-expression domain for Vglut3 within central noradrenergic neurons. Interestingly, and in contrast to previous studies, the authors demonstrated that glutamatergic signaling in central noradrenergic neurons does not exert any influence on breathing and metabolic control either under normoxic/normocapnic conditions or after chemoreflex stimulation. Also, they showed for the first-time the Vglut3-expressing NA population in C2/A2 nuclei. In addition, they were also able to demonstrate Vglut2 expression in anterior NA populations, such as LC neurons, by using more refined techniques, unlike previous studies.

A major strength of the study is the use of a set of techniques to investigate the participation of NA-based glutamatergic signaling in breathing and metabolic control. The authors provided a full characterization of the recombinase-based cumulative fate maps for Vglut transporters. They performed real-time mRNA expression of Vglut transporters in central NA neurons of adult mice. Further, they evaluated the effect of knocking down Vglut2 expression in NA neurons using a DBH-Cre; Vglut2cKO mice on breathing and control in unanesthetized mice. Finally, they injected the AAV virus containing Cre-dependent Td tomato into LC of v-Glut2 Cre mice to verify the VGlut2 expression in LC-NA neurons. A very positive aspect of the article is that the authors combined ventilation with metabolic measurements. This integration holds particular significance, especially when delving into the exploration of respiratory chemosensitivity. Furthermore, the sample size of the experiments is excellent.

Despite the clear strengths of the paper, some weaknesses exist. It is not clear in the manuscript if the experiments were performed in males and females and if the data were combined. I believe that the study would have benefited from a more comprehensive analysis exploring the sex specific differences. The reason I think this is particularly relevant is the developmental disorders mentioned by the authors, such as SIDS and Rett syndrome, which could potentially arise from disruptions in central noradrenergic (NA) function, exhibit varying degrees of sex predominance. Moreover, some of the noradrenergic cell groups are sexually dimorphic. For instance, female Wistar rats exhibit a larger LC size and more LC-NA neurons than male subjects (Pinos et al., 2001; Garcia-Falgueras et al., 2005). More recently, a detailed transcriptional profiling investigation has unveiled the identities of over 3,000 genes in the LC. This revelation has highlighted significant sexual dimorphisms, with more than 100 genes exhibiting differential expression within LC-NA neurons at the transcript level. Furthermore, this investigation has convincingly showcased that these distinct gene expression patterns have the capacity to elicit disparate behavioral responses between sexes (Mulvey et al., 2018). Therefore, the authors should compare the fate maps, Vglut transporters in males and females, at least considering LC-NA neurons. Even in the absence of identified sex differences, this information retains significant importance.

An important point well raised by the authors is that although suggestive, these experiments do not definitively rule out that NA-Vglut2 based glutamatergic signaling has a role in breathing control. Subsequent experiments will be necessary to validate this hypothesis.

An improvement could be made in terms of measuring body temperature. Opting for implanted sensors over rectal probes would circumvent the need to open the chamber, thereby preventing alterations in gas composition during respiratory measurements. Further, what happens to body temperature phenotype in these animals under different gas exposures? These data should be included in the Tables.

Is it plausible that another neurotransmitter within NA neurons might be released in higher amounts in DBH-Cre; Vglut2 cKO mice to compensate for the deficiency in glutamate and prevent changes in ventilation?

Continuing along the same line of inquiry is there a possibility that Vglut2 cKO from NA neurons not only eliminates glutamate release but also reduces NA release? A similar mechanism was previously found in VGLUT2 cKO from DA neurons in previous studies (Alsio et al., 2011; Fortin et al., 2012; Hnasko et al., 2010). Additionally, does glutamate play a role in the vesicular loading of NA? Therefore, could the lack of effect on breathing be explained by the lack of noradrenaline and not glutamate?

Reviewer #3 (Public Review):

Summary:

The authors, Y Chang and colleagues, have performed elegant studies in transgenic mouse models that were designed to examine glutamatergic transmission in noradrenergic neurons, with a focus on respiratory regulation. They generated 3 different transgenic lines, in which a red fluorophore was expressed in dopamine-B-hydroxylase (DBH; noradrenergic and adrenergic neurons) neurons that did not express a vesicular glutamate transporter (Vglut) and a green fluorophore in DBH neurons that did express one of either Vglut1, Vglut2 or Vglut3.

Further experiments generated a transgenic mouse with knockout of Vglut2 in DBH neurons. The authors used plethysmography to measure respiratory parameters in conscious, unrestrained mice in response to various challenges.

Strengths:

The distribution of the Vglut expression is broadly in agreement with other studies, but with the addition of some novel Vglut3 expression. Validation of the transgenic results, using in situ hybridization histochemistry to examine mRNA expression, revealed potential modulation of Vglut2 expression during phases of development. This dataset is comprehensive, well-presented and very useful.

In the physiological studies the authors observed that neither baseline respiratory parameters, nor respiratory responses to hypercapnea (5, 7, 10% CO2) or hypoxia (10% O2) were different between knockout mice and littermate controls. The studies are well-designed and comprehensive. They provide observations that are supportive of previous reports using similar methodology.

Weaknesses:

In relation to the expression of Vglut2, the authors conclude that modulation of expression occurs, such that in adulthood there are differences in expression patterns in some (nor)adrenergic cell groups. Altered sensitivity is provided as an explanation for different results between studies examining mRNA expression. These are likely explanations; however, the conclusion would really be definitive with inclusion of a conditional cre expressing mouse. Given the effort taken to generate this dataset, it seems to me that taking that extra step would be of value for the overall understanding of glutamatergic expression in these catecholaminergic neurons

The respiratory physiology is very convincing and provides clear support for the view that Vglut2 is not required for modulation of the respiratory parameters measured and the reflex responses tested. It is stated that this is surprising. However, comparison with the data from Abbott et al., Eur J Neurosci (2014) in which the same transgenic approach was used, shows that they also observed no change in baseline breathing frequency. Differences were observed with strong, coordinated optogenetic stimulation, but, as discussed in this manuscript, it is not clear what physiological function this is relevant to. It just shows that some C1 neurons can use glutamate as a signaling molecule. Further, Holloway et al., Eur J Neurosci (2015), using the same transgenic mouse approach, showed that the respiratory response to optogenetic activation of Phox2 expressing neurons is not altered in DBH-Vglut2 KO mice. The conclusion seems to be that some C1 neuron effects are reliant upon glutamatergic transmission (C1-DMV for example), and some not.

Further contrast is made in this manuscript to the work of Malheiros-Lima and colleagues (eLife 2020) who showed that the activation of abdominal expiratory nerve activity in response to peripheral chemoreceptor activation with cyanide was dependent upon C1 neurons and could be attenuated by blockade of glutamate receptors in the pFRG - i.e. the supposition that glutamate release from C1 neurons was responsible for the function. However, it is interesting to observe that diaphragm EMG responses to hypercapnia (10% CO2) or cyanide, and the expiratory activation to hypercapnia, were not affected by the glutamate receptor blockade. Thus, a very specific response is affected and one that was not measured in the current study.

These previous published observations are consistent with the current study which provides a more comprehensive analysis of the role of glutamatergic contributions respiratory physiology. A more nuanced discussion of the data and acknowledgement of the differences, which are not actually at odds, would improve the paper and place the information within a more comprehensive model.

Author response:

The following is the authors’ response to the previous reviews.

eLife assessment

Chang et al. provide glutamate co-expression profiles in the central noradrenergic system and test the requirement of Vglut2-based glutamatergic release in respiratory and metabolic activity under physiologically relevant gas challenges. Their experiments provide compelling evidence that conditional deletion of Vglut2 in noradrenergic neurons does not impact steadystate breathing or metabolic activity in room air, hypercapnia, or hypoxia. This study provides an important contribution to our understanding of how noradrenergic neurons regulate respiratory homeostasis in conscious adult mice.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

Chang et al. provide glutamate co-expression profiles in the central noradrenergic system and test the requirement of Vglut2-based glutamatergic release in respiratory and metabolic activity under physiologically relevant gas challenges. Their experiments show that conditional deletion of Vglut2 in NA neurons does not impact steady-state breathing or metabolic activity in room air, hypercapnia, or hypoxia. Their observations challenge the importance of glutamatergic signaling from Vglut2 expressing NA neurons in normal respiratory homeostasis in conscious adult mice.

Strengths:

The comprehensive Vglut1, Vglut2, and Vglut3 co-expression profiles in the central noradrenergic system and the combined measurements of breathing and oxygen consumption are two major strengths of this study. Observations from these experiments provide previously undescribed insights into (1) expression patterns for subtypes of the vesicular glutamate transporter protein in the noradrenergic system and (2) the dispensable nature of Vglut2-dependent glutamate signaling from noradrenergic neurons to breathing responses to physiologically relevant gas challenges in adult conscious mice.

Weaknesses:

Although the cellular expression profiles for the vesicular glutamate transporters are provided, the study fails to document that glutamatergic-based signaling originating from noradrenergic neurons is evident at the cellular level under normal, hypoxic, and/or hypercapnic conditions. This limits the reader's understanding of why conditional Vglut2 knockdown is dispensable for breathing under the conditions tested.

We thank the reviewers for their positive evaluation of our work. First, we would like to highlight that multiple studies have provided anatomical evidence of innervation of multiple cardio-respiratory nuclei by Vglut2+ noradrenergic fibers. Thus, the anatomical substrates are present for noradrenergic based Vglut2 signaling to either play a direct role in breathing control or, upon perturbation, to indirectly affect breathing through disrupted metabolic or cardiovascular control. We have included supplemental table 1 that summarizes central noradrenergic Vglut2+ innervations of respiratory and autonomic nuclei. Additionally, Ultrastructural evidence shows asymmetric synaptic contacts assuming glutamatergic transmission between C1 neurons and LC, A1, A2 and the dorsal motor nucleus of the vagus (DMV) (Milner et al., 1989; Abbott et al., 2012; Holloway et al., 2013; DePuy et al., 2013).

Functionally, electrophysiological evidence showed that photostimulating C1 neurons activate LC, A1, A2 noradrenergic neurons monosynaptically by releasing glutamate (Holloway et al., 2013; DePuy et al., 2013) and optogenetic stimulation of LC neurons excite the downstream parabrachial nucleus (PBN) neurons by releasing glutamate. Thus, at least the glutamatergic signaling from C1 and LC noradrenergic neurons (two noradrenergic nuclei that have been shown to play a role in breathing control) is evident at the cellular level under normal conditions. Other evidence, highlighted in our manuscript, is more circumstantial.

Reviewer #2 (Public Review):

The authors characterized the recombinase-based cumulative fate maps for vesicular glutamate transporters (Vglut1, Vglut2 and Vglut3) expression and compared those maps to their real-time expression profiles in central NA neurons by RNA in situ hybridization in adult mice. Authors have revealed a new and intriguing expression pattern for Vglut2, along with an entirely uncharted co-expression domain for Vglut3 within central noradrenergic neurons. Interestingly, and in contrast to previous studies, the authors demonstrated that glutamatergic signaling in central noradrenergic neurons does not exert any influence on breathing and metabolic control either under normoxic/normocapnic conditions or after chemoreflex stimulation. Also, they showed for the first-time the Vglut3-expressing NA population in C2/A2 nuclei. In addition, they were also able to demonstrate Vglut2 expression in anterior NA populations, such as LC neurons, by using more refined techniques, unlike previous studies.

A major strength of the study is the use of a set of techniques to investigate the participation of NA-based glutamatergic signaling in breathing and metabolic control. The authors provided a full characterization of the recombinase-based cumulative fate maps for Vglut transporters. They performed real-time mRNA expression of Vglut transporters in central NA neurons of adult mice. Further, they evaluated the effect of knocking down Vglut2 expression in NA neurons using a DBH-Cre; Vglut2cKO mice on breathing and control in unanesthetized mice. Finally, they injected the AAV virus containing Cre-dependent Td tomato into LC of v-Glut2 Cre mice to verify the VGlut2 expression in LC-NA neurons. A very positive aspect of the article is that the authors combined ventilation with metabolic measurements. This integration holds particular significance, especially when delving into the exploration of respiratory chemosensitivity. Furthermore, the sample size of the experiments is excellent.

Despite the clear strengths of the paper, some weaknesses exist. It is not clear in the manuscript if the experiments were performed in males and females and if the data were combined. I believe that the study would have benefited from a more comprehensive analysis exploring the sex specific differences. The reason I think this is particularly relevant is the developmental disorders mentioned by the authors, such as SIDS and Rett syndrome, which could potentially arise from disruptions in central noradrenergic (NA) function, exhibit varying degrees of sex predominance. Moreover, some of the noradrenergic cell groups are sexually dimorphic. For instance, female Wistar rats exhibit a larger LC size and more LC-NA neurons than male subjects (Pinos et al., 2001; Garcia-Falgueras et al., 2005). More recently, a detailed transcriptional profiling investigation has unveiled the identities of over 3,000 genes in the LC. This revelation has highlighted significant sexual dimorphisms, with more than 100 genes exhibiting differential expression within LC-NA neurons at the transcript level. Furthermore, this investigation has convincingly showcased that these distinct gene expression patterns have the capacity to elicit disparate behavioral responses between sexes (Mulvey et al., 2018). Therefore, the authors should compare the fate maps, Vglut transporters in males and females, at least considering LC-NA neurons. Even in the absence of identified sex differences, this information retains significant importance.

All experiments contained both males and females as described in the original submission. In our analysis of breathing and metabolism, sex was included in the analysis and no significant phenotypic difference was observed. For the fate map and in situ experiments, we did not see obvious differences in the expression patterns in the three glutamate transporters between females and males, though the group size is small. Though all the anatomical and phenotypic data in this manuscript are presented as combined graphs, we have differentially labeled our data points by sex. The reviewer does raise important questions regarding possible sexual dimorphisms in the central noradrenergic system and whether such dimorphisms may extend to glutamate transporter co-expression. Our thorough interrogation of respiratory-metabolic parameters fails to reveal any sex specific differences in control or experimental mice. Thus, it is unclear if any of the previously described and cited dimorphisms are functionally relevant in this setting. Given the large differences in the real time expression and cumulative fate maps of Vglut2, a worthwhile interrogation of differential glutamate transporter expression would be best served by longitudinal studies with large group sizes across age as it is not clear what underlies the dynamic VGlut2 expression changes. Such changes may at times be greater in males and other times in females, driven by experience or physiological challenges etc., but resulting in averaged cumulative fatemaps that are similar between sexes. Such a longitudinal quantitative study of real-time and fatemapped cell populations across the central NA system would be of a scale that is beyond the scope of this report, especially when no phenotypic changes have been observed in our respiratory data.

An important point well raised by the authors is that although suggestive, these experiments do not definitively rule out that NA-Vglut2 based glutamatergic signaling has a role in breathing control. Subsequent experiments will be necessary to validate this hypothesis.

As noted, we discuss that we only address requirement, not sufficiency, of NA Vglut2 in breathing. Functional sufficiency experiments usually involve increasing the relevant output. However, these experiments can lead to non-specific, pleiotropic effects that would be difficult to disambiguate, even if done with high cellular specificity. Viral or genetic overexpression of Vglut2 in NA neurons may be a feasible approach. Conditional ablation of TH or DBH with concurrent chemo or optogenetic stimulation may also be informative. These approaches would require significant investments in mouse model generation and suffer additional experimental limitations.

An improvement could be made in terms of measuring body temperature. Opting for implanted sensors over rectal probes would circumvent the need to open the chamber, thereby preventing alterations in gas composition during respiratory measurements. Further, what happens to body temperature phenotype in these animals under different gas exposures? These data should be included in the Tables.

While surgical implantation of sensors would provide a more direct assessment of temperature, it requires components that were not available at the time of the study and addresses a question (temperature changes during a time course of gas exposure) that go beyond the scope of the current work focused on respiratory response. As we have done for prior experiments (Martinez et al., 2019; Ray et al., 2011), the body temperature was measured immediately before and after measuring breathing only. Our flow through system using inline gas sensors (AEI P-61B CO2 sensor and AEI N-22M O2 sensor) ensure that gas challenges were constant and consistent across all measurements. Any disruption in gas composition would have been noted by our software analysis system, Breathe Easy, and the data rejected. We did not observe any such perturbations.

Is it plausible that another neurotransmitter within NA neurons might be released in higher amounts in DBH-Cre; Vglut2 cKO mice to compensate for the deficiency in glutamate and prevent changes in ventilation?

We agree that compensation is always a possibility at the synaptic, cellular, and circuit levels that may involve a variety of transcriptional, translational, cellular, and circuit mechanisms (i.e., synaptic strength). This could be interrogated by combining multiple conditional alleles and recombinase drivers for various transmitters and receptors, but would, in our experience, take multiple years for the requisite breeding to be completed.

Continuing along the same line of inquiry is there a possibility that Vglut2 cKO from NA neurons not only eliminates glutamate release but also reduces NA release? A similar mechanism was previously found in VGLUT2 cKO from DA neurons in previous studies (Alsio et al., 2011; Fortin et al., 2012; Hnasko et al., 2010). Additionally, does glutamate play a role in the vesicular loading of NA? Therefore, could the lack of effect on breathing be explained by the lack of noradrenaline and not glutamate?

These are all excellent points, but prior studies suggest that reductions in NA signaling would itself have an apparent effect (Zanella et al., 2006; Kuo et al., 2016). Although several studies showed that LC and C1 NA neurons co-release noradrenaline and glutamate, no direct evidence yet makes clear that glutamate facilitates NA release or vice versa. However, it would be of great interest to test if reduced or lack of NA compensated for loss of glutamate in the future. We do fully acknowledge that compensation in the manuscript that any number of compensatory events could be at play in these findings.

Reviewer #3 (Public Review):

Summary:

The authors, Y Chang and colleagues, have performed elegant studies in transgenic mouse models that were designed to examine glutamatergic transmission in noradrenergic neurons, with a focus on respiratory regulation. They generated 3 different transgenic lines, in which a red fluorophore was expressed in dopamine-B-hydroxylase (DBH; noradrenergic and adrenergic neurons) neurons that did not express a vesicular glutamate transporter (Vglut) and a green fluorophore in DBH neurons that did express one of either Vglut1, Vglut2 or Vglut3.

Further experiments generated a transgenic mouse with knockout of Vglut2 in DBH neurons. The authors used plethysmography to measure respiratory parameters in conscious, unrestrained mice in response to various challenges.

Strengths:

The distribution of the Vglut expression is broadly in agreement with other studies, but with the addition of some novel Vglut3 expression. Validation of the transgenic results, using in situ hybridization histochemistry to examine mRNA expression, revealed potential modulation of Vglut2 expression during phases of development. This dataset is comprehensive, wellpresented and very useful.

In the physiological studies the authors observed that neither baseline respiratory parameters, nor respiratory responses to hypercapnea (5, 7, 10% CO2) or hypoxia (10% O2) were different between knockout mice and littermate controls. The studies are well-designed and comprehensive. They provide observations that are supportive of previous reports using similar methodology.

Weaknesses:

In relation to the expression of Vglut2, the authors conclude that modulation of expression occurs, such that in adulthood there are differences in expression patterns in some (nor)adrenergic cell groups. Altered sensitivity is provided as an explanation for different results between studies examining mRNA expression. These are likely explanations; however, the conclusion would really be definitive with inclusion of a conditional cre expressing mouse. Given the effort taken to generate this dataset, it seems to me that taking that extra step would be of value for the overall understanding of glutamatergic expression in these catecholaminergic neurons

The seemingly dynamic Vglut2 expression pattern across the NA system is intriguing. As noted in our comments to reviewer 2, a robust age dependent interrogation would require a large magnitude study. The reviewer correctly points out that a temporally controlled recombinase fate mapping experiment would offer greater insight into the dynamic expression of Vglut2. We strongly agree with that idea and did work to develop a Vglut2-CreER targeted allele that, despite our many other successes in mouse genetic engineering (Lusk et al., 2022; Sun and Ray, 2016), did not succeed on the first attempt. We aim to complete the line in the near future so that we may better understand the Vglut2 expression pattern in central noradrenergic neurons in a time-specific manner and sex specific manner.

The respiratory physiology is very convincing and provides clear support for the view that Vglut2 is not required for modulation of the respiratory parameters measured and the reflex responses tested. It is stated that this is surprising. However, comparison with the data from Abbott et al., Eur J Neurosci (2014) in which the same transgenic approach was used, shows that they also observed no change in baseline breathing frequency. Differences were observed with strong, coordinated optogenetic stimulation, but, as discussed in this manuscript, it is not clear what physiological function this is relevant to. It just shows that some C1 neurons can use glutamate as a signaling molecule. Further, Holloway et al., Eur J Neurosci (2015), using the same transgenic mouse approach, showed that the respiratory response to optogenetic activation of Phox2 expressing neurons is not altered in DBH-Vglut2 KO mice. The conclusion seems to be that some C1 neuron effects are reliant upon glutamatergic transmission (C1DMV for example), and some not.

We agree that activation of C1 neurons may be sufficient to modulate breathing when artificially stimulated and that such stimulation relies on glutamatergic transmission for its effect. This is why we find our results surprising and important in clarifying for the field that glutamatergic signaling in noradrenergic cells is dispensable for breathing and hypoxic and hypercapnic responses under physiological conditions.

Further contrast is made in this manuscript to the work of Malheiros-Lima and colleagues (eLife 2020) who showed that the activation of abdominal expiratory nerve activity in response to peripheral chemoreceptor activation with cyanide was dependent upon C1 neurons and could be attenuated by blockade of glutamate receptors in the pFRG - i.e. the supposition that glutamate release from C1 neurons was responsible for the function. However, it is interesting to observe that diaphragm EMG responses to hypercapnia (10% CO2) or cyanide, and the expiratory activation to hypercapnia, were not affected by the glutamate receptor blockade. Thus, a very specific response is affected and one that was not measured in the current study.

As we mention above, we do not dispute that glutamate signaling can be manipulated to create a response in non-physiological conditions – we suggest that framing the interpretation around the glutamatergic role in a model that better matches physiological conditions should inform our interpretation. Furthermore, we do include an examination of expiratory flow – which was not impacted by loss of glutamatergic activity in NA neurons – which would be likely to have been impacted if abdominal expiratory nerve activity was modified.

These previous published observations are consistent with the current study which provides a more comprehensive analysis of the role of glutamatergic contributions respiratory physiology. A more nuanced discussion of the data and acknowledgement of the differences, which are not actually at odds, would improve the paper and place the information within a more comprehensive model.

Thank you for the comments. As noted in the original and extended discussion, we respectfully disagree with the perspective that our results align with prior results.

Recommendations for the authors:

The three reviewers believe this is an important study. They have numerous suggestions for improvement of the manuscript (outlined below), but no new experiments are required. The Editor requests some nomenclature changes as indicated in attachment 1.

Reviewer #1 (Recommendations For The Authors):

Abstract/Introduction: Although the need for this study is obvious, it is important that the authors explicitly communicate their working hypothesis < before the start of the work> to the reader. In the current form, it is unclear whether the authors aimed to test the hypothesis that glutamatergic signaling from noradrenergic neurons is important to breathing or whether to test the hypothesis that glutamatergic signaling from noradrenergic neurons is not important to breathing. If it is the latter-it is not important-then the study (related to the breathing measurements) is poorly justified and designed, as additional orthogonal approaches (e.g., actual measurements of glutamatergic signaling at the cellular level) are almost requisite. If the authors' hypothesis was originally based on existing literature suggesting that glutamatergic signaling from noradrenergic neurons is important to breathing, then the experimental design appropriate.

Thank you for the suggestion. The working hypothesis has been added in the abstract (line 2425) and the introduction (line 92-94)), making clear that we initially hypothesized that glutamatergic signaling from noradrenergic neurons is important in breathing.

Results: While the steady state measurements for breathing metrics are clearly important in defining how glutamatergic signaling may contribute to be pulmonary function, the role of glutamatergic signaling may have a greater role in the dynamics of patterns (i.e., regularity of the breathing rhythms) such traits can be described using SD1 and SD2 from Poincare maps, and/or entropy measurements. Such an analysis should be performed.

Thank you for the suggestion. The dynamic patterns of respiratory rate (Vf), tidal volume (VT), minute ventilation (VE), inspiratory duration (TI), expiratory duration (TE), breath cycle duration (TTOT), inspiratory flow rate (VT/TI), expiratory flow rate (VT/TE) have been shown as Poincaré plots and quantified and tested using the SD1 and SD2 statistics in the supplemental figures of Figure 4-7.

Results: Analyses of Inspiratory time (Ti) and flow rate (i.e., Tidal Volume / Ti) should be assessed and included.

Thank you for the suggestion. Inspiratory duration (Ti), expiratory duration (TE), breath cycle duration (TTOT), inspiratory flow rate (VT/Ti), and expiratory flow rate (VT/TE) have been included in the Figures 4-7.

Results/Methods: If similar analytical approaches were used in the current study as to that in Lusk et al. 2022, it appears that data was discontinuously sampled, rejecting periods of movement and only including periods of quiescent breathing. Were the periods of quiescent breathing different? Information should be provided to describe the total sampling duration included.

For room air, the entire gas condition was used for data analysis. For hypercapnia (5% CO2, 7% CO2, 10% CO2), only the last 5 minutes of the gas challenge period was used for data analysis. For hypoxia (10% O2), we analyzed the breathing trace of three 5-minute epochs following initiation of the gas exposure separately, e.g., epoch 1 = 5-10min, epoch 2 = 10-15min, and epoch 3 = 15-20min. All breaths included as quiescent breathing were analyzed in the aggregate for each group and experimental condition, we did not compare individual periods of quiescent breathing within or across an animal(s)/group(s)/experimental condition(s). We have added the details in the Materials and Methods (line 637-642).

Results: As mice were conscious in this study, were sniff periods (transient periods of fast breathing, i.e.,>8Hz) included in the analysis?

No, only regular quiescent breathing periods were included in the analysis.

Discussion: The authors need to discuss the limitations of their findings.

How should the reader interpret the findings? Concluding that glutamatergic signaling is dispensable implies that it occurs in room air, hypoxia, and hypercapnia.

We have edited our discussion for clarity to highlight our conclusions that Vglut2-based glutamatergic signaling from noradrenergic neurons is ultimately dispensable for baseline breathing and hypercapnia and hypoxic chemoreflex in unanesthetized and unrestrained mice.

Assuming that glutamatergic signaling is active during the conditions tested, then the authors should discuss what may be the potential compensations.

We have provided additional discussion surrounding potential compensatory events that may have taken place and could result in the unchanged phenotype in the experimental group.

The authors need to discuss how age and state of consciousness may play a role in their finds. The current discussion gives the impression that their findings are broadly applicable in all cases, but the lack of differences in this study may not hold true under different conditions.

The study was done in adult (6–8-week-old) unanesthetized and unrestrained mice. In the discussion (line 472-474), we highlight that in our unpublished results, loss of NA-expressed Vglut2 does not change the survival curve in P7 neonate mice undergoing repeated bouts of autoresuscitation until death. Thus, we believed that Vglut2-based glutamatergic signaling in central NA neurons is dispensable for baseline breathing and the hypercapnic and hypoxic chemoreflexes in unanesthetized and unrestrained mice across different ages. Otherwise, we do not imply that we have interrogated any other aspects of breathing in our discussion.

Methods: Further description of the analysis window for the respiratory metrics should be provided. Were breath values for each condition taken throughout the entire condition? This is particularly important for hypoxia, where the stereotypical respiratory response is biphasic.

For room air, the entire gas condition was used for data analysis. For hypercapnia (5% CO2, 7% CO2, 10% CO2), only the last 5min of the gas challenge period was used for data analysis. For hypoxia (10% O2), we analyzed the breathing trace of three 5min time periods separately including 5-10min, 10-15min, and 15-20min during the hypoxic challenge as noted in our original manuscript, we graph and assess three 5min epochs during hypoxic exposure to capture the dynamic nature of the hypoxic ventilatory response. We have added the details in the Materials and Methods (line 637-642).

Methods: How was consciousness determined?

The conscious mice mentioned in the manuscript refer to the mice without anesthesia. We have replaced “awake” and “conscious” with “unanesthetized” in the text.

Reviewer #2 (Recommendations For The Authors):

Since no EEG/EMG recording was performed it would be more appropriate to remove "awake" and "conscious" throughout the manuscript and include the term "unanesthetized".

Thank you for the suggestion. “Awake” and “conscious” have been replaced by “unanesthetized” in the text.

Line 545: Why 32C? Isn't this temperature too high for animals?

30-32°C is the thermoneutral zone for mice. It is the range of ambient temperature where mice can maintain a stable core temperature with their minimal metabolic rate (Gordon, 1985). Whole-body plethysmography uses the barometric technique to detect pressure oscillations caused by changes in temperature and humidity with each breathing act when an animal sits in a sealed chamber (Mortola et al., 2013). Thus, maintaining the chamber temperature near the thermoneutral zone during the plethysmography assay is required to maintain constancy in respiratory and metabolic parameters from trial to trial as well as to maintain linearity of ventilatory pressure changes due to humidification, rarefaction, and thermal expansion and contraction during inspiration and expiration (Ray et al., 2011). The chamber temperature that has been used for adult plethysmography has been set across a range 30-34°C (Hodges et al., 2008; Ray et al., 2011; Hennessy et al., 2017). We use 32°C in this manuscript which is consistent with previously published literature from other groups and our own work (Sun et al., 2017; Lusk et al., 2022).

I would include the units of the physiological variables in the tables.

Thank you for the suggestion. The units of the physiological variables have been added in all the tables.

Reviewer #3 (Recommendations For The Authors):

Why is the C3 group not considered in this study?

The C3 adrenergic group, best characterized in rat, is only seen in rodents but not in many other species including primates (including human) (Kitahama et al., 1994). Thus, the C3 group is not the focus of this study where we aim to discuss if glutamate derived from noradrenergic neurons could be the potential therapeutic target of human respiratory disorders. The C3 adrenergic group is typically described as a population containing only about 30 neurons. We have added the fate map data and the adult expression pattern for the three vesicular glutamate transporters for the C3 group in the figure 1 and 2 supplements for reference.

Sub CD/CV does not appear to be defined in the manuscript.

Thank you for the point. The definition of sub CD/CV has been added in the text (line 126).

The data on line 131-133 is interesting but could be described more effectively and clearly.

Thank you for the suggestion. The text has been modified accordingly.

The end of the paragraph at lines 140 onwards is rather repeated in the paragraph that starts at line 146.

The repeated text has been removed accordingly.

Whilst anterior and posterior are correct anatomical terms, for a quadraped, rostral and caudal are more widely used - particularly in the brainstem field. Is there a particular reason for using anterior/posterior?

We followed the anatomical terminations in the Robertson et al. (2013) where they used anterior/posterior to describe C2/A2 and C1/A1.

On the protocol lines include in Figure 4-7 it would be worth adding the test day. This seems a little strange. Why wait up to one week after the habituation to perform the stimulation. How many mice were left for each day between habituation and experimentation, and does this timing affect responses? Do mice forget the habituation after a period?

Thank you for the point. We have added the test day for plethysmography in figures 4-7. After the 5 days of habituation, we began the plethysmography recordings on the sixth day. A maximum of 6 mice can be assayed for plethysmography per day due to the limited number of barometric flow through plethysmography and metabolic measurement systems we have. Thus, all animals were finished with plethysmography “within” one week of the last day of habituation. This protocol is consistent with our previous published work (Martinez et al., 2019; Lusk et al., 2022; Lusk et al., 2023). For the experiments in this manuscript, mice were assayed within 3 days after habituation. As noted in our methods and figures, each mouse is given as much as 40 mins to acclimate to the chamber (determined by directly observed quiet breathing) before data acquisition. We have no reason or evidence that indicates testing order and thus timing was a factor. The detailed explanation for the plethysmography protocol has been added in the material and methods section (line 606-625).

Please state clearly that each mouse is only exposed to one gas mixture (what I interpret is the case), or could one mouse be exposed to several different stimuli?

Each mouse is only exposed to one gas challenge (5% CO2, 7% CO2, 10% CO2, or 10% O2) in a testing period. Each testing period for an individual mouse was separated by 24hs to allow for a full recovery. The protocol is to put the mouse under room air for 45mins, switch to one gas challenge for 20mins, and switch back to room air for 20mins.

With apologies if I missed this, but did each of the respiratory stimuli produce a statistically significant response in the control mice? For example, the response to 10%O2?

Yes, each respiratory stimuli including 5/7/10% CO2 and 10% O2 produced a statistically significant response in both mutant and control mice. We have labeled the statistical significance in the Figures 4-7. Thank you for pointing this out.

Line 312: Optogenetic stimulation induced an increase from 130 to 180 breaths per min (Abbott et al., EJN 2014). It is surprising that this is called "modest". Baseline respiratory frequency was presented.

Thank you for the point. The word “modest” has been removed and the discussion has been changed accordingly (line 355-360).

Line 338: This discussion is not sufficiently nuanced. It is the increased Dia amplitude (to KCN only, not 10%CO2 ) and the stimulation of active expiration, to both stimuli, that is blocked by kyn in pFRG. There is no effect of breathing frequency. The current study would not detect such differences in active expiration.

Thank you for the suggestion. The discussion has been modified accordingly (line 382-388).

Version 2

Abstract

Abstract

Central noradrenergic (NA) neurons are key constituents of the respiratory homeostatic network. NA dysfunction is implicated in several developmental respiratory disorders including Congenital Central Hyperventilation Syndrome (CCHS), Sudden Infant Death Syndrome (SIDS) and Rett Syndrome. The current unchallenged paradigm in the field, supported by multiple studies, is that glutamate co-transmission in subsets of central NA neurons plays a role in breathing control. If true, NA-glutamate co-transmission may also be mechanistically important in respiratory disorders. However, the requirement of NA-derived glutamate in breathing has not been directly tested and the extent of glutamate co-transmission in the central NA system remains uncharacterized. Therefore, we fully characterized the cumulative fate maps and acute adult expression patterns of all three Vesicular Glutamate Transporters (Slc17a7 (Vglut1), Slc17a6 (Vglut2), and Slc17a8 (Vglut3)) in NA neurons, identifying a novel, dynamic expression pattern for Vglut2 and an undescribed co-expression domain for Vglut3 in the NA system. In contrast to our initial hypothesis that NA derived glutamate is required to breathing, our functional studies showed that loss of Vglut2 throughout the NA system failed to alter breathing or metabolism under room air, hypercapnia, or hypoxia in unrestrained and unanesthetized mice. These data demonstrate that Vglut2-based glutamatergic signaling within the central NA system is not required for normal baseline breathing and hypercapnic, hypoxic chemosensory reflexes. These outcomes challenge the current understanding of central NA neurons in the control of breathing and suggests that glutamate may not be a critical target to understand NA neuron dysfunction in respiratory diseases.

Introduction

Breathing is a vital and life-sustaining function supporting homeostatic processes, most critically maintaining blood pH/CO2 and O2 levels within a narrow physiological range. Respiratory homeostasis is mediated by neuron-modulated lung ventilation adjustments in response to physiological deviations resulting in high pCO2 or low pO2 blood and tissue levels, known as the hypercapnic and hypoxic reflexes respectively (Del Negro et al., 2018; Dick et al., 2018). These reflexes are part of a complex brainstem neural network that integrates a multitude of information streams across the central and peripheral nervous systems to regulate respiratory output. Within this brainstem network, central noradrenergic (NA) neurons are known to be an important component that plays a variety of roles in modulating breathing. Furthermore, various perturbations across the central NA system have been implicated in several developmental disorders with respiratory and chemosensory features such as Congenital Central Hyperventilation Syndrome (CCHS), Sudden Infant Death Syndrome (SIDS) and Rett Syndrome (Beltrán-Castillo et al., 2017; Feldman et al., 2013; Gauda et al., 2007; Viemari, 2008). Thus, understanding how central NA neurons modulate respiratory chemoreflexes or chemosensory breathing is critically important for the development of new diagnostic and therapeutic interventions to address respiratory pathophysiology.

Central noradrenergic neurons are commonly thought to exert their effect on breathing through their primary neurotransmitter noradrenaline and adrenaline. However, in addition to noradrenaline, several studies provide strong evidence that glutamate is co-transmitted in subsets of central NA neurons. Vesicular glutamate transporter 2 (Vglut2), a gene marker of glutamatergic signaling, has been shown to be co-expressed in subsets of central NA neurons including, C1/A1, C2/A2, A5, and LC in adult rats and mice (DePuy et al., 2013; Souza et al., 2022a; Stornetta et al., 2002a, 2002b; Yang et al., 2021), though NA specific expression of related Vglut1 and Vglut3 transporters remains unknown. Additionally, it has been well documented that central NA neurons co-expressing Vglut2 innervate key respiratory centers, such as preBötzinger complex and parafacial region (pFRG). Additionally, central NA neurons co-expressing Vglut2 project to many key autonomic brainstem, spinal cord, and forebrain centers, such as the dorsal motor nucleus of the vagus, intermediolateral nucleus (IML) and sympathetic system, and the hypothalamus, all of which could drive a change in breathing if perturbed, i.e., indirect effects from cardio-vascular and metabolic dysregulation (Supplemental table 1). Thus, it has become a dominant paradigm in the field of respiratory physiology that NA-based glutamate release is a critical form of neurotransmission in the control of breathing. Abbott et al. (2014) showed that Vglut2 is required for an increase in respiratory rate when anterior C1 neurons are unilaterally opto-genetically stimulated. Malheiros-Lima et al. (2020) showed that Vglut2-expressing C1 neurons project to the pFRG region, and hypoxic breathing was blunted after blockade of ionotropic glutamatergic receptors at the pFRG site in anesthetized rats, together supporting a role for anterior C1 neurons releasing glutamate at the pFRG site to regulate breathing under hypoxia. Similarly, Malheiros-Lima et al. (2022) and Malheiros-Lima et al. (2018) showed that Vglut2-expressing C1 neurons project to the NA A5 region and the preBötzinger complex, and, again, the blockade of ionotropic glutamatergic receptors at the A5 region or preBötzinger complex reduced the increase in phrenic nerve activity and respiratory frequency elicited by optogenetic stimulation of C1 cells in an anesthetized preparation. In addition, Guyenet et al. (2013) and others speculated that the apparent lack of plasmalemmal monoamine transporter in C1 fibers indicates reduced or absent noradrenergic or adrenergic signaling due a lack of re-uptake and neurotransmitter pool depletion (Comer et al., 1998; Lorang et al., 1994). Cumulatively, the studies argue for glutamate as the predominant functional neurotransmitter for C1 NA neurons in the breathing neural network. To our knowledge, this dominant perspective has not been otherwise previously challenged. Although these studies are informative, the evidence supporting the role of NA-based Vglut2 signaling in respiratory control are either indirect and circumstantial or cannot be seen as physiological given experimental limitations, such as the focal nature of opto-genetic stimulation. Thus, it is not yet clear what the requirement is for NA-based glutamatergic signaling in homeostatic breathing in the unanesthetized and unrestrained animal and how that might inform upon disease.

To better understand the role of NA-based glutamatergic signaling in breathing, we sought to both fully characterize the molecular profiles of the central noradrenergic system with respect to glutamate co-expression and to test the hypothesis that Vglut2-based glutamatergic release is required in respiratory control under physiological chemosensory challenges in unanesthetized and unrestrained mice. We first fully characterized the recombinase-based cumulative fate maps for Vglut1, Vglut2 and Vglut3 expression and compared those maps to their real time expression profiles in central NA neurons by RNA in situ hybridization in adult mice. We found a novel dynamic expression pattern for Vglut2 and an entirely undescribed co-expression domain for Vglut3 in the central NA system. Secondly, to determine if Vglut2-based glutamatergic signaling in NA neurons is required for respiratory homeostasis, we conditionally ablated Vglut2 in all NA neurons and tested respiratory, chemosensory, and metabolic function in unrestrained and unanesthetized mice. Using the same genetic model in prior studies (Abbott et al., 2014), conditional deletion of Vglut2 in NA neurons did not significantly impact breathing under room air, hypercapnic, or hypoxic conditions. These results demonstrate, for the first time, that NA Vglut2-based glutamatergic signaling is dispensable for respiratory control, which challenges the prevalent perspective on the role of C1 and other Vglut2 expressing NA neurons in respiratory homeostasis and suggests that glutamate may not be a critical target to understand NA neuron dysfunction in respiratory diseases.

Results

Cumulative fate maps of central noradrenergic neurons co-expressing Vglut1, Vglut2 and Vglut3

To fully characterize the expression profiles of all three glutamate markers Vglut1, Vglut2 and Vglut3 in central noradrenergic neurons, we first used an intersectional genetic strategy (Figure 1A). We bred three Cre drivers Slc17a7Cre(Vglut1-Cre), Slc17a6Cre (Vglut2-Cre) and Slc17a8Cre(Vglut3-Cre) to Dbhp2a-Flpo (DBH-p2a-Flpo) (targeting noradrenergic neurons) mice, respectively.

Cumulative fate maps of glutamate co-expressing NA neurons characterized by intersectional genetics.

(A) Breeding schematic to generate the three intersectional reporter lines Slc17a7Cre; Dbhp2a-Flpo; Rosa26RC::FLTG, Slc17a6Cre; Dbhp2a-Flpo; Rosa26RC::FLTG, Slc17a8Cre; Dbhp2a- Flpo; Rosa26RC::FLTG. In each of the three intersectional reporter crosses, NA neurons co-expressing either Vglut1, Vglut2, or Vglut3 are labeled by green fluorescent protein (eGFP) while NA neurons without any Vglut1,2, or 3 expressions are labelled by red fluorescent protein (tdTomato). Shown as an example is the Slc17a6Cre; Dbhp2a-Flpo; Rosa26RC::FLTG (Vglut2-Cre/+; DBH-p2a-Flpo/+; RC::FLTG/+) intersectional reporter line. (B) Quantification of the percentage of Vglut2-coexpressing NA neurons among NA neurons in each brainstem NA nuclei including A7, LC, A5, sub CD/CV, anterior C1/A1 and C2/A2, posterior C1/A1 and C2/A2. Pink data points represent female data while blue data points represent male data. (C) Quantification of the percentage of Vglut3-coexpressing NA neurons among NA neurons in each NA nuclei. Pink data points represent female data while blue data points represent male data. (D) Fluorescent expression of tdTomato (red) and eGFP (green) in coronal sections of Vglut1, Vglut2, or Vglut3 intersectional reporter lines in brainstem NA nuclei. Nucleus is labeled by DAPI (blue). Scale bar 50μm.

The three compound lines of Slc17a7Cre; Dbhp2a-Flpo, Slc17a6Cre; Dbhp2a-Flpo, and Slc17a8Cre; Dbhp2a- Flpo were then crossed with the Rosa26RC::FLTG (RC::FLTG). RC::FLTG are intersectional reporter mice that express tdTomato in cells expressing only flippase (Flpo) and express eGFP in cells co-expressing both Flpo and Cre recombinases. Thus, in each of the three intersectional reporter crosses [e.g. 1) Slc17a7Cre; Dbhp2a-Flpo; Rosa26RC::FLTG, 2) Slc17a6Cre; Dbhp2a-Flpo; Rosa26RC::FLTG and 3) Slc17a8Cre; Dbhp2a-Flpo; Rosa26RC::FLTG], NA neurons without any Vglut1, 2, or 3 expression are labelled by red fluorescent protein (tdTomato) while NA neurons co-expressing either Vglut1, Vglut2, or Vglut3 are labeled by green fluorescent protein (eGFP). By using this method, we characterized and quantified the expression profiles of all three vesicular glutamate transporters Vglut1, Vglut2, and Vglut3 across every anatomically defined NA nucleus in adult mice including A7, Locus Coeruleus (LC), dorsal/ ventral subcoeruleus nucleus (sub CD/CV), A5, and the anterior - posterior dimensions of C1/A1, C2/A2 (Robertson et al., 2013), and C3 (Figure 1B-D and Figure 1 supplement 1A-B). Vglut1-Cre was not co-expressed in any central NA nuclei in adult mice. However, Vglut2-Cre and Vglut3-Cre both showed co-expression in central NA neurons. Vglut3-Cre-expressing NA neurons were restricted to posterior C2/A2 and posterior C1/A1, with greatest expression in posterior C2/A2 where 26.9 ± 3.16% of NA neurons were Vglut3-Cre positive. In posterior C1/A1, only 1.26 ± 0.559% of NA neurons were Vglut3-Cre positive. Surprisingly, 84.6 ± 3.75% of NA neurons in total showed Vglut2-Cre co-expression and each NA nucleus was predominantly labelled by Vglut2-Cre expression. Over 50% of NA neurons in each NA nucleus were Vglut2-Cre positive. The percentages of Vglut2-Cre positive NA neurons in each NA nucleus were as follows: 69.8 ± 8.29% in A7, 80.5 ± 3.78% in LC, 84.8 ± 11.3% in A5, 51.5 ± 2.61% in sub CD/CV, 99.0 ± 0.681% in anterior C1/A1, 95.7 ± 2.40% in posterior C1/A1, 96.1 ± 1.98% in anterior C2/A2, 97.8 ± 0.771% in posterior C2/A2, and 100 ± 0.00% in C3 (Mean ± SEM). The presence of Vglut2-Cre co-expression in anterior NA groups was unexpected as previous in situ data in adult rats found Slc17a6 (Vglut2) positive NA neurons only in the posterior C2/A2 and C1/A1 (Stornetta et al., 2002a, 2002b). Notably, however, the Vglut2 co-expression in the LC region agrees with Yang et al. (2021) which also showed that 89% of LC NA neurons are Vglut2 positive using a different intersectional strategy in mice: ThFlpo; Slc17a6Cre; Rosa26Ai65 (TH-Flpo; Vglut2-Cre; Ai65). The differences in our and other fate maps compared to in situ hybridization may either reflect early gene expression that is down regulated in the adult or may reflect low levels of expression not detectable by in situ hybridization but that are nonetheless sufficient to affect recombination in the intersectional genetic strategy.

Real time mRNA expression patterns of Vglut1, Vglut2 and Vglut3 in central noradrenergic neurons in adult mice

To verify if the real time NA-based Vglut2 expression pattern in adult mice is comparable to previous reports in adult rats (DePuy et al., 2013; Stornetta et al., 2002b, 2002a), and to further characterize the expression patterns of the other two vesicular glutamate transporters (Vglut1 and Vglut3) in adult mice, we performed fluorescent RNA in situ hybridization experiments. We co-stained Slc17a7 (Vglut1), Slc17a6 (Vglut2), or Slc17a8 (Vglut3) with Dbh respectively in brain tissue of adult mice and characterized the colocalization of Slc17a7, Slc17a6, Slc17a8 with Dbh in brainstem NA nuclei A7, LC, A5, sub CD/CV, anterior C1/A1 and C2/A2, posterior C1/A1 and C2/A2, and C3 (Figure 2A-C and Figure 2 supplement 2A-B). No Slc17a7/Dbh double positive neurons were detected in any part of the central NA system, consistent with our fate map. Slc17a8 mRNA colocalization with Dbh was only found in the posterior part of C2/A2 where 27.1 ± 1.86% of Dbh positive neurons demonstrated Slc17a8 colocalization, again consistent with our fate map. However, we only found detectable levels of Slc17a6 mRNA in the C1/A1, C2/A2 and C3 NA regions, but not in the A7, LC, A5, sub CD/CV. The percentages of Slc17a6/Dbh double positive neurons for regions where detected were as follows: 84.7 ± 5.82% in anterior C1/A1, 66.3 ± 0.335% in anterior C2/A2, 35.7 ± 3.01% in posterior C1/A1, 90.1 ± 2.45% in posterior C2/A2, and 79.7 ± 3.94% in C3. This result is consistent with the previous in situ data in adult rats (Stornetta et al., 2002a, 2002b) suggesting that there is no obvious expression difference of Slc17a6 in the central NA system between mouse and rat. However, the Slc17a6 in situ data did not show expression in anterior NA populations. This difference between our in situ data and fate map data supports our previous hypothesis that many NA neurons expressed Slc17a6 (Vglut2) at some point from early development toward adulthood but the expression of Slc17a6 (Vglut2) is diminished during adulthood. Interestingly, Yang et al. (2021) did show adult Vglut2 co-expression in LC by a viral injection of both Cre and Flpo-dependent eYFP into the LC of a bi-transgenic mouse with both ThFlpoand Slc17a6Cre. To verify this result, we injected an AAV virus containing Cre-dependent tdTomato (pAAV-EF1a-DIO-tdTomato-WPRE) into the LC region of adult Slc17a6Cre mice. Consistent with Yang et al. (2021), we observed sparse Vglut2-Cre and TH-immunopositive double labelled neurons in the LC (Figure 2 supplement 1). Additionally, Souza et al. (2022a) showed about 28.5% of A5 neurons are Slc17a6 mRNA positive in adult rats by RNA scope, a method which is more sensitive than the traditional in situ hybridization we used here. These data suggest that some adult NA neurons in LC and A5 have Slc17a6 (Vglut2) co-expression, but the Slc17a6 mRNA level cannot be detected by traditional in situ hybridization.

Characterization of real time co-expression of Vglut1, Vglut2 and Vglut3 in brainstem NA neurons in adult mice by fluorescent in situ hybridization.

(A) Representative images of Slc17a7 (Vglut1), Slc17a6 (Vglut2), and Slc17a8 (Vglut3) and Dbh (DBH) double ISH in coronal sections of WT mice in all of brainstem NA nuclei in adult mice. DBH (green), Vglut1/2/3 (red), DAPI (cyan). Scale bar 50μm. (B) Quantification of the percentage of Slc17a6 (Vglut2) co-expression in NA neurons in each NA nuclei. Female data (pink), male data (blue). (C) Quantification of the percentage of Slc17a8 (Vglut3) co-expression in NA neurons in each NA nuclei. Female data (pink), male data (blue).

Vglut2 expression is effectively knocked down in the whole central noradrenergic system in Dbh-Cre; Vglut2 cKO mice

To investigate the requirement of Vglut2-based glutamatergic signaling in the central NA system in breathing under physiological challenges, we used Dbh-Cre; Slc17a6flox/flox(Dbh-Cre; Vglut2 cKO) to remove Vglut2 expression from all noradrenergic neurons. This is the same model used by Abbott et al. (2014) to show that Vglut2 expression in the NA system was required for an increase in respiratory frequency following unilateral optogenetic stimulation of anterior C1. To verify if Vglut2 expression was effectively removed from central NA neurons in Dbh-Cre; Vglut2 cKO mice, we used fluorescent mRNA in situ hybridization for Slc17a6 (Vglut2) and Dbh in the mouse brainstems and compared the Slc17a6 (Vglut2) signal intensity in Dbh positive neurons in Dbh-Cre; Vglut2 cKO mice to their littermate controls. We found that Slc17a6 mRNA expression was 96.4 ± 0.226% decreased compared to controls, indicating that the Slc17a6 (Vglut2) was effectively recombined to abrogate expression in NA neurons (Figure 3A-B and Figure 3 supplement 1).

DBH with Vglut2 co-staining by fluorescent in situ hybridization confirms that Vglut2 expression is disrupted in the whole central NA system in Dbh-Cre; Vglut2 cKO mice.

(A) Representative images of Slc17a6 (Vglut2) and Dbh (DBH) double ISH in anterior C1/A1, posterior C1/A1, anterior C2/A2 and posterior C2/A2 in control and Dbh-Cre; Vglut2 cKO mice. DBH (green), Vglut2 (red), DAPI (cyan). Scale bar 20μm.(B) Quantification of intensity of Slc17a6 (Vglut2) signal in control and Dbh-Cre; Vglut2 cKO mice.

Vglut2-based glutamatergic signaling in central NA neurons is not required for baseline breathing nor for the hypercapnic ventilatory reflex under distinct CO2 challenges

Multiple prior studies provide indirect evidence that NA glutamate transmission may play roles in respiratory function, particularly that anterior C1 neurons mediate the hypoxic ventilatory response through the pFRG/RTN and/or the NA A5 group (Malheiros-Lima et al., 2022, 2020). Additionally, the A5 group has been implicated in the hypercapnic reflex (Haxhiu et al., 1996). Thus, we sought to determine if Vglut2-based glutamatergic signaling in central NA neurons is necessary to regulate baseline breathing and the hypercapnic chemoreflex in unanesthetized and unrestrained animals. We used whole-body barometric plethysmography to measure the breathing and metabolism of unanesthetized and unrestrained mice. After a five-day habituation protocol, the respiratory responses of Dbh-Cre; Vglut2 cKO mice and their littermate controls, including the overall respiratory output or ventilatory equivalent of oxygen (VE/VO2), respiratory rate (Vf), tidal volume (VT), minute ventilation (VE), metabolic demand (VO2), inspiratory duration (TI), expiratory duration (TE), breath cycle duration (TTOT), inspiratory flow rate (VT/TI), and expiratory flow rate (VT/TE), were measured under room air and hypercapnia (5% CO2) (Figure 4A). 5% CO2 gas challenge produced a significant change in all the respiratory parameters compared to room air in both Dbh-Cre; Vglut2 cKO and control mice. However, surprisingly, Dbh-Cre; Vglut2 cKO mice did not show significant differences in any respiratory or metabolic parameter mentioned above compared to their littermate controls under both room air and 5% CO2 challenges (Figure 4B and Table 1-2). In addition, to investigate if Vglut2-based NA derived glutamate is required for regulating the dynamic patterns of breathing, such as the breathing rhythms, we generated Poincaré plots for TI, TE, TTOT, VT/TI, VT/TE, Vf, VT, and VE and we calculated the SD1 and SD2 statistics for each parameter under room air and 5% CO2. We found that Dbh-Cre; Vglut2 cKO mice only showed a marginally statistically significant increase in the SD1 of inspiratory duration (TI) (p=0.0492). Outside of this single comparison, however, we found no significant differences in the other parameters between the two groups (Figure 4 supplement 4A-H). Thus, we theorize that the statistically significant result for the SD1 of TI seen here is likely either a result of measurement noise or a consequence of a type 1 error. The overall pattern of results suggested that Vglut2-based glutamatergic signaling is not required for regulating either the steady state of or the dynamic patterns of baseline breathing and hypercapnic ventilatory reflex under 5% CO2.

Vglut2 conditional knockout in central noradrenergic neurons fails to alter baseline breathing and the hypercapnic ventilatory reflex (5% CO2).

(A) Mouse model schematic and experimental protocol including habituation and hypercapnia protocol (5% CO2). (B) Under both room air and hypercapnia (5% CO2), Dbh-Cre; Vglut2 cKO mice didn’t show significant changes in respiratory output (VE/VO2), minute ventilation (VE), respiratory rate (Vf), tidal volume (VT), metabolism demand (VO2), inspiratory duration (TI), expiratory duration (TE), breath cycle duration (TTOT), inspiratory flow rate (VT/TI), and expiratory flow rate (VT/TE). Linear mixed-effects regression model, ****p<0.0001, **p<0.01, *p<0.05, ns: p20.05.

To further investigate the role of Vglut2-based glutamatergic signaling in the hypercapnic response, we interrogated the system further with more severe hypercapnic exposures and measured the breathing of Dbh-Cre; Vglut2 cKO mice under 7% CO2 and 10% CO2. Both 7% and 10% CO2 stimuli produced a significant respiratory response in both mutant and control mice. Under the 7% CO2 condition, Dbh-Cre; Vglut2 cKO mice showed a significantly decreased tidal volume (VT) (p=0.023) and a significantly reduced inspiratory flow rate (VT/TI) (p=0.0016), but the overall respiratory output (VE/VO2), respiratory rate (Vf), minute ventilation (VE), metabolic demand (VO2), inspiratory duration (TI), expiratory duration (TE), breath cycle duration (TTOT), and expiratory flow rate (VT/TE) did not show a significant difference between the mutant mice and their sibling controls (Figure 5A-B and Table 3). Additionally, none of the dynamic patterns of any respiratory parameter showed a significant difference between the mutant and control groups (Figure 5 supplement 1A-H). This result suggests that Vglut2-based glutamatergic signaling in central NA neurons is not required for regulating the hypercapnic ventilatory reflex and breathing regularity under 7% CO2.

Vglut2 conditional knockout in central noradrenergic neurons fails to alter the majority of hypercapnic ventilation parameters and metabolism under 7% CO2.

(A) Mouse model schematic and experimental protocol including habituation and hypercapnia protocol (7% CO2). (B) Knocking out Vglut2 in the whole NA system failed to alter VE/VO2, Vf, VE, VO2, TI, TE, TTOT, VT/TE but only showed a significantly reduce in VT and VT/TI under 7% CO2 condition. Linear mixed-effects regression model, ****p<0.0001, **p<0.01, *p<0.05, ns: p20.05.

Under 10% CO2 challenge, none of the steady state breathing parameters including the overall respiratory outputs (VE/VO2), respiratory rate (Vf), tidal volume (VT), minute ventilation (VE), metabolic demand (VO2), inspiratory duration (TI), expiratory duration (TE), breath cycle duration (TTOT), inspiratory flow rate (VT/TI), and expiratory flow rate (VT/TE) were significantly different between Dbh-Cre; Vglut2 cKO mice and their littermate controls, suggesting that Vglut2-based glutamatergic signaling in central NA neurons is not required for the hypercapnic ventilatory reflex under 10% CO2 (Figure 6A-B and Table 4). For the dynamic patterns of breathing, the SD1 of breath cycle duration (TTOT) (p=0.028) and SD1 of respiratory rate (Vf) (p=0.038) showed a significant increase in mutant mice compared to the controls (Figure 6 supplement 1A-H). This suggests that removing Vglut2 from central NA neurons may increase the breathing irregularity under high CO2 challenges like 10% CO2.

Vglut2 conditional knockout in central noradrenergic neurons fails to alter the hypercapnic ventilatory reflex (10% CO2).

(A) Mouse model schematic and experimental protocol including habituation and hypercapnia protocol (10% CO2). (B) Knocking out Vglut2 in the whole NA system didn’t significantly alter VE/VO2, Vf, VT, VE, VO2, TI, TE, TTOT, VT/TI, and VT/TE under 10% CO2 condition. Linear mixed-effects regression model, ****p<0.0001, **p<0.01, *p<0.05, ns: p20.05.

Vglut2-based glutamatergic signaling in central NA neurons is not required for the hypoxic ventilatory reflex under 10% O2

To determine if Vglut2-based glutamatergic signaling in central NA neurons is necessary to regulate the hypoxic chemoreflex, we measured the ventilation response of Dbh-Cre; Vglut2 cKO mice under 10% O2. We analyzed the ten respiratory parameters in three 5-min epochs following 10% O2 challenge onset. Similar to the hypercapnic challenges, the 10% O2 challenge produced a significant breathing response in every 5-min epoch compared to room air, but no significant difference was found in the overall respiratory output (VE/VO2), respiratory rate (Vf), tidal volume (VT), minute ventilation (VE), metabolism demand (VO2), inspiratory duration (TI), expiratory duration (TE), breath cycle duration (TTOT), inspiratory flow rate (VT/TI), and expiratory flow rate (VT/TE) between Dbh-Cre; Vglut2 cKO mice and controls in any of the three 5-min 10% O2 epochs (Figure 7A-B and Table 5). These data suggest that Vglut2-based glutamatergic signaling in central NA neurons is not required for the hypoxic ventilatory reflex under 10% O2. For the dynamic patterns of breathing, the SD2 of inspiratory flow rate (VT/TI) (p=0.005) and the SD2 of minute ventilation (VE) (p=0.026) were significantly increased during the first epoch (first 5 minutes) under the 10% O2 challenge in mutant mice, suggesting that Vglut2-based glutamate may play a role in maintaining breathing regularity during the initial, or reflexive, response to hypoxia (Figure 7 supplement 1A-H).

Vglut2 conditional knockout in central noradrenergic neurons fails to alter the hypoxic ventilatory reflex (10% O2).

(A) Mouse model schematic and experimental protocol including habituation and hypoxia protocol (10% O2). (B) Knocking out Vglut2 in the whole NA system did not show significant breathing changes including VE/VO2, VE, Vf, VT, VO2, TI, TE, TTOT, VT/TI, and VT/TE in either of the three five-minute time periods under hypoxia (10% O2). Linear mixed-effects regression model, ****p<0.0001, **p<0.01, *p<0.05, ns: p20.05.

Discussion

Neuron co-transmission of two or more neurotransmitters across excitatory, inhibitory, and neuro-modulatory facets is becoming an increasingly appreciated phenomenon in the central nervous system. Functional interrogation of co-transmission is essential not only to understand the distinct and cooperative roles that multiple signaling molecules may play at the synapse but also to determine the most critical targets for potential therapeutics. To our knowledge, the unopposed paradigm in the field until now has been that vesicular glutamate transporter 2 (Vglut2)-based glutamate transmission from central noradrenergic neurons is important in respiratory homeostasis. Anatomically, it has long been appreciated that central NA neurons co-express Vglut2, the major glutamate marker among three glutamate transporters (DePuy et al., 2013; Souza et al., 2022a; Stornetta et al., 2002a, 2002b; Yang et al., 2021). Also, it has been well documented that NA Vglut2 positive fibers innervate known central respiratory centers or other autonomic centers, that, when perturbed, could potentially disrupt respiratory homeostasis (Supplemental Table 1). Functionally, multiple reports have suggested that Vglut2-based glutamate transmission plays a role in respiratory homeostasis (Abbott et al., 2014; Malheiros-Lima et al., 2022, 2020, 2018). It has also been suggested that, at least for C1 neurons, the apparent lack of a plasmalemmal monoamine transporter may attenuate or eliminate noradrenergic or adrenergic release from these fibers (Guyenet et al., 2013). However, as discussed below, much of the functional evidence supporting a role for Vglut2-based glutamate transmission from central noradrenergic neurons in breathing is circumstantial or of a non-physiological nature, and absent or attenuated release of adrenaline and noradrenaline has not been demonstrated. Nonetheless, the dominant perspective that NA-Vglut2 glutamate transmission plays a role in breathing has remained unchallenged. Our studies, in contrast to prior work, show that loss of Vglut2 in NA neurons does not appreciably change baseline or chemosensory breathing in unanesthetized and unrestrained mice. Additionally, our work uncovers a novel dynamic expression pattern for Vglut2 and an entirely undescribed co-expression domain for Vglut3 in central NA neurons.

Vglut2 has been shown to be co-expressed in subsets of central NA neurons, however, to our knowledge, the potential for the central NA neurons to express the other glutamate transporters Vglut1 or Vglut3 has not been reported at the anatomical level. To determine if central NA neurons express either Vglut1 or Vglut3 and to confirm the Vglut2 expression, we carried out cumulative intersectional fate mapping using Slc17a7Cre; Dbhp2a-Flpo; Rosa26RC::FLTG, Slc17a6Cre; Dbhp2a-Flpo; Rosa26RC::FLTG and Slc17a8Cre; Dbhp2a-Flpo; Rosa26RC::FLTG compound mouse lines. As this approach maps the cumulative history of expression, we also examined acute adult expression of Slc17a7 (Vglut1), Slc17a6 (Vglut2) and Slc17a8 (Vglut3) using fluorescent RNA in situ hybridization. Our in situ results for Slc17a6 (Vglut2) agreed with earlier published outcomes (DePuy et al., 2013; Stornetta et al., 2002a, 2002b) showing as much as 80% expression in posterior NA neurons. However, there was no appreciable signal in anterior groups (A7, LC, A5, sub CD/CV). This lack of expression was in stark contrast to the cumulative fate map for Vglut2, showing as much as 75% of the anterior central NA neurons were recombined by the Vglut2-Cre over the lifetime of the animal. To further interrogate the acute activity of the Vglut2-Cre in the Locus Coeruleus in the adult mouse, we injected a Cre responsive virus in LC of Slc17a6Cre mice and found sparse Vglut2-Cre positive NA neurons in the LC. The outcomes from the cumulative fate map and acute injections are in agreement with published results from Yang et al. (2021). Notably, our use of the Dbhp2a-Flpoavoided expression in the parabrachial nucleus (PBN), which was a confounding variable in the ThFlpo mouse. In addition, 28.5% of A5 neurons were shown to be Slc17a6 (Vglut2) positive in adult rats by RNA scope (Souza et al., 2022b). It is notable that both our lab and the work published by (Stornetta et al., 2002a, 2002b) failed to detect Slc17a6 (Vglut2) expression in anterior groups, but that apparently very low levels can be detected by more sensitive recombinase and RNA scope methods. Additional electrophysiological evidence suggests that this expression is functional (Yang et al., 2021). However, it is notable that we see nearly 80% of the LC and 85% of A5 recombined in the intersectional strategy representing lifetime expression, but our study and others find only 20% of the LC and 28.5% of A5 may be expressing Slc17a6 (Vglut2) at any given moment in adults. Thus, it suggests that Vglut2 co-expression in anterior NA groups is time dependent, though the underlying temporal dynamics of Vglut2 remain unknown. We hypothesize that this could be either a temporal restriction in expression as the animal matures or may reflect dynamic change driven by behavioral or physiological experiences, which could be tested with the use of inducible conditional strategies.

For Slc17a7 (Vglut1), neither fate map nor in situ hybridization revealed any co-expression within the central NA system. However, Slc17a8 (Vglut3) expression was found in the posterior part of NA neurons including C2/A2 and C1/A1. This study is the first to characterize this expression. The majority of this Vglut3 co-expression is restricted to the posterior part of the C2/A2 NA population. This NA region has not been heavily implicated in respiratory control nonetheless it readily lends itself to interrogation by intersectional genetic methods and in situ hybridization. Given its distal expression from anterior C1, it is less likely that Vglut3 would compensate for loss of Vglut2 in anterior C1, though this has not been ruled out. Further, while Vglut3 is a marker of glutamatergic neurons, its role in neurotransmission is not wholly defined as Vglut3 is typically found in soma and dendrites (Fremeau et al., 2004).

The cumulative fate map and in situ hybridization experiments were performed in both females and males and no obvious sex difference was observed in the expression pattern of Vglut1, Vglut2 or Vglut3. However, only three mice were characterized in each dataset, thus, larger sample sizes may capture more subtle outcomes for potential sex differences.

Based on the Vglut2 co-expression in NA neurons and their projections to multiple brain regions important in respiration, several studies suggest that glutamate in NA populations is important in breathing control. However, the evidence those studies provided is circumstantial, indirect, or limited due to experimental caveats or technical limitations. Abbott et al. (2014), using exactly the same mouse model as this study, showed that conditionally knocking out Vglut2 in the whole central NA system did not significantly change respiratory rate under room air, which is consistent with our findings here. Our results differ in that they reported removing Vglut2 from central NA neurons attenuated an increase in respiratory rate resulting from unilateral C1 optogenetic stimulation. However, while the optogenetic stimulation of C1 neurons did drive a Vglut2 dependent increase in respiratory rate, it is not clear if the resulting change in breathing reflects a native feature of the network that might be engaged in chemosensory responses or other aspects of breathing. It remains possible that strong unilateral optogenetic stimulation (up to 20Hz) of a small, isolated population drives a Vglut2 dependent, but ectopic, function that is not typically engaged in the normal operation of the breathing network. Such ectopic outcomes could come from several potential, non-exclusive mechanisms. First, unilateral stimulation could drive a network hysteresis or dysregulation similar to a focal injury that is not a typical biological feature (Ducros et al., 2003). This point is well made in Abbott et al. (2013) in their discussion of C1 optogenetic stimulation where they state, “resulting cardiorespiratory response pattern should not be considered strictly “physiological,” because a physiological response is never initiated by selective activation of a single cluster of CNS neurons and the various subsets of C1 neurons are presumably never recruited en bloc under any physiological condition”. Second, strong stimulation could result in overwhelming synaptic mechanisms (i.e., glial uptake) resulting in glutamate spillover and unintended cross talk to affect extra-synaptic neurons proximal to targeted NA fibers. Activity dependent spillover transmission has been documented elsewhere in the nervous system (Henneberger et al., 2020; Hülsmann et al., 2000). Third, the outcomes could be secondary to a different autonomic or behavioral function requiring NA-derived glutamatergic signaling involving metabolism or circulation (DePuy et al., 2013; Yang et al., 2021). Additionally, respiratory parameters except for respiratory rate (Vf), such as tidal volume (VT) and minute ventilation (VE) were not reported, making comprehensive comparison difficult.

More circumstantially, Malheiros-Lima et al. (2020) suggested C1 neurons potentially release glutamate at the pFRG site to regulate active expiration under hypoxia in anesthetized rats by providing three indirect lines of evidence: 1) Vglut2-expressing C1 neurons project to the pFRG region; 2) Increased abdominal expiratory nerve activity (AbdEMG) was blunted after blockade of ionotropic glutamatergic receptors at the pFRG site under anesthesia using cytotoxic KCN mediated hypoxia; and 3) Depletion of C1 neurons eliminated the increased AbdEMG elicited by hypoxia. However, 1) the glutamatergic signaling targeting the pFRG region is not necessarily from C1 NA neurons since glutamatergic neurons from other regions also project to the pFRG site (Yang and Feldman, 2018). Thus, it is possible that the glutamatergic signaling that is required for hypoxic response is derived from other glutamatergic neurons such as the preBötzinger complex and the RTN itself. Furthermore, 2) C1 neuron depletion knocks out all the signaling modalities in the C1 population including both noradrenaline and glutamate. It is not clear which signaling drives the hypoxic response. Finally, 3) KCN injection mimics a hypoxic challenge by activating peripheral chemoreceptors, but the physiological responses differ from those seen during environmental hypoxia exposure. Thus, the abdominal activity changes due to KCN injection may not be recaptured under physiological hypoxic challenge.

Similarly, Malheiros-Lima et al. (2022) and Malheiros-Lima et al. (2018) showed that Vglut2-expressing C1 neurons project to both the NA A5 region and the preBötzinger complex. The blockade of ionotropic glutamatergic receptors at the A5 region or preBötzinger complex reduced the increase in phrenic nerve activity and the breathing frequency caused by optogenetic stimulation of C1 cells in an anesthetized preparation. Together these results suggest that Vglut2-expressing C1 neurons communicate with A5 or preBötzinger complex neurons by releasing glutamate in turn to regulate phrenic nerve activity or breathing frequency. Again, these two studies showed that C1 neurons are important to regulate phrenic nerve activity or breathing frequency, however, it is not necessarily through a direct C1-A5 or C1-preBötzinger complex glutamatergic pathway, whereas an indirect multi-synaptic pathway originating with C1 noradrenaline and ending with glutamate transmission from an intermediate to A5 or preBötzinger complex cannot be ruled out. Importantly, in all four studies noted here, respiratory measurements were limited or proxies of breathing under anesthesia were used. The role of NA-Vglut2 signaling was not reported in the context of an unanesthetized, unrestrained animal under normoxic, hypoxic, or hypercapnic breathing in these previous studies.

Thus, the limitations and indirect lines of the evidence in these studies raised the question as to whether or not glutamatergic signaling in central NA neurons is required or necessary to modulate respiratory homeostasis under more physiologically relevant circumstances and if NA-derived glutamate is an important mechanism to understand the pathophysiology of NA neurons in respiratory diseases. The goal of our study was to test the requirement of NA-based glutamatergic signaling in breathing homeostasis in the unanesthetized and unrestrained animal under conditions that would engage the whole respiratory control network rather than a singular component that may drive hysteresis or artifactual outcomes. Our in vivo breathing data under room air, hypoxic, and multiple hypercapnic conditions failed to show the requirement of NA-derived Vglut2 in normal breathing and respiratory chemoreflexes directly. For the dynamic patterns of breathing, we measured the SD1 and SD2 of eight parameters under each gas condition. We found no consistent change across multiple hypercapnic challenges or hypoxia. We did find that under severe hypercapnic challenge (10% CO2), NA derived Vglut2 removal resulted in an increase in the SD1 of breath cycle duration and respiratory rate, which was not observed under 5% and 7% CO2 conditions. Under hypoxia (10% O2), Dbh-Cre; Vglut2 cKO mice only showed an increase in the SD2 of inspiratory flow rate and minute ventilation during the first 5-minute of 10% O2 exposure. These findings may suggest that Vglut2-based glutamatergic signaling in central NA neurons may play a role in modulating breathing regularity under high CO2 challenge and hypoxia. However, the phenotypes were of a small magnitude and not consistent across challenges, and not at all seen under homeostatic breathing.

To ensure a rigorous and robust conclusion, we included several extra precautions and additional measures in our experimental design and analysis. First, we use exactly the same mouse model, DBH-Cre; Vglut2 cKO, as Abbott et al. (2014) used, and confirmed, by a second method (fluorescent RNA in situ hybridization) that Vglut2 expression was indeed abrogated in the whole central NA system. Our results are in strong agreement with DePuy et al. (2013) that recombination is efficient. Notably, we assayed for expression at the cell body, rather than the fibers and see no singular cell body that is scored positive. Second, we measured the breathing of unanesthetized, unrestrained, and habituated mice under physiological challenges including normoxia, hypercapnia and hypoxia. Third, to visualize breathing function comprehensively, the metabolic parameter and multiple respiratory parameters (both the steady state and the dynamic patterns) were measured and reported in absolute values including oxygen consumption, respiratory rate, tidal volume, minute ventilation, overall respiratory output (minute ventilation normalized to oxygen consumption), inspiratory duration, expiratory duration, breath cycle duration, inspiratory flow rate, and expiratory flow rate. Fourth, our experimental design was overpowered (i.e., n=16-21 vs 5-13 needed for power based on power analysis and n=7-8 reported in other manuscripts). Each respiratory/metabolic parameter for each physiological condition including room air, hypercapnia, and hypoxia was derived by quantifying the entire respiratory trace and analyzed between mutant and sibling controls by using a linear mixed-effects regression model with animal type (experimental vs. control) as fixed effects, animal ID as a random effect, and sex as an experimental covariate (for which we did not find a significant effect) (Lusk et al., 2023).

Despite our experimental design and the extensive nature of our measurements, there remain several possibilities or considerations in the interpretation of our data. The first possibility is Vglut3 in posterior CA domain (first characterized in this study) compensates for the effect of the loss of Vglut2 in NA neurons in breathing. However, we argue that this is unlikely because 1) Our model is the same used in the other study where a Vglut2-dependent effect was seen with optogenetic stimulation. 2) Vglut3 is expressed in the posterior C2/A2 NA neurons which are anatomically distant from anterior C1 and the LC (the two most likely NA candidates with Vglut2 co-expression to impact breathing). Also, posterior C2/A2 hasn’t been heavily implicated in breathing regulation. Future work will need to further investigate the role of Vglut3 in posterior CA in respiratory control in order to test this possibility. The second possibility is that pre- or post-natal developmental compensation corrects for the loss of NA derived glutamate. Again, we argue that is unlikely, as 1) we are using the same adult model that provided the most direct evidence and first functionally demonstrated a potential role for C1-derived glutamate in breathing; 2) Abbott et al. (2014) highlight that C1 NA neurons showed no obvious abnormalities in number, morphology, and projection patterns after knocking out Vglut2; and 3) we have examined the requirement of NA-expressed Vglut2 in P7 neonate mice in the autoresuscitation reflex and saw no differences (data not shown). However, it is still possible that developmental compensation occurs before P7, which is something that could be better tested by using the Dbh-CreERT2 (Stubbusch et al., 2011) to more acutely remove Vglut2 expression (though this leaves a window of 2-3 weeks for compensation to occur). Third, as we only address the loss of NA-Vglut2 based signaling across the entire NA system, it remains a formal possibility that this ultimately results in the removal of Vglut2 dependent signaling from two counter balancing regions of the NA system and are therefore left with null results whereas removal from anterior C1 alone would show a requirement. To test this possibility, it would require an intersectional conditional knockout/loss of function approach, something that has not yet been shown to be efficient and effective and would be beyond the scope of these studies. Fourth, it is possible that another neurotransmitter/neuropeptide within NA neurons might be released in higher amounts in Dbh-Cre; Vglut2 cKO mice to compensate for the deficiency of glutamate in breathing. Neuropeptide Y (NPY) and galanin have been shown to co-exist in LC NA neurons and are involved in regulating energy metabolism, stress, or anxiety (Ruohonen et al., 2012; Tasan et al., 2010; Weinshenker and Holmes, 2016). In our studies, we account for potential metabolic changes that may underlie or be coordinate with changes in breathing. However, we did not test for potential changes in blood pressure or other autonomic functions as well as affective state that may, in some obscure way, compensate for changes in breathing. Further interrogation of NPY or galanin signaling across NA neurons and potential compensatory effects in breathing through changes in metabolism or other autonomic function under stress or anxiety may yield notable insights. Lastly, our results are loss of function in nature. We do not test sufficiency or gain of function and cannot fully rule out a role for NA-Vglut2 based glutamatergic signaling in the control of breathing.

From a translational perspective, our data question whether or not glutamatergic signaling in NA neurons is likely to be a key mechanism and therefore a therapeutic target for breathing disorders, such as Rett syndrome and SIDS. Mecp2-deficient mice (a mouse model of Rett Syndrome that phenocopies many human symptoms) showed both a deficiency of NA populations (reduced number of NA neurons in C2/A2 and C1/A1 group) and highly variable respiratory rhythm at around 4-5 weeks of age (Roux et al., 2007; Viemari et al., 2005). 80% of Rett Syndrome patients experience breathing issues, such as unstable breathing, episodes of hyperventilation, and breath holds, throughout their lifespan (Ramirez et al., 2020). SIDS decedents show NA abnormalities and are hypothesized to ultimately succumb from a failure in cardiorespiratory autoresuscitation (Chigr et al., 1989; Garcia et al., 2013; Kopp et al., 1993; Mansouri et al., 2001; Ozawa et al., 2003; Takashima and Becker, 1991). However, Dbh-Cre; Vglut2 cKO mice, have normal baseline and chemosensory respiratory parameters as adults and neonate mice show normal autoresuscitation indicating that perturbed NA based glutamatergic signaling may not be a key driver in these or other related respiratory pathophysiologies.

In conclusion, our studies show that Vglut2 has a dynamic and extensive expression profile across the central NA system. We were able to characterize a novel Vglut3-expressing NA population in the posterior C2/A2 nuclei. Despite prior studies providing indirect and circumstantial evidence that noradrenergic-based glutamatergic signaling may play a role in control of breathing, our conditional loss of function studies in adult unanesthetized and unrestrained mice failed to consistently and significantly change room air breathing, the hypercapnic ventilatory reflex, and the hypoxic ventilatory reflex. These outcomes offer a contrasting perspective from the current field view and provides further insight into the potential role of NA-glutamate transmission in the control of breathing and NA neuron dysfunction in respiratory disorders.

Materials and Methods

Ethical Approval

Studies were approved by the Baylor College of Medicine Institutional Animal Care and Use Committee (IACUC) under protocol AN-6171, and all experiments reported here were performed in accordance with relevant guidelines and regulations.

Breeding, Genetic Background, and Maintenance of Mice

We maintained all our heterozygous mouse strains by backcrossing to wildtype C57BL/6J mice and homozygous mouse strains by sibling crosses. For immunofluorescence experiments, heterozygous B6;129S-Slc17a7tm1.1(cre)Hze/J (Slc17a7Cre, Vglut1-Cre) (Jax Stock No: 023527), Slc17a6tm2(cre)Lowl/J (Slc17a6Cre, Vglut2-Cre) (Jax Stock No: 016963) and B6;129S-Slc17a8tm1.1(cre)Hze/J (Slc17a8Cre, Vglut3-Cre) (Jax Stock No: 018147) were mated with heterozygous Dbhem2.1(flpo)Rray (Dbhp2a-Flpo, DBH-p2a-Flpo) (MMRRC ID: 41575) (Sun and Ray, 2016) mice respectively to derive compound lines with both Cre and Flpo alleles (Slc17a7Cre; Dbhp2a-Flpo, Slc17a6Cre; Dbhp2a-Flpo, and Slc17a8Cre; Dbhp2a-Flpo). Then these three compound lines were each mated with homozygous B6.Cg-Gt(ROSA)26Sortm1.3(CAG-tdTomato,-EGFP)Pjen/J (Rosa26RC::FLTG) (Jax Stock No: 026932) to derive three different intersectional mouse lines Slc17a7Cre; Dbhp2a-Flpo; Rosa26RC::FLTG, Slc17a6Cre; Dbhp2a-Flpo; Rosa26RC::FLTG and Slc17a8Cre; Dbhp2a-Flpo; Rosa26RC::FLTG. For in situ hybridization experiments, wildtype C57BL/6J mice were ordered from the Center of Comparative Medicine (CCM), Baylor College of Medicine. For NA-Vglut2 conditional loss of function in situ hybridization and plethysmography experiments, hemizygous transgene Tg(Dbh-cre)KH212Gsat (Dbh-Cre) (MMRRC ID: 036778-UCD GENSAT) mice were mated with homozygous Slc17a6tm1Lowl/J (Slc17a6flox/flox) (Jax Stock No: 012898) to derive Dbh-Cre; Slc17a6flox/+. Dbh-Cre; Slc17a6flox/+mice were mated with Slc17a6flox/flox to derive Dbh-Cre; Slc17a6flox/flox(Dbh-Cre; Vglut2 cKO). Sibling mice that lacked the Cre allele or carried the Cre allele but lacked the floxed Vglut2 alleles were used as controls. Rosa26 specific primers for the Rosa26RC::FLTG mice were 5’-GCACTTGCTCTCCCAAAGTC, 5’-GGGCGTACTTGGCATATGAT, and 5’-CTTTAAGCCTGCCCAGAAGA (Ray et al., 2011) and yield a 495 bp band (targeted) and 330 bp band (wt). Cre-specific primers for all Cre drivers were 5’-ATCGCCATCTTCCAGCAGGCGCACCATTGCCC and 5’-GCATTTCTGGGGATTGCTTA and yielded a 550 bp band if positive. Flpo-specific primers for Dbhp2a- Flpo are 5’-CACGCCCAGGTACTTGTTCT and 5’-CCACAGCAAGAAGATGCTGA (Sun and Ray, 2016) and yielded a 226 bp band if positive. Slc17a6flox specific primers for Vglut2-floxed mice are available at The Jackson Laboratory website (https://www.jax.org/strain/012898).

Immunofluorescence staining

Slc17a7Cre; Dbhp2a-Flpo; Rosa26RC::FLTG, Slc17a6Cre; Dbhp2a-Flpo; Rosa26RC::FLTGand Slc17a8Cre; Dbhp2a- Flpo; Rosa26RC::FLTG adult mice of both sexes were sacrificed and transcardially perfused with 0.1 M phosphate-buffered saline (PBS) then with 4% paraformaldehyde (PFA) in PBS. Mouse brains were dissected out and fixed for 2h in 4% PFA before a PBS rinse and dehydration in 30% sucrose in PBS. Brains were embedded in OCT blocks, sectioned at the thickness of 30μm, mounted on slides, and stored at -80°C until they were ready for staining. The slides were hydrated in 0.1% Triton-X in PBS (PBST) for 15 mins, blocked with 5% donkey serum in 0.1% PBST for 1h at room temperature and then incubated with primary antibodies for 72 hours at 4°C in 0.1% PBST with 5% donkey serum. Tissues were washed in 0.1% PBST 3 times for 10 min each and then incubated with secondary antibodies for 2hs at room temperature in 0.1% PBST with 5% donkey serum. Slides were washed with 0.1% PBST for 10 min and washed in PBS twice for 10 min each, stained for DAPI, washed 3 times for 10 min each with PBS and mounted in ProLong Glass (Invitrogen). The following primary and secondary antibodies were used: chicken anti-GFP (1:1,000, Abcam ab13970), rabbit anti-dsRed (1:1,000, Clontech 632496), donkey anti-chicken Cy2 (1:500, Jackson 703-225-155), donkey anti-rabbit Cy3 (1:500, Jackson 711-165-152).

In situ hybridization

Mouse brains were dissected out from adult mice of both sexes with 6-8 week of age, sectioned into 25µm brain sections and mounted on slides. We generated a digoxigenin (DIG)-labeled mRNA antisense probe against Slc17a7 (Vglut1), Slc17a6 (Vglut2), and Slc17a8 (Vglut3) and fluorescein (FITC)-labeled mRNA against Dbh using reverse-transcribed mouse cDNA as a template and an RNA DIG or FITC-labeling kits from Roche (Sigma). Primer and probe sequences for the Slc17a7 (Vglut1), Slc17a6 (Vglut2), and Slc17a8 (Vglut3) and Dbh probes are available in the Allen Brain Atlas (http://www.brain-map.org). For the Slc17a6 (Vglut2) and Dbh double ISH in Dbh-Cre; Vglut2 cKO and their littermate controls, we generated a new Slc17a6 (Vglut2) probe targeting exon 2 of Slc17a6 specifically (Tong et al., 2007) and the probe sequence is 892-1144bp as Slc17a6 transcript variant 1 and the size is 253bp. ISH was performed by the RNA In Situ Hybridization Core at Baylor College of Medicine using an automated robotic platform as previously described (Yaylaoglu et al., 2005) with modifications of the protocol for double ISH. Modifications in brief [see Yaylaoglu et al. (2005) for buffer descriptions]: both probes were hybridized to the tissue simultaneously. After the described washes and blocking steps the DIG-labeled probes were visualized using tyramide-Cy3 Plus (1/75 dilution, 15-minute incubation, Akoya Biosciences). After washes in TNT, the remaining HRP-activity was quenched by a 10-minute incubation in 0.2M HCl. The sections were then washed in TNT, blocked in TNB for 15 minutes, and incubated at room temperature for 30 minutes with HRP-labeled sheep anti-FITC antibody (1/500 in TNB, Roche/Sigma). Following washes in TNT, the FITC-labeled probe was visualized using tyramide-FITC Plus (1/50 dilution, 15-minute incubation, Akoya Biosciences). Following washes in TNT, the slides were stained with DAPI (invitrogen), washed again, removed from the machine, and mounted in ProLong Diamond (Invitrogen).

Viral Injection

To verify Vglut2 co-expression in LC in adult mice (Yang et al., 2021), 6-week-old Slc17a7Cre mice were injected with pAAV-EF1a-DIO-tdTomato-WPRE virus (RRID:Addgene_133786, obtained from Joshua Ortiz at the Optogenetics and Viral Vectors Core at the Jan and Dan Duncan Neurological Research Institute, 1ul at 7.90E+10 Gc/mL) into the LC (coordinates from bregma anteroposterior -5.4 mm, lateral + 0.8 mm and dorsoventral – 4.0 mm) and allowed to incubate for 4 weeks.

Plethysmography

Plethysmography on unanesthetized and free-moving mice was carried out as described in (Ray et al., 2011). 6–8-week-old adult mice of both sexes with minimum group sizes n=16 were used in both the experimental and control group for each experiment. Animals were subjected to a five-day habituation protocol with each day including several minutes of handling, temperature taken by rectal probe and at least 30 min exposure in the plethysmography chamber (Martinez et al., 2019). Plethysmography started to be performed on the sixth day and was finished for all animals within a week of the last day of habituation (Maximum 6 mice can be assayed for plethysmography per day due to the limited number of plethysmography rigs). On the day of testing, mice were taken out from their home cage, weighed, and rectal temperature was taken. Animals were then placed into a flow-through, temperature-controlled (about 32°C with real-time temperature recording) plethysmography chamber and allowed to acclimate for 20-40 mins in room air (21% O2/79% N2) conditions. After acclimation (indicated by a steady respiratory trace free from movement artifact), a 20-min baseline breathing trace was taken under room air, then the chamber gas was switched to a hypercapnic or hypoxic mixture of 5% CO2/21% O2/74% N2, 7% CO2/21% O2/72% N2, 10% CO2/21% O2/69% N2, or 10% O2/90% N2, depending on the protocol, for 20 min. Chamber gas was then switched back to room air for another 20 min. The animals were removed from the chamber and rectal temperature was measured immediately after the mice were taken out from the chamber. Each testing period for an individual mouse was separated by 24hs to allow for a full recovery.

Plethysmography data analysis and statistics

Details were previously described in (Martinez et al., 2019; Ray et al., 2011). Plethysmography pressure changes were measured using a Validyne DP45 differential pressure transducer, CD15 carrier demodulator and a reference chamber, and were recorded with LabChart Pro in real time. Respiratory waveforms were analyzed by the SASSI module of the Breathe Easy software to determine respiratory frequency (Vf), tidal volume (VT), minute ventilation (VE), oxygen consumption (VO2), ventilatory equivalents for oxygen (VE/VO2), inspiratory duration (TI), expiratory duration (TE), breath cycle duration (TTOT), inspiratory flow rate (VT/TI), and expiratory flow rate (VT/TE) (Lusk et al., 2023). Poincaré plots for TI, TE, TTOT, VT/TI, VT/TE, Vf, VT, VE and their SD1 and SD2 properties were generated and characterized by the STAGG module of the Breathe Easy software. Analysis windows of the respiratory parameters were as follows: for room air, the entire experimental period was considered; for hypercapnia, only the last 5 min challenge for each of 5/7/10% CO2 conditions were analyzed; and for hypoxia, three separate 5-min time periods, namely the 5th-10th min, 10th-15th min, and 15th-20th min intervals, were used due to the stereotypical biphasic respiratory response under this condition. Only steady quiescent breathing periods were included in the data analysis. A power analysis was performed using the reported effect size in Abbott et al. (2014), which used the same mouse model as we used here; 5-13 mice were necessary to observe a statistically significant result (Abbott et al. (2014) was able to see a significant difference between two groups with n=7 mice). In our experiments, the sample size for each group (Dbh-Cre; Vglut2 cKO and control) exceeded 13. Steady state results (Vf, VE, VT, VO2, VE/VO2, TI, TE, TTOT, VT/TI, VT/TE) for room air and hypercapnic or hypoxic data were compared between Dbh-Cre; Vglut2 cKO cohorts and sibling controls using a linear mixed-effects regression model with animal type (experimental vs. control) as fixed effects and animal ID as a random effect (Lusk et al., 2023). SD1 and SD2 results for room air and hypercapnic or hypoxic data were compared between Dbh-Cre; Vglut2 cKO cohorts and sibling controls using a Mann-Whitney U test (Ferreira et al., 2022). A p-value threshold of p < 0.05 was used to test for statistical significance. Individual data points, means, and standard errors of the mean are shown on all charts. The graphs were plotted by Prism 8.

Image quantification

Images were taken by using a Zeiss upright epifluorescent microscope and a Zeiss LSM 880 with Airyscan FAST confocal microscope. Images were captured using Zen software with z-stack function from top to bottom with 0.34μm intervals, exported, and then analyzed in Imaris using the spots and surface functions. For quantification of immunofluorescence staining, each GFP positive area coincident with DAPI (to denote nuclear localization) was defined as Vglut2 or Vglut3 expressing NA neurons while each tdTomato positive area coincident with DAPI was defined as NA neurons without any Vglut2 or Vglut3 co-expression. For quantification of Slc17a6 (Vglut2) or Slc17a8 (Vglut3) with Dbh double in situ hybridization in adult WT mice, Dbh positive areas coinciding with DAPI were identified as NA neurons and the Dbh positive areas overlapped with Slc17a6 (Vglut2) or Slc17a8 (Vglut3) positive pixels and DAPI were defined as NA neurons colocalized with Slc17a6 (Vglut2) or Slc17a8 (Vglut3). The number of Slc17a6 (Vglut2) or Slc17a8 (Vglut3) positive NA neurons and Slc17a6 (Vglut2) or Slc17a8 (Vglut3) negative NA neurons was counted in each image for both immunofluorescence and in situ experiments and the percentage of Slc17a6 (Vglut2) or Slc17a8 (Vglut3) positive NA neurons among all NA neurons in each NA nucleus in each mouse brainstem was calculated every other brain section unilaterally. For quantification of Slc17a6 (Vglut2) and Dbh double ISH in Dbh-Cre; Vglut2 cKO and their littermate controls, Slc17a6 (Vglut2) pixel intensities in Dbh positive areas coincident with DAPI (NA neurons) were measured in mutant and control images separately. The Slc17a6 (Vglut2) pixel intensity in NA neurons of control brains was normalized as 1 and the relative Slc17a6 (Vglut2) pixel intensity of NA neurons in mutant brains compared to that in control brains was calculated. At least three mouse brains were examined for each set of experiments in each group. All quantitative results were graphed using Prism 8.

eLife Assessment

Chang et al. provide glutamate co-expression profiles in the central noradrenergic system and test the requirement of Vglut2-based glutamatergic release in respiratory and metabolic activity under physiologically relevant gas challenges. Their experiments provide compelling evidence that conditional deletion of vesicular glutamate transporters from noradrenergic neurons does not impact steady-state breathing or metabolic activity in room air, hypercapnia, or hypoxia. This study provides an important contribution to our understanding of how noradrenergic neurons regulate respiratory homeostasis in conscious adult mice.

Reviewer #1 (Public Review):

Summary:

Chang et al. provide glutamate co-expression profiles in the central noradrenergic system and test the requirement of Vglut2-based glutamatergic release in respiratory and metabolic activity under physiologically relevant gas challenges. Their experiments show that conditional deletion of Vglut2 in NA neurons does not impact steady-state breathing or metabolic activity in room air, hypercapnia, or hypoxia. Their observations challenge the importance of glutamatergic signaling from Vglut2 expressing NA neurons in normal respiratory homeostasis in conscious adult mice.

Strengths:

The comprehensive Vglut1, Vglut2, and Vglut3 co-expression profiles in the central noradrenergic system and the combined measurements of breathing and oxygen consumption are two major strengths of this study. Observations from these experiments provide previously undescribed insights into (1) expression patterns for subtypes of the vesicular glutamate transporter protein in the noradrenergic system and (2) the dispensable nature of Vglut2-dependent glutamate signaling from noradrenergic neurons to breathing responses to physiologically relevant gas challenges in adult conscious mice.

Weaknesses:

Although the cellular expression profiles for the vesicular glutamate transporters are provided, the study does not document that glutamatergic-based signaling originating from noradrenergic neurons is evident at the cellular level under normal, hypoxic, and/or hypercapnic conditions. The authors effectively recognize this issue and appropriately discuss their findings in this context.

Reviewer #2 (Public Review):

The authors characterized the recombinase-based cumulative fate maps for vesicular glutamate transporters (Vglut1, Vglut2 and Vglut3) expression and compared those maps to their real-time expression profiles in central NA neurons by RNA in situ hybridization in adult mice. Authors have revealed a new and intriguing expression pattern for Vglut2, along with an entirely uncharted co-expression domain for Vglut3 within central noradrenergic neurons. Interestingly, and in contrast to previous studies, the authors demonstrated that glutamatergic signaling in central noradrenergic neurons does not exert any influence on breathing and metabolic control either under normoxic/normocapnic conditions or after chemoreflex stimulation. Also, they showed for the first-time the Vglut3-expressing NA population in C2/A2 nuclei. In addition, they were also able to demonstrate Vglut2 expression in anterior NA populations, such as LC neurons, by using more refined techniques, unlike previous studies.

A major strength of the study is the use of a set of techniques to investigate the participation of NA-based glutamatergic signaling in breathing and metabolic control. The authors provided a full characterization of the recombinase-based cumulative fate maps for Vglut transporters. They performed real-time mRNA expression of Vglut transporters in central NA neurons of adult mice. Further, they evaluated the effect of knocking down Vglut2 expression in NA neurons using a DBH-Cre; Vglut2cKO mice on breathing and control in unanesthetized mice. Finally, they injected the AAV virus containing Cre-dependent Td tomato into LC of v-Glut2 Cre mice to verify the VGlut2 expression in LC-NA neurons. A very positive aspect of the article is that the authors combined ventilation with metabolic measurements. This integration holds particular significance, especially when delving into the exploration of respiratory chemosensitivity. Furthermore, the sample size of the experiments is excellent.

Despite the clear strengths of the paper, some weaknesses exist. It is not clear in the manuscript if the experiments were performed in males and females and if the data were combined. I believe that the study would have benefited from a more comprehensive analysis exploring the sex specific differences. The reason I think this is particularly relevant is the developmental disorders mentioned by the authors, such as SIDS and Rett syndrome, which could potentially arise from disruptions in central noradrenergic (NA) function, exhibit varying degrees of sex predominance. Moreover, some of the noradrenergic cell groups are sexually dimorphic. For instance, female Wistar rats exhibit a larger LC size and more LC-NA neurons than male subjects (Pinos et al., 2001; Garcia-Falgueras et al., 2005). More recently, a detailed transcriptional profiling investigation has unveiled the identities of over 3,000 genes in the LC. This revelation has highlighted significant sexual dimorphisms, with more than 100 genes exhibiting differential expression within LC-NA neurons at the transcript level. Furthermore, this investigation has convincingly showcased that these distinct gene expression patterns have the capacity to elicit disparate behavioral responses between sexes (Mulvey et al., 2018). Therefore, the authors should compare the fate maps, Vglut transporters in males and females, at least considering LC-NA neurons. Even in the absence of identified sex differences, this information retains significant importance.

An important point well raised by the authors is that although suggestive, these experiments do not definitively rule out that NA-Vglut2 based glutamatergic signaling has a role in breathing control. Subsequent experiments will be necessary to validate this hypothesis.

An improvement could be made in terms of measuring body temperature. Opting for implanted sensors over rectal probes would circumvent the need to open the chamber, thereby preventing alterations in gas composition during respiratory measurements. Further, what happens to body temperature phenotype in these animals under different gas exposures? These data should be included in the Tables.

Is it plausible that another neurotransmitter within NA neurons might be released in higher amounts in DBH-Cre; Vglut2 cKO mice to compensate for the deficiency in glutamate and prevent changes in ventilation?

Continuing along the same line of inquiry is there a possibility that Vglut2 cKO from NA neurons not only eliminates glutamate release but also reduces NA release? A similar mechanism was previously found in VGLUT2 cKO from DA neurons in previous studies (Alsio et al., 2011; Fortin et al., 2012; Hnasko et al., 2010). Additionally, does glutamate play a role in the vesicular loading of NA? Therefore, could the lack of effect on breathing be explained by the lack of noradrenaline and not glutamate?

Reviewer #4 (Public Review):

Summary:

Although previous research suggested that noradrenergic glutamatergic signaling could influence respiratory control, the work performed by Chang and colleagues reveals that excitatory (specifically Vglut2) neurons is dynamically and widely expressed throughout the central noradrenergic system, but it is not significantly crucial to change baseline breathing as well the hypercapnia and hypoxia ventilatory responses. The central point that will make a significant change in the field is how NA-glutamate transmission may influence breathing control and the dysfunction of NA neurons in respiratory disorders.

Strengths:

There are several strengths such as the comprehensive analysis of Vglut1, Vglut2, and Vglut3 expression in the central noradrenergic system and the combined measurements of breathing parameters in conscious unrestrained mice.

Other considerations :

These results strongly suggest that glutamate may not be necessary for modulating breathing under normal conditions or even when faced with high levels of carbon dioxide (hypercapnia) or low oxygen levels (hypoxia). This finding is unexpected, considering many studies have underscored glutamate's vital role in respiratory regulation, more so than catecholamines. This leads us to question the significance of catecholamines in controlling respiration. Moreover, if glutamate is not essential for this function, we need to explore its role in other physiological processes such as sympathetic nerve activity (SNA), thermoregulation, and sensory physiology.

Author response:

The following is the authors’ response to the previous reviews.

All of the reviewers indicate that their major concerns have been adequately addressed, but they each have a few comments that the authors should consider before submitting a final version (without further review) for publication. For example, a statement about the sex of the mice used in the studies and whether any differences were noted if both sexes were used. The idea that the loss of glutamate transport might affect NA loading into vesicles is also worth considering. Finally, the authors might want to mention that the role of neuropeptide release from NA neurons needs further examination.

As noted in the prior submitted revision, all experiments contained both males and females and this was addressed in our re-submission. In our analysis of breathing and metabolism, sex was included in the analysis and no significant phenotypic difference was observed (The statement of no sex difference is in line 451-456). For the fate map and in situ experiments, although the group size is small, we did not see obvious differences in the expression patterns in the three glutamate transporters between females and males (line 347-350). All the anatomical and phenotypic data in this manuscript are presented as combined graphs (figure 1, figure 1 supplement 1, figure 2, figure 2 supplement 2, figure 4,5,6,7) and we had differentially labeled our data points by sex (female data is pink and male data is blue).

The possibility that loss of Vglut2 might affect NA release has been added in the discussion (line 485-491) of the current revision. Dopamine Beta Hydroxylase (DBH) converts dopamine to noradrenaline in the vesicles, thus, glutamate may not directly affect noradrenaline loading into vesicles. However, since loss of Vglut2 reduced dopamine release in subsets of dopaminergic neurons, it remains possible that glutamate affects dopamine loading in NA neurons and in turn perturbs DA to NA conversion in the vesicle by DBH and subsequent noradrenaline release. Future work could examine this hypothesis using fast-scan cyclic voltammetry (FSCV) or microdialysis.

The further examination of the role of neuropeptide release from NA neurons is mentioned in the discussion (line 491-494 and line 497-499 of the pre).

eLife assessment

Chang et al. provide glutamate co-expression profiles in the central noradrenergic system and test the requirement of Vglut2-based glutamatergic release in respiratory and metabolic activity under physiologically relevant gas challenges. Their experiments provide compelling evidence that conditional deletion of vesicular glutamate transporters from noradrenergic neurons does not impact steady-state breathing or metabolic activity in room air, hypercapnia, or hypoxia. This study provides an important contribution to our understanding of how noradrenergic neurons regulate respiratory homeostasis in conscious adult mice.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

Chang et al. provide glutamate co-expression profiles in the central noradrenergic system and test the requirement of Vglut2-based glutamatergic release in respiratory and metabolic activity under physiologically relevant gas challenges. Their experiments show that conditional deletion of Vglut2 in NA neurons does not impact steady-state breathing or metabolic activity in room air, hypercapnia, or hypoxia. Their observations challenge the importance of glutamatergic signaling from Vglut2 expressing NA neurons in normal respiratory homeostasis in conscious adult mice.

Strengths:

The comprehensive Vglut1, Vglut2, and Vglut3 co-expression profiles in the central noradrenergic system and the combined measurements of breathing and oxygen consumption are two major strengths of this study. Observations from these experiments provide previously undescribed insights into (1) expression patterns for subtypes of the vesicular glutamate transporter protein in the noradrenergic system and (2) the dispensable nature of Vglut2dependent glutamate signaling from noradrenergic neurons to breathing responses to physiologically relevant gas challenges in adult conscious mice.

Weaknesses:

Although the cellular expression profiles for the vesicular glutamate transporters are provided, the study does not document that glutamatergic-based signaling originating from noradrenergic neurons is evident at the cellular level under normal, hypoxic, and/or hypercapnic conditions. The authors effectively recognize this issue and appropriately discuss their findings in this context.

We thank the reviewer for the positive evaluation of our work.

Reviewer #2 (Public Review):

The authors characterized the recombinase-based cumulative fate maps for vesicular glutamate transporters (Vglut1, Vglut2 and Vglut3) expression and compared those maps to their realtime expression profiles in central NA neurons by RNA in situ hybridization in adult mice. Authors have revealed a new and intriguing expression pattern for Vglut2, along with an entirely uncharted co-expression domain for Vglut3 within central noradrenergic neurons. Interestingly, and in contrast to previous studies, the authors demonstrated that glutamatergic signaling in central noradrenergic neurons does not exert any influence on breathing and metabolic control either under normoxic/normocapnic conditions or after chemoreflex stimulation. Also, they showed for the first-time the Vglut3-expressing NA population in C2/A2 nuclei. In addition, they were also able to demonstrate Vglut2 expression in anterior NA populations, such as LC neurons, by using more refined techniques, unlike previous studies.

A major strength of the study is the use of a set of techniques to investigate the participation of NA-based glutamatergic signaling in breathing and metabolic control. The authors provided a full characterization of the recombinase-based cumulative fate maps for Vglut transporters. They performed real-time mRNA expression of Vglut transporters in central NA neurons of adult mice. Further, they evaluated the effect of knocking down Vglut2 expression in NA neurons using a DBH-Cre; Vglut2cKO mice on breathing and control in unanesthetized mice. Finally, they injected the AAV virus containing Cre-dependent Td tomato into LC of v-Glut2 Cre mice to verify the VGlut2 expression in LC-NA neurons. A very positive aspect of the article is that the authors combined ventilation with metabolic measurements. This integration holds

particular significance, especially when delving into the exploration of respiratory chemosensitivity. Furthermore, the sample size of the experiments is excellent. Despite the clear strengths of the paper, some weaknesses exist. It is not clear in the manuscript if the experiments were performed in males and females and if the data were combined. I believe that the study would have benefited from a more comprehensive analysis exploring the sex specific differences. The reason I think this is particularly relevant is the developmental disorders mentioned by the authors, such as SIDS and Rett syndrome, which could potentially arise from disruptions in central noradrenergic (NA) function, exhibit varying degrees of sex predominance. Moreover, some of the noradrenergic cell groups are sexually dimorphic. For instance, female Wistar rats exhibit a larger LC size and more LC-NA neurons than male subjects (Pinos et al., 2001; Garcia-Falgueras et al., 2005). More recently, a detailed transcriptional profiling investigation has unveiled the identities of over 3,000 genes in the LC. This revelation has highlighted significant sexual dimorphisms, with more than 100 genes exhibiting differential expression within LC-NA neurons at the transcript level. Furthermore, this investigation has convincingly showcased that these distinct gene expression patterns have the capacity to elicit disparate behavioral responses between sexes (Mulvey et al., 2018).

Therefore, the authors should compare the fate maps, Vglut transporters in males and females, at least considering LC-NA neurons. Even in the absence of identified sex differences, this information retains significant importance.

An important point well raised by the authors is that although suggestive, these experiments do not definitively rule out that NA-Vglut2 based glutamatergic signaling has a role in breathing control. Subsequent experiments will be necessary to validate this hypothesis.

An improvement could be made in terms of measuring body temperature. Opting for implanted sensors over rectal probes would circumvent the need to open the chamber, thereby preventing alterations in gas composition during respiratory measurements. Further, what happens to body temperature phenotype in these animals under different gas exposures? These data should be included in the Tables.

Is it plausible that another neurotransmitter within NA neurons might be released in higher amounts in DBH-Cre; Vglut2 cKO mice to compensate for the deficiency in glutamate and prevent changes in ventilation?

Continuing along the same line of inquiry is there a possibility that Vglut2 cKO from NA neurons not only eliminates glutamate release but also reduces NA release? A similar mechanism was previously found in VGLUT2 cKO from DA neurons in previous studies (Alsio et al., 2011; Fortin et al., 2012; Hnasko et al., 2010). Additionally, does glutamate play a role in the vesicular loading of NA? Therefore, could the lack of effect on breathing be explained by the lack of noradrenaline and not glutamate?

We thank the reviewer for the positive evaluation and further suggestions. Please see our response in “Author Response” to the previous version of Reviewer #2 (Public review).

Reviewer #4 (Public Review):

Summary:

Although previous research suggested that noradrenergic glutamatergic signaling could influence respiratory control, the work performed by Chang and colleagues reveals that excitatory (specifically Vglut2) neurons is dynamically and widely expressed throughout the central noradrenergic system, but it is not significantly crucial to change baseline breathing as well the hypercapnia and hypoxia ventilatory responses. The central point that will make a significant change in the field is how NA-glutamate transmission may influence breathing control and the dysfunction of NA neurons in respiratory disorders.

Strengths:

There are several strengths such as the comprehensive analysis of Vglut1, Vglut2, and Vglut3 expression in the central noradrenergic system and the combined measurements of breathing parameters in conscious unrestrained mice.

Other considerations :

These results strongly suggest that glutamate may not be necessary for modulating breathing under normal conditions or even when faced with high levels of carbon dioxide (hypercapnia) or low oxygen levels (hypoxia). This finding is unexpected, considering many studies have underscored glutamate's vital role in respiratory regulation, more so than catecholamines. This leads us to question the significance of catecholamines in controlling respiration. Moreover, if glutamate is not essential for this function, we need to explore its role in other physiological processes such as sympathetic nerve activity (SNA), thermoregulation, and sensory physiology.

We thank the reviewer for the positive evaluation and further suggestions. The potential role of noradrenergic-derived glutamate in other processes, which is beyond the scope of this study, should be addressed in the future.

Recommendations for the authors:

Reviewer #2 (Recommendations For The Authors):

All of my concerns were effectively resolved, leading me to accept the paper. However, I suggest that the authors consider investing in a more reliable system for measuring body temperature, as accurate measurements of this parameter are crucial for whole body plethysmography.

Thank you for the suggestion. The real-time measurement of body temperature is a goal in future studies.

Reviewer #4 (Recommendations For The Authors):

Because I am revising a revised version, I believe the authors have addressed most, if not all, the concerns raised by already 3 reviewers. In my understanding the authors achieved their aims and the results are totally supported by the conclusions. The impact of this work on the respiratory field is significant and is likely to advance the field. The methods and data utilized, which combine standard techniques with genetic tools, will be highly beneficial to the research community.

In my understanding I still have one concern that if glutamate is not critical, then what is? Could we potentially disable the noradrenergic (NA) system while preserving glutamate functionality to determine if the NA system is indeed crucial for respiratory physiology? This approach might provide clearer insights into the mechanisms underlying respiratory control.

We agree that there remain several exciting questions about the respective roles of noradrenaline, glutamate, and other neuropeptides such as Neuropeptide Y (NPY) and galanin. We are currently devising strategies to address the respective and combinatorial roles for all these candidates in breathing control. Most simply, we can conditionally, mutagenized each of them in the central noradrenergic system in an acute manner using DBH-CreER mice to determine if any of them are critical to respiratory control with the advantage of minimizing developmental compensatory events.

Version 3

Abstract

Abstract

Central noradrenergic (NA) neurons are key constituents of the respiratory homeostatic network. NA dysfunction is implicated in several developmental respiratory disorders including Congenital Central Hyperventilation Syndrome (CCHS), Sudden Infant Death Syndrome (SIDS) and Rett Syndrome. The current unchallenged paradigm in the field, supported by multiple studies, is that glutamate co-transmission in subsets of central NA neurons plays a role in breathing control. If true, NA-glutamate co-transmission may also be mechanistically important in respiratory disorders. However, the requirement of NA-derived glutamate in breathing has not been directly tested and the extent of glutamate co-transmission in the central NA system remains uncharacterized. Therefore, we fully characterized the cumulative fate maps and acute adult expression patterns of all three Vesicular Glutamate Transporters (Slc17a7 (Vglut1), Slc17a6 (Vglut2), and Slc17a8 (Vglut3)) in NA neurons, identifying a novel, dynamic expression pattern for Vglut2 and an undescribed co-expression domain for Vglut3 in the NA system. In contrast to our initial hypothesis that NA derived glutamate is required to breathing, our functional studies showed that loss of Vglut2 throughout the NA system failed to alter breathing or metabolism under room air, hypercapnia, or hypoxia in unrestrained and unanesthetized mice. These data demonstrate that Vglut2-based glutamatergic signaling within the central NA system is not required for normal baseline breathing and hypercapnic, hypoxic chemosensory reflexes. These outcomes challenge the current understanding of central NA neurons in the control of breathing and suggests that glutamate may not be a critical target to understand NA neuron dysfunction in respiratory diseases.

Introduction

Breathing is a vital and life-sustaining function supporting homeostatic processes, most critically maintaining blood pH/CO2 and O2 levels within a narrow physiological range. Respiratory homeostasis is mediated by neuron-modulated lung ventilation adjustments in response to physiological deviations resulting in high pCO2 or low pO2 blood and tissue levels, known as the hypercapnic and hypoxic reflexes respectively (Del Negro et al., 2018; Dick et al., 2018). These reflexes are part of a complex brainstem neural network that integrates a multitude of information streams across the central and peripheral nervous systems to regulate respiratory output. Within this brainstem network, central noradrenergic (NA) neurons are known to be an important component that plays a variety of roles in modulating breathing. Furthermore, various perturbations across the central NA system have been implicated in several developmental disorders with respiratory and chemosensory features such as Congenital Central Hyperventilation Syndrome (CCHS), Sudden Infant Death Syndrome (SIDS) and Rett Syndrome (Beltrán-Castillo et al., 2017; Feldman et al., 2013; Gauda et al., 2007; Viemari, 2008). Thus, understanding how central NA neurons modulate respiratory chemoreflexes or chemosensory breathing is critically important for the development of new diagnostic and therapeutic interventions to address respiratory pathophysiology.

Central noradrenergic neurons are commonly thought to exert their effect on breathing through their primary neurotransmitter noradrenaline and adrenaline. However, in addition to noradrenaline, several studies provide strong evidence that glutamate is co-transmitted in subsets of central NA neurons. Vesicular glutamate transporter 2 (Vglut2), a gene marker of glutamatergic signaling, has been shown to be co-expressed in subsets of central NA neurons including, C1/A1, C2/A2, A5, and LC in adult rats and mice (DePuy et al., 2013; Souza et al., 2022a; Stornetta et al., 2002a, 2002b; Yang et al., 2021), though NA specific expression of related Vglut1 and Vglut3 transporters remains unknown. Additionally, it has been well documented that central NA neurons co-expressing Vglut2 innervate key respiratory centers, such as preBötzinger complex and parafacial region (pFRG). Additionally, central NA neurons co-expressing Vglut2 project to many key autonomic brainstem, spinal cord, and forebrain centers, such as the dorsal motor nucleus of the vagus, intermediolateral nucleus (IML) and sympathetic system, and the hypothalamus, all of which could drive a change in breathing if perturbed, i.e., indirect effects from cardio-vascular and metabolic dysregulation (Supplemental table 1). Thus, it has become a dominant paradigm in the field of respiratory physiology that NA-based glutamate release is a critical form of neurotransmission in the control of breathing. Abbott et al. (2014) showed that Vglut2 is required for an increase in respiratory rate when anterior C1 neurons are unilaterally opto-genetically stimulated. Malheiros-Lima et al. (2020) showed that Vglut2-expressing C1 neurons project to the pFRG region, and hypoxic breathing was blunted after blockade of ionotropic glutamatergic receptors at the pFRG site in anesthetized rats, together supporting a role for anterior C1 neurons releasing glutamate at the pFRG site to regulate breathing under hypoxia. Similarly, Malheiros-Lima et al. (2022) and Malheiros-Lima et al. (2018) showed that Vglut2-expressing C1 neurons project to the NA A5 region and the preBötzinger complex, and, again, the blockade of ionotropic glutamatergic receptors at the A5 region or preBötzinger complex reduced the increase in phrenic nerve activity and respiratory frequency elicited by optogenetic stimulation of C1 cells in an anesthetized preparation. In addition, Guyenet et al. (2013) and others speculated that the apparent lack of plasmalemmal monoamine transporter in C1 fibers indicates reduced or absent noradrenergic or adrenergic signaling due a lack of re-uptake and neurotransmitter pool depletion (Comer et al., 1998; Lorang et al., 1994). Cumulatively, the studies argue for glutamate as the predominant functional neurotransmitter for C1 NA neurons in the breathing neural network. To our knowledge, this dominant perspective has not been otherwise previously challenged. Although these studies are informative, the evidence supporting the role of NA-based Vglut2 signaling in respiratory control are either indirect and circumstantial or cannot be seen as physiological given experimental limitations, such as the focal nature of opto-genetic stimulation. Thus, it is not yet clear what the requirement is for NA-based glutamatergic signaling in homeostatic breathing in the unanesthetized and unrestrained animal and how that might inform upon disease.

To better understand the role of NA-based glutamatergic signaling in breathing, we sought to both fully characterize the molecular profiles of the central noradrenergic system with respect to glutamate co-expression and to test the hypothesis that Vglut2-based glutamatergic release is required in respiratory control under physiological chemosensory challenges in unanesthetized and unrestrained mice. We first fully characterized the recombinase-based cumulative fate maps for Vglut1, Vglut2 and Vglut3 expression and compared those maps to their real time expression profiles in central NA neurons by RNA in situ hybridization in adult mice. We found a novel dynamic expression pattern for Vglut2 and an entirely undescribed co-expression domain for Vglut3 in the central NA system. Secondly, to determine if Vglut2-based glutamatergic signaling in NA neurons is required for respiratory homeostasis, we conditionally ablated Vglut2 in all NA neurons and tested respiratory, chemosensory, and metabolic function in unrestrained and unanesthetized mice. Using the same genetic model in prior studies (Abbott et al., 2014), conditional deletion of Vglut2 in NA neurons did not significantly impact breathing under room air, hypercapnic, or hypoxic conditions. These results demonstrate, for the first time, that NA Vglut2-based glutamatergic signaling is dispensable for respiratory control, which challenges the prevalent perspective on the role of C1 and other Vglut2 expressing NA neurons in respiratory homeostasis and suggests that glutamate may not be a critical target to understand NA neuron dysfunction in respiratory diseases.

Results

Cumulative fate maps of central noradrenergic neurons co-expressing Vglut1, Vglut2 and Vglut3

To fully characterize the expression profiles of all three glutamate markers Vglut1, Vglut2 and Vglut3 in central noradrenergic neurons, we first used an intersectional genetic strategy (Figure 1A). We bred three Cre drivers Slc17a7Cre(Vglut1-Cre), Slc17a6Cre (Vglut2-Cre) and Slc17a8Cre(Vglut3-Cre) to Dbhp2a-Flpo (DBH-p2a-Flpo) (targeting noradrenergic neurons) mice, respectively. The three compound lines of Slc17a7Cre; Dbhp2a-Flpo, Slc17a6Cre; Dbhp2a-Flpo, and Slc17a8Cre; Dbhp2a- Flpo were then crossed with the Rosa26RC::FLTG (RC::FLTG). RC::FLTG are intersectional reporter mice that express tdTomato in cells expressing only flippase (Flpo) and express eGFP in cells co-expressing both Flpo and Cre recombinases. Thus, in each of the three intersectional reporter crosses [e.g. 1) Slc17a7Cre; Dbhp2a-Flpo; Rosa26RC::FLTG, 2) Slc17a6Cre; Dbhp2a-Flpo; Rosa26RC::FLTG and 3) Slc17a8Cre; Dbhp2a-Flpo; Rosa26RC::FLTG], NA neurons without any Vglut1, 2, or 3 expression are labelled by red fluorescent protein (tdTomato) while NA neurons co-expressing either Vglut1, Vglut2, or Vglut3 are labeled by green fluorescent protein (eGFP). By using this method, we characterized and quantified the expression profiles of all three vesicular glutamate transporters Vglut1, Vglut2, and Vglut3 across every anatomically defined NA nucleus in adult mice including A7, Locus Coeruleus (LC), dorsal/ ventral subcoeruleus nucleus (sub CD/CV), A5, and the anterior - posterior dimensions of C1/A1, C2/A2 (Robertson et al., 2013), and C3 (Figure 1B-D and Figure 1 supplement 1A-B). Vglut1-Cre was not co-expressed in any central NA nuclei in adult mice. However, Vglut2-Cre and Vglut3-Cre both showed co-expression in central NA neurons. Vglut3-Cre-expressing NA neurons were restricted to posterior C2/A2 and posterior C1/A1, with greatest expression in posterior C2/A2 where 26.9 ± 3.16% of NA neurons were Vglut3-Cre positive. In posterior C1/A1, only 1.26 ± 0.559% of NA neurons were Vglut3-Cre positive. Surprisingly, 84.6 ± 3.75% of NA neurons in total showed Vglut2-Cre co-expression and each NA nucleus was predominantly labelled by Vglut2-Cre expression. Over 50% of NA neurons in each NA nucleus were Vglut2-Cre positive. The percentages of Vglut2-Cre positive NA neurons in each NA nucleus were as follows: 69.8 ± 8.29% in A7, 80.5 ± 3.78% in LC, 84.8 ± 11.3% in A5, 51.5 ± 2.61% in sub CD/CV, 99.0 ± 0.681% in anterior C1/A1, 95.7 ± 2.40% in posterior C1/A1, 96.1 ± 1.98% in anterior C2/A2, 97.8 ± 0.771% in posterior C2/A2, and 100 ± 0.00% in C3 (Mean ± SEM). The presence of Vglut2-Cre co-expression in anterior NA groups was unexpected as previous in situ data in adult rats found Slc17a6 (Vglut2) positive NA neurons only in the posterior C2/A2 and C1/A1 (Stornetta et al., 2002a, 2002b). Notably, however, the Vglut2 co-expression in the LC region agrees with Yang et al. (2021) which also showed that 89% of LC NA neurons are Vglut2 positive using a different intersectional strategy in mice: ThFlpo; Slc17a6Cre; Rosa26Ai65 (TH-Flpo; Vglut2-Cre; Ai65). The differences in our and other fate maps compared to in situ hybridization may either reflect early gene expression that is down regulated in the adult or may reflect low levels of expression not detectable by in situ hybridization but that are nonetheless sufficient to affect recombination in the intersectional genetic strategy.

Cumulative fate maps of glutamate co-expressing NA neurons characterized by intersectional genetics.

(A) Breeding schematic to generate the three intersectional reporter lines Slc17a7Cre; Dbhp2a-Flpo; Rosa26RC::FLTG, Slc17a6Cre; Dbhp2a-Flpo; Rosa26RC::FLTG, Slc17a8Cre; Dbhp2a- Flpo; Rosa26RC::FLTG. In each of the three intersectional reporter crosses, NA neurons co-expressing either Vglut1, Vglut2, or Vglut3 are labeled by green fluorescent protein (eGFP) while NA neurons without any Vglut1,2, or 3 expressions are labelled by red fluorescent protein (tdTomato). Shown as an example is the Slc17a6Cre; Dbhp2a-Flpo; Rosa26RC::FLTG (Vglut2-Cre/+; DBH-p2a-Flpo/+; RC::FLTG/+) intersectional reporter line. (B) Quantification of the percentage of Vglut2-coexpressing NA neurons among NA neurons in each brainstem NA nuclei including A7, LC, A5, sub CD/CV, anterior C1/A1 and C2/A2, posterior C1/A1 and C2/A2. Pink data points represent female data while blue data points represent male data. (C) Quantification of the percentage of Vglut3-coexpressing NA neurons among NA neurons in each NA nuclei. Pink data points represent female data while blue data points represent male data. (D) Fluorescent expression of tdTomato (red) and eGFP (green) in coronal sections of Vglut1, Vglut2, or Vglut3 intersectional reporter lines in brainstem NA nuclei. Nucleus is labeled by DAPI (blue). Scale bar 50μm.

Real time mRNA expression patterns of Vglut1, Vglut2 and Vglut3 in central noradrenergic neurons in adult mice

To verify if the real time NA-based Vglut2 expression pattern in adult mice is comparable to previous reports in adult rats (DePuy et al., 2013; Stornetta et al., 2002b, 2002a), and to further characterize the expression patterns of the other two vesicular glutamate transporters (Vglut1 and Vglut3) in adult mice, we performed fluorescent RNA in situ hybridization experiments. We co-stained Slc17a7 (Vglut1), Slc17a6 (Vglut2), or Slc17a8 (Vglut3) with Dbh respectively in brain tissue of adult mice and characterized the colocalization of Slc17a7, Slc17a6, Slc17a8 with Dbh in brainstem NA nuclei A7, LC, A5, sub CD/CV, anterior C1/A1 and C2/A2, posterior C1/A1 and C2/A2, and C3 (Figure 2A-C and Figure 2 supplement 2A-B). No Slc17a7/Dbh double positive neurons were detected in any part of the central NA system, consistent with our fate map. Slc17a8 mRNA colocalization with Dbh was only found in the posterior part of C2/A2 where 27.1 ± 1.86% of Dbh positive neurons demonstrated Slc17a8 colocalization, again consistent with our fate map. However, we only found detectable levels of Slc17a6 mRNA in the C1/A1, C2/A2 and C3 NA regions, but not in the A7, LC, A5, sub CD/CV. The percentages of Slc17a6/Dbh double positive neurons for regions where detected were as follows: 84.7 ± 5.82% in anterior C1/A1, 66.3 ± 0.335% in anterior C2/A2, 35.7 ± 3.01% in posterior C1/A1, 90.1 ± 2.45% in posterior C2/A2, and 79.7 ± 3.94% in C3. This result is consistent with the previous in situ data in adult rats (Stornetta et al., 2002a, 2002b) suggesting that there is no obvious expression difference of Slc17a6 in the central NA system between mouse and rat. However, the Slc17a6 in situ data did not show expression in anterior NA populations. This difference between our in situ data and fate map data supports our previous hypothesis that many NA neurons expressed Slc17a6 (Vglut2) at some point from early development toward adulthood but the expression of Slc17a6 (Vglut2) is diminished during adulthood. Interestingly, Yang et al. (2021) did show adult Vglut2 co-expression in LC by a viral injection of both Cre and Flpo-dependent eYFP into the LC of a bi-transgenic mouse with both ThFlpoand Slc17a6Cre. To verify this result, we injected an AAV virus containing Cre-dependent tdTomato (pAAV-EF1a-DIO-tdTomato-WPRE) into the LC region of adult Slc17a6Cre mice. Consistent with Yang et al. (2021), we observed sparse Vglut2-Cre and TH-immunopositive double labelled neurons in the LC (Figure 2 supplement 1). Additionally, Souza et al. (2022a) showed about 28.5% of A5 neurons are Slc17a6 mRNA positive in adult rats by RNA scope, a method which is more sensitive than the traditional in situ hybridization we used here. These data suggest that some adult NA neurons in LC and A5 have Slc17a6 (Vglut2) co-expression, but the Slc17a6 mRNA level cannot be detected by traditional in situ hybridization.

Characterization of real time co-expression of Vglut1, Vglut2 and Vglut3 in brainstem NA neurons in adult mice by fluorescent in situ hybridization.

(A) Representative images of Slc17a7 (Vglut1), Slc17a6 (Vglut2), and Slc17a8 (Vglut3) and Dbh (DBH) double ISH in coronal sections of WT mice in all of brainstem NA nuclei in adult mice. DBH (green), Vglut1/2/3 (red), DAPI (cyan). Scale bar 50μm. (B) Quantification of the percentage of Slc17a6 (Vglut2) co-expression in NA neurons in each NA nuclei. Female data (pink), male data (blue). (C) Quantification of the percentage of Slc17a8 (Vglut3) co-expression in NA neurons in each NA nuclei. Female data (pink), male data (blue).

Vglut2 expression is effectively knocked down in the whole central noradrenergic system in Dbh-Cre; Vglut2 cKO mice

To investigate the requirement of Vglut2-based glutamatergic signaling in the central NA system in breathing under physiological challenges, we used Dbh-Cre; Slc17a6flox/flox(Dbh-Cre; Vglut2 cKO) to remove Vglut2 expression from all noradrenergic neurons. This is the same model used by Abbott et al. (2014) to show that Vglut2 expression in the NA system was required for an increase in respiratory frequency following unilateral optogenetic stimulation of anterior C1. To verify if Vglut2 expression was effectively removed from central NA neurons in Dbh-Cre; Vglut2 cKO mice, we used fluorescent mRNA in situ hybridization for Slc17a6 (Vglut2) and Dbh in the mouse brainstems and compared the Slc17a6 (Vglut2) signal intensity in Dbh positive neurons in Dbh-Cre; Vglut2 cKO mice to their littermate controls. We found that Slc17a6 mRNA expression was 96.4 ± 0.226% decreased compared to controls, indicating that the Slc17a6 (Vglut2) was effectively recombined to abrogate expression in NA neurons (Figure 3A-B and Figure 3 supplement 1).

DBH with Vglut2 co-staining by fluorescent in situ hybridization confirms that Vglut2 expression is disrupted in the whole central NA system in Dbh-Cre; Vglut2 cKO mice.

(A) Representative images of Slc17a6 (Vglut2) and Dbh (DBH) double ISH in anterior C1/A1, posterior C1/A1, anterior C2/A2 and posterior C2/A2 in control and Dbh-Cre; Vglut2 cKO mice. DBH (green), Vglut2 (red), DAPI (cyan). Scale bar 20μm.(B) Quantification of intensity of Slc17a6 (Vglut2) signal in control and Dbh-Cre; Vglut2 cKO mice.

Vglut2-based glutamatergic signaling in central NA neurons is not required for baseline breathing nor for the hypercapnic ventilatory reflex under distinct CO2 challenges

Multiple prior studies provide indirect evidence that NA glutamate transmission may play roles in respiratory function, particularly that anterior C1 neurons mediate the hypoxic ventilatory response through the pFRG/RTN and/or the NA A5 group (Malheiros-Lima et al., 2022, 2020). Additionally, the A5 group has been implicated in the hypercapnic reflex (Haxhiu et al., 1996). Thus, we sought to determine if Vglut2-based glutamatergic signaling in central NA neurons is necessary to regulate baseline breathing and the hypercapnic chemoreflex in unanesthetized and unrestrained animals. We used whole-body barometric plethysmography to measure the breathing and metabolism of unanesthetized and unrestrained mice. After a five-day habituation protocol, the respiratory responses of Dbh-Cre; Vglut2 cKO mice and their littermate controls, including the overall respiratory output or ventilatory equivalent of oxygen (VE/VO2), respiratory rate (Vf), tidal volume (VT), minute ventilation (VE), metabolic demand (VO2), inspiratory duration (TI), expiratory duration (TE), breath cycle duration (TTOT), inspiratory flow rate (VT/TI), and expiratory flow rate (VT/TE), were measured under room air and hypercapnia (5% CO2) (Figure 4A). 5% CO2 gas challenge produced a significant change in all the respiratory parameters compared to room air in both Dbh-Cre; Vglut2 cKO and control mice. However, surprisingly, Dbh-Cre; Vglut2 cKO mice did not show significant differences in any respiratory or metabolic parameter mentioned above compared to their littermate controls under both room air and 5% CO2 challenges (Figure 4B and Table 1-2). In addition, to investigate if Vglut2-based NA derived glutamate is required for regulating the dynamic patterns of breathing, such as the breathing rhythms, we generated Poincaré plots for TI, TE, TTOT, VT/TI, VT/TE, Vf, VT, and VE and we calculated the SD1 and SD2 statistics for each parameter under room air and 5% CO2. We found that Dbh-Cre; Vglut2 cKO mice only showed a marginally statistically significant increase in the SD1 of inspiratory duration (TI) (p=0.0492). Outside of this single comparison, however, we found no significant differences in the other parameters between the two groups (Figure 4 supplement 1A-H). Thus, we theorize that the statistically significant result for the SD1 of TI seen here is likely either a result of measurement noise or a consequence of a type 1 error. The overall pattern of results suggested that Vglut2-based glutamatergic signaling is not required for regulating either the steady state of or the dynamic patterns of baseline breathing and hypercapnic ventilatory reflex under 5% CO2.

Vglut2 conditional knockout in central noradrenergic neurons fails to alter baseline breathing and the hypercapnic ventilatory reflex (5% CO2).

(A) Mouse model schematic and experimental protocol including habituation and hypercapnia protocol (5% CO2). (B) Under both room air and hypercapnia (5% CO2), Dbh-Cre; Vglut2 cKO mice didn’t show significant changes in respiratory output (VE/VO2), minute ventilation (VE), respiratory rate (Vf), tidal volume (VT), metabolism demand (VO2), inspiratory duration (TI), expiratory duration (TE), breath cycle duration (TTOT), inspiratory flow rate (VT/TI), and expiratory flow rate (VT/TE). Linear mixed-effects regression model, ****p<0.0001, **p<0.01, *p<0.05, ns: p20.05.

To further investigate the role of Vglut2-based glutamatergic signaling in the hypercapnic response, we interrogated the system further with more severe hypercapnic exposures and measured the breathing of Dbh-Cre; Vglut2 cKO mice under 7% CO2 and 10% CO2. Both 7% and 10% CO2 stimuli produced a significant respiratory response in both mutant and control mice. Under the 7% CO2 condition, Dbh-Cre; Vglut2 cKO mice showed a significantly decreased tidal volume (VT) (p=0.023) and a significantly reduced inspiratory flow rate (VT/TI) (p=0.0016), but the overall respiratory output (VE/VO2), respiratory rate (Vf), minute ventilation (VE), metabolic demand (VO2), inspiratory duration (TI), expiratory duration (TE), breath cycle duration (TTOT), and expiratory flow rate (VT/TE) did not show a significant difference between the mutant mice and their sibling controls (Figure 5A-B and Table 3). Additionally, none of the dynamic patterns of any respiratory parameter showed a significant difference between the mutant and control groups (Figure 5 supplement 1A-H). This result suggests that Vglut2-based glutamatergic signaling in central NA neurons is not required for regulating the hypercapnic ventilatory reflex and breathing regularity under 7% CO2.

Vglut2 conditional knockout in central noradrenergic neurons fails to alter the majority of hypercapnic ventilation parameters and metabolism under 7% CO2.

(A) Mouse model schematic and experimental protocol including habituation and hypercapnia protocol (7% CO2). (B) Knocking out Vglut2 in the whole NA system failed to alter VE/VO2, Vf, VE, VO2, TI, TE, TTOT, VT/TE but only showed a significantly reduce in VT and VT/TI under 7% CO2 condition. Linear mixed-effects regression model, ****p<0.0001, **p<0.01, *p<0.05, ns: p20.05.

Under 10% CO2 challenge, none of the steady state breathing parameters including the overall respiratory outputs (VE/VO2), respiratory rate (Vf), tidal volume (VT), minute ventilation (VE), metabolic demand (VO2), inspiratory duration (TI), expiratory duration (TE), breath cycle duration (TTOT), inspiratory flow rate (VT/TI), and expiratory flow rate (VT/TE) were significantly different between Dbh-Cre; Vglut2 cKO mice and their littermate controls, suggesting that Vglut2-based glutamatergic signaling in central NA neurons is not required for the hypercapnic ventilatory reflex under 10% CO2 (Figure 6A-B and Table 4). For the dynamic patterns of breathing, the SD1 of breath cycle duration (TTOT) (p=0.028) and SD1 of respiratory rate (Vf) (p=0.038) showed a significant increase in mutant mice compared to the controls (Figure 6 supplement 1A-H). This suggests that removing Vglut2 from central NA neurons may increase the breathing irregularity under high CO2 challenges like 10% CO2.

Vglut2 conditional knockout in central noradrenergic neurons fails to alter the hypercapnic ventilatory reflex (10% CO2).

(A) Mouse model schematic and experimental protocol including habituation and hypercapnia protocol (10% CO2). (B) Knocking out Vglut2 in the whole NA system didn’t significantly alter VE/VO2, Vf, VT, VE, VO2, TI, TE, TTOT, VT/TI, and VT/TE under 10% CO2 condition. Linear mixed-effects regression model, ****p<0.0001, **p<0.01, *p<0.05, ns: p20.05.

Vglut2-based glutamatergic signaling in central NA neurons is not required for the hypoxic ventilatory reflex under 10% O2

To determine if Vglut2-based glutamatergic signaling in central NA neurons is necessary to regulate the hypoxic chemoreflex, we measured the ventilation response of Dbh-Cre; Vglut2 cKO mice under 10% O2. We analyzed the ten respiratory parameters in three 5-min epochs following 10% O2 challenge onset. Similar to the hypercapnic challenges, the 10% O2 challenge produced a significant breathing response in every 5-min epoch compared to room air, but no significant difference was found in the overall respiratory output (VE/VO2), respiratory rate (Vf), tidal volume (VT), minute ventilation (VE), metabolism demand (VO2), inspiratory duration (TI), expiratory duration (TE), breath cycle duration (TTOT), inspiratory flow rate (VT/TI), and expiratory flow rate (VT/TE) between Dbh-Cre; Vglut2 cKO mice and controls in any of the three 5-min 10% O2 epochs (Figure 7A-B and Table 5). These data suggest that Vglut2-based glutamatergic signaling in central NA neurons is not required for the hypoxic ventilatory reflex under 10% O2. For the dynamic patterns of breathing, the SD2 of inspiratory flow rate (VT/TI) (p=0.005) and the SD2 of minute ventilation (VE) (p=0.026) were significantly increased during the first epoch (first 5 minutes) under the 10% O2 challenge in mutant mice, suggesting that Vglut2-based glutamate may play a role in maintaining breathing regularity during the initial, or reflexive, response to hypoxia (Figure 7 supplement 1A-H).

Vglut2 conditional knockout in central noradrenergic neurons fails to alter the hypoxic ventilatory reflex (10% O2).

(A) Mouse model schematic and experimental protocol including habituation and hypoxia protocol (10% O2). (B) Knocking out Vglut2 in the whole NA system did not show significant breathing changes including VE/VO2, VE, Vf, VT, VO2, TI, TE, TTOT, VT/TI, and VT/TE in either of the three five-minute time periods under hypoxia (10% O2). Linear mixed-effects regression model, ****p<0.0001, **p<0.01, *p<0.05, ns: p20.05.

Discussion

Neuron co-transmission of two or more neurotransmitters across excitatory, inhibitory, and neuro-modulatory facets is becoming an increasingly appreciated phenomenon in the central nervous system. Functional interrogation of co-transmission is essential not only to understand the distinct and cooperative roles that multiple signaling molecules may play at the synapse but also to determine the most critical targets for potential therapeutics. To our knowledge, the unopposed paradigm in the field until now has been that vesicular glutamate transporter 2 (Vglut2)-based glutamate transmission from central noradrenergic neurons is important in respiratory homeostasis. Anatomically, it has long been appreciated that central NA neurons co-express Vglut2, the major glutamate marker among three glutamate transporters (DePuy et al., 2013; Souza et al., 2022a; Stornetta et al., 2002a, 2002b; Yang et al., 2021). Also, it has been well documented that NA Vglut2 positive fibers innervate known central respiratory centers or other autonomic centers, that, when perturbed, could potentially disrupt respiratory homeostasis (Supplemental Table 1). Functionally, multiple reports have suggested that Vglut2-based glutamate transmission plays a role in respiratory homeostasis (Abbott et al., 2014; Malheiros-Lima et al., 2022, 2020, 2018). It has also been suggested that, at least for C1 neurons, the apparent lack of a plasmalemmal monoamine transporter may attenuate or eliminate noradrenergic or adrenergic release from these fibers (Guyenet et al., 2013). However, as discussed below, much of the functional evidence supporting a role for Vglut2-based glutamate transmission from central noradrenergic neurons in breathing is circumstantial or of a non-physiological nature, and absent or attenuated release of adrenaline and noradrenaline has not been demonstrated. Nonetheless, the dominant perspective that NA-Vglut2 glutamate transmission plays a role in breathing has remained unchallenged. Our studies, in contrast to prior work, show that loss of Vglut2 in NA neurons does not appreciably change baseline or chemosensory breathing in unanesthetized and unrestrained mice. Additionally, our work uncovers a novel dynamic expression pattern for Vglut2 and an entirely undescribed co-expression domain for Vglut3 in central NA neurons.

Vglut2 has been shown to be co-expressed in subsets of central NA neurons, however, to our knowledge, the potential for the central NA neurons to express the other glutamate transporters Vglut1 or Vglut3 has not been reported at the anatomical level. To determine if central NA neurons express either Vglut1 or Vglut3 and to confirm the Vglut2 expression, we carried out cumulative intersectional fate mapping using Slc17a7Cre; Dbhp2a-Flpo; Rosa26RC::FLTG, Slc17a6Cre; Dbhp2a-Flpo; Rosa26RC::FLTG and Slc17a8Cre; Dbhp2a-Flpo; Rosa26RC::FLTG compound mouse lines. As this approach maps the cumulative history of expression, we also examined acute adult expression of Slc17a7 (Vglut1), Slc17a6 (Vglut2) and Slc17a8 (Vglut3) using fluorescent RNA in situ hybridization. Our in situ results for Slc17a6 (Vglut2) agreed with earlier published outcomes (DePuy et al., 2013; Stornetta et al., 2002a, 2002b) showing as much as 80% expression in posterior NA neurons. However, there was no appreciable signal in anterior groups (A7, LC, A5, sub CD/CV). This lack of expression was in stark contrast to the cumulative fate map for Vglut2, showing as much as 75% of the anterior central NA neurons were recombined by the Vglut2-Cre over the lifetime of the animal. To further interrogate the acute activity of the Vglut2-Cre in the Locus Coeruleus in the adult mouse, we injected a Cre responsive virus in LC of Slc17a6Cre mice and found sparse Vglut2-Cre positive NA neurons in the LC. The outcomes from the cumulative fate map and acute injections are in agreement with published results from Yang et al. (2021). Notably, our use of the Dbhp2a-Flpoavoided expression in the parabrachial nucleus (PBN), which was a confounding variable in the ThFlpo mouse. In addition, 28.5% of A5 neurons were shown to be Slc17a6 (Vglut2) positive in adult rats by RNA scope (Souza et al., 2022b). It is notable that both our lab and the work published by (Stornetta et al., 2002a, 2002b) failed to detect Slc17a6 (Vglut2) expression in anterior groups, but that apparently very low levels can be detected by more sensitive recombinase and RNA scope methods. Additional electrophysiological evidence suggests that this expression is functional (Yang et al., 2021). However, it is notable that we see nearly 80% of the LC and 85% of A5 recombined in the intersectional strategy representing lifetime expression, but our study and others find only 20% of the LC and 28.5% of A5 may be expressing Slc17a6 (Vglut2) at any given moment in adults. Thus, it suggests that Vglut2 co-expression in anterior NA groups is time dependent, though the underlying temporal dynamics of Vglut2 remain unknown. We hypothesize that this could be either a temporal restriction in expression as the animal matures or may reflect dynamic change driven by behavioral or physiological experiences, which could be tested with the use of inducible conditional strategies.

For Slc17a7 (Vglut1), neither fate map nor in situ hybridization revealed any co-expression within the central NA system. However, Slc17a8 (Vglut3) expression was found in the posterior part of NA neurons including C2/A2 and C1/A1. This study is the first to characterize this expression. The majority of this Vglut3 co-expression is restricted to the posterior part of the C2/A2 NA population. This NA region has not been heavily implicated in respiratory control nonetheless it readily lends itself to interrogation by intersectional genetic methods and in situ hybridization. Given its distal expression from anterior C1, it is less likely that Vglut3 would compensate for loss of Vglut2 in anterior C1, though this has not been ruled out. Further, while Vglut3 is a marker of glutamatergic neurons, its role in neurotransmission is not wholly defined as Vglut3 is typically found in soma and dendrites (Fremeau et al., 2004).

The cumulative fate map and in situ hybridization experiments were performed in both females and males and no obvious sex difference was observed in the expression pattern of Vglut1, Vglut2 or Vglut3. However, only three mice were characterized in each dataset, thus, larger sample sizes may capture more subtle outcomes for potential sex differences.

Based on the Vglut2 co-expression in NA neurons and their projections to multiple brain regions important in respiration, several studies suggest that glutamate in NA populations is important in breathing control. However, the evidence those studies provided is circumstantial, indirect, or limited due to experimental caveats or technical limitations. Abbott et al. (2014), using exactly the same mouse model as this study, showed that conditionally knocking out Vglut2 in the whole central NA system did not significantly change respiratory rate under room air, which is consistent with our findings here. Our results differ in that they reported removing Vglut2 from central NA neurons attenuated an increase in respiratory rate resulting from unilateral C1 optogenetic stimulation. However, while the optogenetic stimulation of C1 neurons did drive a Vglut2 dependent increase in respiratory rate, it is not clear if the resulting change in breathing reflects a native feature of the network that might be engaged in chemosensory responses or other aspects of breathing. It remains possible that strong unilateral optogenetic stimulation (up to 20Hz) of a small, isolated population drives a Vglut2 dependent, but ectopic, function that is not typically engaged in the normal operation of the breathing network. Such ectopic outcomes could come from several potential, non-exclusive mechanisms. First, unilateral stimulation could drive a network hysteresis or dysregulation similar to a focal injury that is not a typical biological feature (Ducros et al., 2003). This point is well made in Abbott et al. (2013) in their discussion of C1 optogenetic stimulation where they state, “resulting cardiorespiratory response pattern should not be considered strictly “physiological,” because a physiological response is never initiated by selective activation of a single cluster of CNS neurons and the various subsets of C1 neurons are presumably never recruited en bloc under any physiological condition”. Second, strong stimulation could result in overwhelming synaptic mechanisms (i.e., glial uptake) resulting in glutamate spillover and unintended cross talk to affect extra-synaptic neurons proximal to targeted NA fibers. Activity dependent spillover transmission has been documented elsewhere in the nervous system (Henneberger et al., 2020; Hülsmann et al., 2000). Third, the outcomes could be secondary to a different autonomic or behavioral function requiring NA-derived glutamatergic signaling involving metabolism or circulation (DePuy et al., 2013; Yang et al., 2021). Additionally, respiratory parameters except for respiratory rate (Vf), such as tidal volume (VT) and minute ventilation (VE) were not reported, making comprehensive comparison difficult.

More circumstantially, Malheiros-Lima et al. (2020) suggested C1 neurons potentially release glutamate at the pFRG site to regulate active expiration under hypoxia in anesthetized rats by providing three indirect lines of evidence: 1) Vglut2-expressing C1 neurons project to the pFRG region; 2) Increased abdominal expiratory nerve activity (AbdEMG) was blunted after blockade of ionotropic glutamatergic receptors at the pFRG site under anesthesia using cytotoxic KCN mediated hypoxia; and 3) Depletion of C1 neurons eliminated the increased AbdEMG elicited by hypoxia. However, 1) the glutamatergic signaling targeting the pFRG region is not necessarily from C1 NA neurons since glutamatergic neurons from other regions also project to the pFRG site (Yang and Feldman, 2018). Thus, it is possible that the glutamatergic signaling that is required for hypoxic response is derived from other glutamatergic neurons such as the preBötzinger complex and the RTN itself. Furthermore, 2) C1 neuron depletion knocks out all the signaling modalities in the C1 population including both noradrenaline and glutamate. It is not clear which signaling drives the hypoxic response. Finally, 3) KCN injection mimics a hypoxic challenge by activating peripheral chemoreceptors, but the physiological responses differ from those seen during environmental hypoxia exposure. Thus, the abdominal activity changes due to KCN injection may not be recaptured under physiological hypoxic challenge.

Similarly, Malheiros-Lima et al. (2022) and Malheiros-Lima et al. (2018) showed that Vglut2-expressing C1 neurons project to both the NA A5 region and the preBötzinger complex. The blockade of ionotropic glutamatergic receptors at the A5 region or preBötzinger complex reduced the increase in phrenic nerve activity and the breathing frequency caused by optogenetic stimulation of C1 cells in an anesthetized preparation. Together these results suggest that Vglut2-expressing C1 neurons communicate with A5 or preBötzinger complex neurons by releasing glutamate in turn to regulate phrenic nerve activity or breathing frequency. Again, these two studies showed that C1 neurons are important to regulate phrenic nerve activity or breathing frequency, however, it is not necessarily through a direct C1-A5 or C1-preBötzinger complex glutamatergic pathway, whereas an indirect multi-synaptic pathway originating with C1 noradrenaline and ending with glutamate transmission from an intermediate to A5 or preBötzinger complex cannot be ruled out. Importantly, in all four studies noted here, respiratory measurements were limited or proxies of breathing under anesthesia were used. The role of NA-Vglut2 signaling was not reported in the context of an unanesthetized, unrestrained animal under normoxic, hypoxic, or hypercapnic breathing in these previous studies.

Thus, the limitations and indirect lines of the evidence in these studies raised the question as to whether or not glutamatergic signaling in central NA neurons is required or necessary to modulate respiratory homeostasis under more physiologically relevant circumstances and if NA-derived glutamate is an important mechanism to understand the pathophysiology of NA neurons in respiratory diseases. The goal of our study was to test the requirement of NA-based glutamatergic signaling in breathing homeostasis in the unanesthetized and unrestrained animal under conditions that would engage the whole respiratory control network rather than a singular component that may drive hysteresis or artifactual outcomes. Our in vivo breathing data under room air, hypoxic, and multiple hypercapnic conditions failed to show the requirement of NA-derived Vglut2 in normal breathing and respiratory chemoreflexes directly. For the dynamic patterns of breathing, we measured the SD1 and SD2 of eight parameters under each gas condition. We found no consistent change across multiple hypercapnic challenges or hypoxia. We did find that under severe hypercapnic challenge (10% CO2), NA derived Vglut2 removal resulted in an increase in the SD1 of breath cycle duration and respiratory rate, which was not observed under 5% and 7% CO2 conditions. Under hypoxia (10% O2), Dbh-Cre; Vglut2 cKO mice only showed an increase in the SD2 of inspiratory flow rate and minute ventilation during the first 5-minute of 10% O2 exposure. These findings may suggest that Vglut2-based glutamatergic signaling in central NA neurons may play a role in modulating breathing regularity under high CO2 challenge and hypoxia. However, the phenotypes were of a small magnitude and not consistent across challenges, and not at all seen under homeostatic breathing.

To ensure a rigorous and robust conclusion, we included several extra precautions and additional measures in our experimental design and analysis. First, we use exactly the same mouse model, DBH-Cre; Vglut2 cKO, as Abbott et al. (2014) used, and confirmed, by a second method (fluorescent RNA in situ hybridization) that Vglut2 expression was indeed abrogated in the whole central NA system. Our results are in strong agreement with DePuy et al. (2013) that recombination is efficient. Notably, we assayed for expression at the cell body, rather than the fibers and see no singular cell body that is scored positive. Second, we measured the breathing of unanesthetized, unrestrained, and habituated mice under physiological challenges including normoxia, hypercapnia and hypoxia. Third, to visualize breathing function comprehensively, the metabolic parameter and multiple respiratory parameters (both the steady state and the dynamic patterns) were measured and reported in absolute values including oxygen consumption, respiratory rate, tidal volume, minute ventilation, overall respiratory output (minute ventilation normalized to oxygen consumption), inspiratory duration, expiratory duration, breath cycle duration, inspiratory flow rate, and expiratory flow rate. Fourth, our experimental design was overpowered (i.e., n=16-21 vs 5-13 needed for power based on power analysis and n=7-8 reported in other manuscripts). Each respiratory/metabolic parameter for each physiological condition including room air, hypercapnia, and hypoxia was derived by quantifying the entire respiratory trace and analyzed between mutant and sibling controls by using a linear mixed-effects regression model with animal type (experimental vs. control) as fixed effects, animal ID as a random effect, and sex as an experimental covariate (for which we did not find a significant effect) (Lusk et al., 2023).

Despite our experimental design and the extensive nature of our measurements, there remain several possibilities or considerations in the interpretation of our data. The first possibility is Vglut3 in posterior CA domain (first characterized in this study) compensates for the effect of the loss of Vglut2 in NA neurons in breathing. However, we argue that this is unlikely because 1) Our model is the same used in the other study where a Vglut2-dependent effect was seen with optogenetic stimulation. 2) Vglut3 is expressed in the posterior C2/A2 NA neurons which are anatomically distant from anterior C1 and the LC (the two most likely NA candidates with Vglut2 co-expression to impact breathing). Also, posterior C2/A2 hasn’t been heavily implicated in breathing regulation. Future work will need to further investigate the role of Vglut3 in posterior CA in respiratory control in order to test this possibility. The second possibility is that pre- or post-natal developmental compensation corrects for the loss of NA derived glutamate. Again, we argue that is unlikely, as 1) we are using the same adult model that provided the most direct evidence and first functionally demonstrated a potential role for C1-derived glutamate in breathing; 2) Abbott et al. (2014) highlight that C1 NA neurons showed no obvious abnormalities in number, morphology, and projection patterns after knocking out Vglut2; and 3) we have examined the requirement of NA-expressed Vglut2 in P7 neonate mice in the autoresuscitation reflex and saw no differences (data not shown). However, it is still possible that developmental compensation occurs before P7, which is something that could be better tested by using the Dbh-CreERT2 (Stubbusch et al., 2011) to more acutely remove Vglut2 expression (though this leaves a window of 2-3 weeks for compensation to occur). Third, as we only address the loss of NA-Vglut2 based signaling across the entire NA system, it remains a formal possibility that this ultimately results in the removal of Vglut2 dependent signaling from two counter balancing regions of the NA system and are therefore left with null results whereas removal from anterior C1 alone would show a requirement. To test this possibility, it would require an intersectional conditional knockout/loss of function approach, something that has not yet been shown to be efficient and effective and would be beyond the scope of these studies. Fourth, it is possible that another neurotransmitter/neuropeptide within NA neurons might be released in higher amounts in Dbh-Cre; Vglut2 cKO mice to compensate for the deficiency of glutamate in breathing. Loss of Vglut2 could reduce dopamine release in subsets of dopaminergic neurons (Alsiö et al., 2011; Fortin et al., 2012; Hnasko et al., 2010). Thus, it remains possible that loss of Vglut2 affects dopamine loading into the vesicles and in turn affecting noradrenaline release as Dopamine Beta Hydroxylase (DBH) synthesizes noradrenaline from dopamine inside the vesicles (Weinshenker, 2007). Changes in noradrenaline release in Vglut2 negative NA neurons could be further examined with fast-scan cyclic voltammetry (FSCV) or microdialysis. Neuropeptide Y (NPY) and galanin have been shown to co-exist in LC NA neurons and are involved in regulating energy metabolism, stress, or anxiety (Ruohonen et al., 2012; Tasan et al., 2010; Weinshenker and Holmes, 2016). In our studies, we account for potential metabolic changes that may underlie or be coordinate with changes in breathing. However, we did not test for potential changes in blood pressure or other autonomic functions as well as affective state that may, in some obscure way, compensate for changes in breathing. Further interrogation of NPY or galanin signaling across NA neurons and potential compensatory effects in breathing through changes in metabolism or other autonomic function under stress or anxiety may yield notable insights. Lastly, our results are loss of function in nature. We do not test sufficiency or gain of function and cannot fully rule out a role for NA-Vglut2 based glutamatergic signaling in the control of breathing.

From a translational perspective, our data question whether or not glutamatergic signaling in NA neurons is likely to be a key mechanism and therefore a therapeutic target for breathing disorders, such as Rett syndrome and SIDS. Mecp2-deficient mice (a mouse model of Rett Syndrome that phenocopies many human symptoms) showed both a deficiency of NA populations (reduced number of NA neurons in C2/A2 and C1/A1 group) and highly variable respiratory rhythm at around 4-5 weeks of age (Roux et al., 2007; Viemari et al., 2005). 80% of Rett Syndrome patients experience breathing issues, such as unstable breathing, episodes of hyperventilation, and breath holds, throughout their lifespan (Ramirez et al., 2020). SIDS decedents show NA abnormalities and are hypothesized to ultimately succumb from a failure in cardiorespiratory autoresuscitation (Chigr et al., 1989; Garcia et al., 2013; Kopp et al., 1993; Mansouri et al., 2001; Ozawa et al., 2003; Takashima and Becker, 1991). However, Dbh-Cre; Vglut2 cKO mice, have normal baseline and chemosensory respiratory parameters as adults and neonate mice show normal autoresuscitation indicating that perturbed NA based glutamatergic signaling may not be a key driver in these or other related respiratory pathophysiologies.

In conclusion, our studies show that Vglut2 has a dynamic and extensive expression profile across the central NA system. We were able to characterize a novel Vglut3-expressing NA population in the posterior C2/A2 nuclei. Despite prior studies providing indirect and circumstantial evidence that noradrenergic-based glutamatergic signaling may play a role in control of breathing, our conditional loss of function studies in adult unanesthetized and unrestrained mice failed to consistently and significantly change room air breathing, the hypercapnic ventilatory reflex, and the hypoxic ventilatory reflex. These outcomes offer a contrasting perspective from the current field view and provides further insight into the potential role of NA-glutamate transmission in the control of breathing and NA neuron dysfunction in respiratory disorders.

Materials and methods

Ethical approval

Studies were approved by the Baylor College of Medicine Institutional Animal Care and Use Committee (IACUC) under protocol AN-6171, and all experiments reported here were performed in accordance with relevant guidelines and regulations.

Breeding, genetic background, and maintenance of mice

We maintained all our heterozygous mouse strains by backcrossing to wildtype C57BL/6J mice and homozygous mouse strains by sibling crosses. For immunofluorescence experiments, heterozygous B6;129S-Slc17a7tm1.1(cre)Hze/J (Slc17a7Cre, Vglut1-Cre) (Jax Stock No: 023527), Slc17a6tm2(cre)Lowl/J (Slc17a6Cre, Vglut2-Cre) (Jax Stock No: 016963) and B6;129S-Slc17a8tm1.1(cre)Hze/J (Slc17a8Cre, Vglut3-Cre) (Jax Stock No: 018147) were mated with heterozygous Dbhem2.1(flpo)Rray (Dbhp2a-Flpo, DBH-p2a-Flpo) (MMRRC ID: 41575) (Sun and Ray, 2016) mice respectively to derive compound lines with both Cre and Flpo alleles (Slc17a7Cre; Dbhp2a-Flpo, Slc17a6Cre; Dbhp2a-Flpo, and Slc17a8Cre; Dbhp2a-Flpo). Then these three compound lines were each mated with homozygous B6.Cg-Gt(ROSA)26Sortm1.3(CAG-tdTomato,-EGFP)Pjen/J (Rosa26RC::FLTG) (Jax Stock No: 026932) to derive three different intersectional mouse lines Slc17a7Cre; Dbhp2a-Flpo; Rosa26RC::FLTG, Slc17a6Cre; Dbhp2a-Flpo; Rosa26RC::FLTGand Slc17a8Cre; Dbhp2a-Flpo; Rosa26RC::FLTG. For in situ hybridization experiments, wildtype C57BL/6J mice were ordered from the Center of Comparative Medicine (CCM), Baylor College of Medicine. For NA-Vglut2 conditional loss of function in situ hybridization and plethysmography experiments, hemizygous transgene Tg(Dbh-cre)KH212Gsat (Dbh-Cre) (MMRRC ID: 036778-UCD GENSAT) mice were mated with homozygous Slc17a6tm1Lowl/J (Slc17a6flox/flox) (Jax Stock No: 012898) to derive Dbh-Cre; Slc17a6flox/+. Dbh-Cre; Slc17a6flox/+mice were mated with Slc17a6flox/flox to derive Dbh-Cre; Slc17a6flox/flox(Dbh-Cre; Vglut2 cKO). Sibling mice that lacked the Cre allele or carried the Cre allele but lacked the floxed Vglut2 alleles were used as controls. Rosa26 specific primers for the Rosa26RC::FLTG mice were 5’-GCACTTGCTCTCCCAAAGTC, 5’-GGGCGTACTTGGCATATGAT, and 5’-CTTTAAGCCTGCCCAGAAGA (Ray et al., 2011) and yield a 495 bp band (targeted) and 330 bp band (wt). Cre-specific primers for all Cre drivers were 5’-ATCGCCATCTTCCAGCAGGCGCACCATTGCCC and 5’-GCATTTCTGGGGATTGCTTA and yielded a 550 bp band if positive. Flpo-specific primers for Dbhp2a- Flpo are 5’-CACGCCCAGGTACTTGTTCT and 5’-CCACAGCAAGAAGATGCTGA (Sun and Ray, 2016) and yielded a 226 bp band if positive. Slc17a6flox specific primers for Vglut2-floxed mice are available at The Jackson Laboratory website (https://www.jax.org/strain/012898).

Immunofluorescence staining

Slc17a7Cre; Dbhp2a-Flpo; Rosa26RC::FLTG, Slc17a6Cre; Dbhp2a-Flpo; Rosa26RC::FLTGand Slc17a8Cre; Dbhp2a- Flpo; Rosa26RC::FLTG adult mice of both sexes were sacrificed and transcardially perfused with 0.1 M phosphate-buffered saline (PBS) then with 4% paraformaldehyde (PFA) in PBS. Mouse brains were dissected out and fixed for 2h in 4% PFA before a PBS rinse and dehydration in 30% sucrose in PBS. Brains were embedded in OCT blocks, sectioned at the thickness of 30μm, mounted on slides, and stored at −80°C until they were ready for staining. The slides were hydrated in 0.1% Triton-X in PBS (PBST) for 15 mins, blocked with 5% donkey serum in 0.1% PBST for 1h at room temperature and then incubated with primary antibodies for 72 hours at 4°C in 0.1% PBST with 5% donkey serum. Tissues were washed in 0.1% PBST 3 times for 10 min each and then incubated with secondary antibodies for 2hs at room temperature in 0.1% PBST with 5% donkey serum. Slides were washed with 0.1% PBST for 10 min and washed in PBS twice for 10 min each, stained for DAPI, washed 3 times for 10 min each with PBS and mounted in ProLong Glass (Invitrogen). The following primary and secondary antibodies were used: chicken anti-GFP (1:1,000, Abcam ab13970), rabbit anti-dsRed (1:1,000, Clontech 632496), donkey anti-chicken Cy2 (1:500, Jackson 703-225-155), donkey anti-rabbit Cy3 (1:500, Jackson 711-165-152).

In situ hybridization

Mouse brains were dissected out from adult mice of both sexes with 6-8 week of age, sectioned into 25µm brain sections and mounted on slides. We generated a digoxigenin (DIG)-labeled mRNA antisense probe against Slc17a7 (Vglut1), Slc17a6 (Vglut2), and Slc17a8 (Vglut3) and fluorescein (FITC)-labeled mRNA against Dbh using reverse-transcribed mouse cDNA as a template and an RNA DIG or FITC-labeling kits from Roche (Sigma). Primer and probe sequences for the Slc17a7 (Vglut1), Slc17a6 (Vglut2), and Slc17a8 (Vglut3) and Dbh probes are available in the Allen Brain Atlas (http://www.brain-map.org). For the Slc17a6 (Vglut2) and Dbh double ISH in Dbh-Cre; Vglut2 cKO and their littermate controls, we generated a new Slc17a6 (Vglut2) probe targeting exon 2 of Slc17a6 specifically (Tong et al., 2007) and the probe sequence is 892-1144bp as Slc17a6 transcript variant 1 and the size is 253bp. ISH was performed by the RNA In Situ Hybridization Core at Baylor College of Medicine using an automated robotic platform as previously described (Yaylaoglu et al., 2005) with modifications of the protocol for double ISH. Modifications in brief [see Yaylaoglu et al. (2005) for buffer descriptions]: both probes were hybridized to the tissue simultaneously. After the described washes and blocking steps the DIG-labeled probes were visualized using tyramide-Cy3 Plus (1/75 dilution, 15-minute incubation, Akoya Biosciences). After washes in TNT, the remaining HRP-activity was quenched by a 10-minute incubation in 0.2M HCl. The sections were then washed in TNT, blocked in TNB for 15 minutes, and incubated at room temperature for 30 minutes with HRP-labeled sheep anti-FITC antibody (1/500 in TNB, Roche/Sigma). Following washes in TNT, the FITC-labeled probe was visualized using tyramide-FITC Plus (1/50 dilution, 15-minute incubation, Akoya Biosciences). Following washes in TNT, the slides were stained with DAPI (invitrogen), washed again, removed from the machine, and mounted in ProLong Diamond (Invitrogen).

Viral injection

To verify Vglut2 co-expression in LC in adult mice (Yang et al., 2021), 6-week-old Slc17a7Cre mice were injected with pAAV-EF1a-DIO-tdTomato-WPRE virus (RRID:Addgene_133786, obtained from Joshua Ortiz at the Optogenetics and Viral Vectors Core at the Jan and Dan Duncan Neurological Research Institute, 1ul at 7.90E+10 Gc/mL) into the LC (coordinates from bregma anteroposterior −5.4 mm, lateral + 0.8 mm and dorsoventral – 4.0 mm) and allowed to incubate for 4 weeks.

Plethysmography

Plethysmography on unanesthetized and free-moving mice was carried out as described in (Ray et al., 2011). 6–8-week-old adult mice of both sexes with minimum group sizes n=16 were used in both the experimental and control group for each experiment. Animals were subjected to a five-day habituation protocol with each day including several minutes of handling, temperature taken by rectal probe and at least 30 min exposure in the plethysmography chamber (Martinez et al., 2019). Plethysmography started to be performed on the sixth day and was finished for all animals within a week of the last day of habituation (Maximum 6 mice can be assayed for plethysmography per day due to the limited number of plethysmography rigs). On the day of testing, mice were taken out from their home cage, weighed, and rectal temperature was taken. Animals were then placed into a flow-through, temperature-controlled (about 32°C with real-time temperature recording) plethysmography chamber and allowed to acclimate for 20-40 mins in room air (21% O2/79% N2) conditions. After acclimation (indicated by a steady respiratory trace free from movement artifact), a 20-min baseline breathing trace was taken under room air, then the chamber gas was switched to a hypercapnic or hypoxic mixture of 5% CO2/21% O2/74% N2, 7% CO2/21% O2/72% N2, 10% CO2/21% O2/69% N2, or 10% O2/90% N2, depending on the protocol, for 20 min. Chamber gas was then switched back to room air for another 20 min. The animals were removed from the chamber and rectal temperature was measured immediately after the mice were taken out from the chamber. Each testing period for an individual mouse was separated by 24hs to allow for a full recovery.

Plethysmography data analysis and statistics

Details were previously described in (Martinez et al., 2019; Ray et al., 2011). Plethysmography pressure changes were measured using a Validyne DP45 differential pressure transducer, CD15 carrier demodulator and a reference chamber, and were recorded with LabChart Pro in real time. Respiratory waveforms were analyzed by the SASSI module of the Breathe Easy software to determine respiratory frequency (Vf), tidal volume (VT), minute ventilation (VE), oxygen consumption (VO2), ventilatory equivalents for oxygen (VE/VO2), inspiratory duration (TI), expiratory duration (TE), breath cycle duration (TTOT), inspiratory flow rate (VT/TI), and expiratory flow rate (VT/TE) (Lusk et al., 2023). Poincaré plots for TI, TE, TTOT, VT/TI, VT/TE, Vf, VT, VE and their SD1 and SD2 properties were generated and characterized by the STAGG module of the Breathe Easy software. Analysis windows of the respiratory parameters were as follows: for room air, the entire experimental period was considered; for hypercapnia, only the last 5 min challenge for each of 5/7/10% CO2 conditions were analyzed; and for hypoxia, three separate 5-min time periods, namely the 5th-10th min, 10th-15th min, and 15th-20th min intervals, were used due to the stereotypical biphasic respiratory response under this condition. Only steady quiescent breathing periods were included in the data analysis. A power analysis was performed using the reported effect size in Abbott et al. (2014), which used the same mouse model as we used here; 5-13 mice were necessary to observe a statistically significant result (Abbott et al. (2014) was able to see a significant difference between two groups with n=7 mice). In our experiments, the sample size for each group (Dbh-Cre; Vglut2 cKO and control) exceeded 13. Steady state results (Vf, VE, VT, VO2, VE/VO2, TI, TE, TTOT, VT/TI, VT/TE) for room air and hypercapnic or hypoxic data were compared between Dbh-Cre; Vglut2 cKO cohorts and sibling controls using a linear mixed-effects regression model with animal type (experimental vs. control) as fixed effects and animal ID as a random effect (Lusk et al., 2023). SD1 and SD2 results for room air and hypercapnic or hypoxic data were compared between Dbh-Cre; Vglut2 cKO cohorts and sibling controls using a Mann-Whitney U test (Ferreira et al., 2022). A p-value threshold of p < 0.05 was used to test for statistical significance. Individual data points, means, and standard errors of the mean are shown on all charts. The graphs were plotted by Prism 8.

Image quantification

Images were taken by using a Zeiss upright epifluorescent microscope and a Zeiss LSM 880 with Airyscan FAST confocal microscope. Images were captured using Zen software with z-stack function from top to bottom with 0.34μm intervals, exported, and then analyzed in Imaris using the spots and surface functions. For quantification of immunofluorescence staining, each GFP positive area coincident with DAPI (to denote nuclear localization) was defined as Vglut2 or Vglut3 expressing NA neurons while each tdTomato positive area coincident with DAPI was defined as NA neurons without any Vglut2 or Vglut3 co-expression. For quantification of Slc17a6 (Vglut2) or Slc17a8 (Vglut3) with Dbh double in situ hybridization in adult WT mice, Dbh positive areas coinciding with DAPI were identified as NA neurons and the Dbh positive areas overlapped with Slc17a6 (Vglut2) or Slc17a8 (Vglut3) positive pixels and DAPI were defined as NA neurons colocalized with Slc17a6 (Vglut2) or Slc17a8 (Vglut3). The number of Slc17a6 (Vglut2) or Slc17a8 (Vglut3) positive NA neurons and Slc17a6 (Vglut2) or Slc17a8 (Vglut3) negative NA neurons was counted in each image for both immunofluorescence and in situ experiments and the percentage of Slc17a6 (Vglut2) or Slc17a8 (Vglut3) positive NA neurons among all NA neurons in each NA nucleus in each mouse brainstem was calculated every other brain section unilaterally. For quantification of Slc17a6 (Vglut2) and Dbh double ISH in Dbh-Cre; Vglut2 cKO and their littermate controls, Slc17a6 (Vglut2) pixel intensities in Dbh positive areas coincident with DAPI (NA neurons) were measured in mutant and control images separately. The Slc17a6 (Vglut2) pixel intensity in NA neurons of control brains was normalized as 1 and the relative Slc17a6 (Vglut2) pixel intensity of NA neurons in mutant brains compared to that in control brains was calculated. At least three mouse brains were examined for each set of experiments in each group. All quantitative results were graphed using Prism 8.

Data availability

All the data supporting this study is available from the corresponding author upon request.